---
title: "ETV Cycle 01 — Coffee Spending"
subtitle: "STAT/DATA 1810 · In-class activity · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT (assessed via lab receipt + drills)
  CLO 2.1 read_csv extraction · CLO 3.1 select/filter/mutate/summarize
  CLO 4.1 ggplot + geom_point · CLO 4.4 center & spread · CLO 4.5 claim
  CLO 12.4 Privacy
Timing: 0 Ethics 8' · E 12' · T 30' · V 20' · Answer 10' = 80'
-->

# 0 · Context & ethics (8 min)

Our data are 821 entries from a personal transaction ledger.

**Talking points (CLO 12.4):** transaction logs can reveal routines, locations,
and financial habits. Model sentence: *"Someone might not expect a personal
ledger to be used to infer their routine or financial situation."*

# 1 · The central question

> **How much do customers spend on coffee, and how does the amount spent vary?**

# 2 · E — Extraction (12 min)

```{r}
library(tidyverse)

# LM1: import the data
coffee_df <- read_csv("Data/coffee_purchases.csv")
```

```{r}
# LM2: inspect
glimpse(coffee_df)
```

**Expected:** 821 rows, 7 variables. `amount` (`<dbl>`) stores debits as
negative dollar values; `description`, `account`, `month`, and `year` provide
the available context.

# 3 · T — Transformation (30 min)

```{r}
# LM3: select the relevant variables
coffee_small_df <- coffee_df |>
  select(description, amount, account, month, year)
```

```{r}
# LM4: keep coffee purchases with recorded debits
coffee_clean_df <- coffee_small_df |>
  filter(str_detect(description, regex("coffee", ignore_case = TRUE)),
         !is.na(amount))

nrow(coffee_small_df) - nrow(coffee_clean_df)
```

**Expected:** 77 rows removed, leaving 744 coffee descriptions. The defense
clause students should produce: *"The description is our operational rule for
identifying coffee, but it may miss or include some purchases."* The negative
sign is expected because the source is a transaction ledger.

```{r}
# LM5: convert ledger debits to positive spending
coffee_clean_df <- coffee_clean_df |>
  mutate(spending = -amount)
```

```{r}
# LM6: center and spread
spending_summary_df <- coffee_clean_df |>
  summarize(
    mean_spending = mean(spending),
    sd_spending   = sd(spending),
    n             = n()
  )

spending_summary_df
```

**Expected:** mean ≈ **$5.95**, sd ≈ **$5.45**, n = 744. The large spread is
partly driven by unusually large ledger entries; interpret, don't memorize.

# 4 · V — Visualization & summary (20 min)

```{r}
# LM7: histogram of spending
ggplot(coffee_clean_df, aes(x = spending)) +
  geom_histogram(binwidth = 0.25, fill = "#12263A", color = "white") +
  labs(x = "Coffee spending ($)", y = "Purchases",
       title = "Coffee spending varies across ledger entries")
```

```{r}
# LM8: geom_point — every purchase visible
ggplot(coffee_clean_df, aes(x = account, y = spending)) +
  geom_point(alpha = 0.5, color = "#12263A") +
  labs(x = "Account", y = "Coffee spending ($)",
       title = "Coffee spending by account — one point per purchase")
```

*(Optional boxplot, if the class is ahead:)*

```{r}
ggplot(coffee_clean_df, aes(x = account, y = spending)) +
  geom_boxplot(fill = "#E3C36F") +
  labs(x = "Account", y = "Coffee spending ($)")
```

**Teaching note for LM8:** points show the full transaction-level spread that
the summary table compresses away. The account groups also make the limits of
the dataset visible: this is a ledger, not a random sample of all customers.

# 5 · Interpretation & final answer (10 min)

**Model answer:** *"A typical recorded coffee purchase is about $5.95, with a
standard deviation of roughly $5.45. The plot shows that the ledger contains
many small purchases and a few much larger entries. We retained 744 of 821
rows whose descriptions contained 'coffee'; that text rule is useful but not
perfect."*

Required elements: **center (mean), spread (sd in dollars), and caveat (the
filter rule)** — that trio is the CLO 4.5 rubric for every cycle this term.
