---
title: "ETV Cycle 01 — Coffee Spending"
subtitle: "STAT/DATA 1810 · In-class activity · Student version"
author: "Your name here"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
  eval: false
---

<!--
CLO ALIGNMENT
  CLO 2 (2.1 read_csv extraction; 2.5 export awareness)
  CLO 3 (3.1 core dplyr verbs: select, filter, mutate, summarize, arrange)
  CLO 4 (4.1 ggplot2 grammar + geom_point; 4.4 center & spread; 4.5 defensible claim)
  CLO 12 (12.4 Privacy — consent & transparency in transaction data)
Time budget: 80 minutes. LM = short learning moment.
-->

# 0 · Context & ethics (8 min)

Our data are 821 entries from a personal transaction ledger. Before touching
them, two questions we owe the people in the data:

- **Privacy (CLO 12.4):** every row records a purchase. Transaction data can
  reveal routines, locations, and financial habits. *Did the people represented
  here knowingly consent to their spending being analyzed?*
- **Transparency:** if the shop uses this analysis to change prices, who should
  be told?

Write one sentence: what is one way this data could be used that a customer
might not expect?

> *Your sentence:*

# 1 · The central question

> **How much do customers spend on coffee, and how does the amount spent vary?**

Every chunk below exists only to answer this. If a line of code doesn't serve
the question, it doesn't belong.

# 2 · E — Extraction (12 min)

**LM1 — import.** The file is `Data/coffee_purchases.csv`. Import it with
`read_csv()` and store it as `coffee_df`.

```{r}
library(tidyverse)

# LM1: import the data
coffee_df <- read_csv("____")
```

**LM2 — inspect.** Look at what arrived. How many rows? Which variable answers
our question, and what type is it?

```{r}
# LM2: inspect with glimpse()

```

> *Which variables identify the purchase, and what does the sign of `amount`
> mean?*

# 3 · T — Transformation (30 min)

**LM3 — select (CLO 3.1).** We don't need every column. Keep
`description`, `amount`, `account`, `month`, and `year`.

```{r}
# LM3: select the relevant variables
coffee_small_df <- coffee_df |>
  select(____)
```

**LM4 — filter.** Keep descriptions that identify coffee purchases and remove
missing amounts. The ledger stores spending as negative debits, so a negative
amount is expected here, not impossible.

```{r}
# LM4: filter out missing / impossible amounts
coffee_clean_df <- coffee_small_df |>
  filter(____)
```

> *How many rows did you remove? Why is filtering by description a reasonable
> operational definition of a coffee purchase?*

**LM5 — mutate.** Create `spending = -amount` so the dollar amount is positive.

```{r}
# LM5: create a new variable

```

**LM6 — summarize.** Compute the **mean** and **standard deviation** of
`spending`, and the count `n()`.

```{r}
# LM6: center and spread in one summarize()
spending_summary_df <- coffee_clean_df |>
  summarize(
    mean_amount = ____,
    sd_amount   = ____,
    n           = n()
  )

spending_summary_df
```

# 4 · V — Visualization & summary (20 min)

**LM7 — the distribution.** A histogram shows *how coffee spending varies*.

```{r}
# LM7: histogram of amount

```

**LM8 — geom_point().** Plot `spending` by `account` as points, one point per
purchase. Points let you see every purchase the summary compressed away.

```{r}
# LM8: ggplot(...) + geom_point()
ggplot(coffee_clean_df, aes(x = ____, y = ____)) +
  geom_point() +
  labs(x = ____, y = ____, title = ____)
```

*(Optional, if time: make a boxplot of `spending` by `account` — does it help
answer the question?)*

# 5 · Interpretation & final answer (10 min)

Using your summary table AND one plot, answer the central question in 2–3
sentences. A complete answer names: a typical amount (the mean), how much
amounts vary (the sd, in dollars), and one caveat (what you filtered out).

> *Your answer:*

<!-- End of class: render this document. Rendering IS the receipt. -->
