---
title: "ETV Cycle 01 — Coffee Spending"
subtitle: "STAT/DATA 1810 · In-class activity · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right-body
    toc-title: "Contents"
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT (assessed via lab receipt + drills)
  CLO 2.1 read_csv extraction · CLO 3.1 select/filter/mutate/summarize
  CLO 4.1 ggplot + geom_point · CLO 4.4 center & spread · CLO 4.5 claim
  CLO 12.4 Privacy
Timing: 0 Ethics 8' · E 12' · T 30' · V 20' · Answer 10' = 80'
-->

::: {.callout-warning title="Timing & CLOs"}
- **Budget:** Ethics 8′ · E 12′ · T 30′ · V 20′ · Answer 10′ = **80′**
- **CLOs:** 2.1 `read_csv` · 3.1 select/filter/mutate/summarize · 4.1 ggplot + `geom_point` · 4.4 center & spread · 4.5 claim · 12.4 privacy
:::

# 0 · Context & ethics [8 min]{.timing}

::: {.callout-important title="Talking points (CLO 12.4)"}
- Our data: **821 entries** from a personal transaction ledger
- Transaction logs can reveal routines, locations, and financial habits
- Model sentence: *"Someone might not expect a personal ledger to be used to infer their routine or financial situation."*
:::

# 1 · The central question

> **How much do customers spend on coffee, and how does the amount spent vary?**

# 2 · E — Extraction [12 min]{.timing}

```{r}
library(tidyverse)

# LM1: import the data
coffee_df <- read_csv("Data/coffee_purchases.csv")
```

```{r}
# LM2: inspect
glimpse(coffee_df)
```

::: {.callout-note title="Expected — LM2"}
- [LM2]{.lm} **821 rows, 7 variables**
- `amount` (`<dbl>`) stores debits as **negative** dollar values
- `description`, `account`, `month`, `year` provide the available context
:::

# 3 · T — Transformation [30 min]{.timing}

```{r}
# LM3: select the relevant variables
coffee_small_df <- coffee_df |>
  select(description, amount, account, month, year)
```

```{r}
# LM4: keep coffee purchases with recorded debits
coffee_clean_df <- coffee_small_df |>
  filter(str_detect(description, regex("coffee", ignore_case = TRUE)),
         !is.na(amount))

nrow(coffee_small_df) - nrow(coffee_clean_df)
```

::: {.callout-note title="Expected — LM4"}
- [LM4]{.lm} **77 rows removed**, leaving **744** coffee descriptions
- Defense clause students should produce: *"The description is our operational rule for identifying coffee, but it may miss or include some purchases."*
- The negative sign is expected — the source is a transaction ledger
:::

```{r}
# LM5: convert ledger debits to positive spending
coffee_clean_df <- coffee_clean_df |>
  mutate(spending = -amount)
```

```{r}
# LM6: center and spread
spending_summary_df <- coffee_clean_df |>
  summarize(
    mean_spending = mean(spending),
    sd_spending   = sd(spending),
    n             = n()
  )

spending_summary_df
```

::: {.callout-note title="Expected — LM6"}
- [LM6]{.lm} mean ≈ **$5.95** · sd ≈ **$5.45** · n = **744**
- The large spread is partly driven by unusually large ledger entries — interpret, don't memorize
:::

# 4 · V — Visualization & summary [20 min]{.timing}

```{r}
# LM7: histogram of spending
ggplot(coffee_clean_df, aes(x = spending)) +
  geom_histogram(binwidth = 0.25, fill = "#12263A", color = "white") +
  labs(x = "Coffee spending ($)", y = "Purchases",
       title = "Coffee spending varies across ledger entries")
```

```{r}
# LM8: geom_point — every purchase visible
ggplot(coffee_clean_df, aes(x = account, y = spending)) +
  geom_point(alpha = 0.5, color = "#12263A") +
  labs(x = "Account", y = "Coffee spending ($)",
       title = "Coffee spending by account — one point per purchase")
```

*(Optional boxplot, if the class is ahead:)*

```{r}
ggplot(coffee_clean_df, aes(x = account, y = spending)) +
  geom_boxplot(fill = "#E3C36F") +
  labs(x = "Account", y = "Coffee spending ($)")
```

::: {.callout-important title="Teaching note — LM8"}
- [LM8]{.lm} Points show the full transaction-level spread that the summary table compresses away
- The account groups make the limits of the dataset visible: this is a **ledger, not a random sample** of all customers
:::

# 5 · Interpretation & final answer [10 min]{.timing}

::: {.callout-tip title="Model answer"}
- *"A typical recorded coffee purchase is about $5.95, with a standard deviation of roughly $5.45. The plot shows that the ledger contains many small purchases and a few much larger entries. We retained 744 of 821 rows whose descriptions contained 'coffee'; that text rule is useful but not perfect."*
- Required elements (the CLO 4.5 rubric for every cycle this term):
  - **center** — the mean
  - **spread** — the sd, in dollars
  - **caveat** — the filter rule
:::
