---
title: "Foundations 02 — Data Types and Data Structures"
subtitle: "STAT/DATA 1810 · Week 2 Monday · Student version"
author: "Your name here"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT
  CLO 1 (1.5 data types: numeric/character/factor/date · 1.6 containers:
         vector, data frame/tibble, list, matrix · 1.4 assignment reinforced)
  CLO 7 preview (7.6 is.na())
  CLO 9 practiced: students INSERT THEIR OWN CHUNKS today
Time: 80 min. Datasets: iris, airquality (built in) + tiny constructed examples.
Rule of the day: a TYPE is what one value is; a STRUCTURE is how values are held.
LM = learning moment. YT = your turn. Times in headers are the section budget.
-->

```{r}
#| label: setup
library(tidyverse)
```

# 0 · The one distinction, and your own chunks (8 min)

> **Type** = what a single value *is* (number? text? TRUE/FALSE?).
> **Structure** = the container holding many values (vector? data frame? list?).

Everything today hangs off that sentence.

**New skill today: you make the chunks.** Last week every chunk was provided.
From now on, when you see **"➕ insert a chunk"**, you create one yourself.
Three ways — pick one and practice it until it's automatic:

1. Keyboard: **Cmd + Shift + I** (Mac) / **Ctrl + Shift + I** (Windows).
2. Command Palette (Cmd/Ctrl + Shift + P) → type *Quarto: Insert Code Cell*.
3. Type it by hand: a line with ```` ```{r} ````, your code, then a line with
   ```` ``` ````.

> *YT0 (3 min). Directly below this line, insert a chunk. Inside it, write a
> comment `# YT0` and the line `1 + 1`. Run it. If output appears under your
> chunk, you're ready.*

<!-- ➕ insert a chunk here -->

# 1 · Data types (22 min)

**LM1 — the four you'll live with (4 min).** Predict each `class()` before
running:

```{r}
# LM1: predict, then run
class(4.75)
class(4L)
class("latte")
class(TRUE)
```

> *Q1. Predictions: ____ · ____ · ____ · ____*

**LM2 — doubles vs integers (2 min).** Both are numeric; `typeof()` tells
them apart.

```{r}
# LM2
typeof(4.75)
typeof(4L)
```

**LM3 — factors (4 min).** A factor is text with a fixed menu of allowed
values (levels). `iris$Species` is built in:

```{r}
# LM3: character vs factor
class(iris$Species)
levels(iris$Species)
```

> *Q2. How many levels does Species have, and what are they?*
>
> *Your answer:*

**LM4 — dates and date-times (2 min).**

```{r}
# LM4
today <- Sys.Date()
class(today)

right_now <- Sys.time()
class(right_now)
```

**LM5 — NA, the missing value (3 min).** `airquality` (built in) has real
missing ozone readings:

```{r}
# LM5: count the missing values in Ozone
sum(is.na(airquality$Ozone))
```

> *Q3. How many Ozone values are missing? ____*

**LM6 — the is.\* family (3 min).** Test types with functions that return
TRUE/FALSE:

```{r}
#| eval: false
# LM6: fill each blank so all four return TRUE, then delete the eval line
is.numeric(____)
is.character(____)
is.logical(____)
is.factor(____)
```

> *YT1 (4 min). ➕ Insert a chunk below. In it, find the `class()` of each of
> the following, predicting first: `"TRUE"` (with quotes) · `3 > 2` ·
> `mtcars$cyl` · `Sys.Date() + 7`. Then in the comment, write which one
> surprised you.*

<!-- ➕ insert a chunk here -->

> *Predictions: ____ · ____ · ____ · ____*

# 2 · Data structures (30 min)

**LM7 — vector: one type, many values (4 min).** Built with `c()`:

```{r}
# LM7: a vector
drinks <- c("latte", "drip", "mocha")
drinks
length(drinks)
class(drinks)
```

> *YT2 (4 min). ➕ Insert a chunk. Build THREE vectors and print each:*
> *(a) `prices` — three numbers, one price per drink above;*
> *(b) `sizes` — three sizes as text ("small", "medium", "large");*
> *(c) `iced` — three TRUE/FALSE values, whether each drink is iced.*
> *Then run `class()` on each. What happens if you run `c(1, "two", 3)`? Try
> it and look at the class.*

<!-- ➕ insert a chunk here -->

> *Q4. `c(1, "two", 3)` has class ____. In one phrase, why?*
>
> *Your answer:*

**LM8 — data frame & tibble: vectors standing side by side (6 min).** Each
column is a vector; every column has the same length.

```{r}
# LM8: same data, two containers
menu_df  <- data.frame(drink = drinks, price = c(4.75, 2.75, 5.25))
menu_tbl <- tibble(drink = drinks, price = c(4.75, 2.75, 5.25))

menu_df
menu_tbl
```

> *Q5. Name one visible difference between how the data frame and the tibble
> printed:*
>
> *Your answer:*

> *YT3 (5 min). ➕ Insert a chunk. Build a tibble called `menu` from YOUR
> vectors from YT2 — columns `drink`, `price`, `size`, `iced`. Print it.
> Then run `nrow(menu)`, `ncol(menu)`, and `class(menu)`. What happens if one
> vector has only two values? Try it (make a broken one on purpose), read
> the error, then fix it.*

<!-- ➕ insert a chunk here -->

**LM9 — tribble: a tibble typed row by row (2 min)** (brief — you'll rarely
need it):

```{r}
# LM9
tribble(
  ~drink,  ~price,
  "chai",   4.50,
  "espresso", 3.00
)
```

**LM10 — list: the junk drawer (4 min).** A list holds *anything*, even other
structures:

```{r}
# LM10
my_list <- list(name = "week 2", nums = c(1, 2, 3), data = menu_tbl)
my_list
length(my_list)
```

> *YT4 (3 min). ➕ Insert a chunk. Build a list called `me` with three named
> parts: your first name (text), the number of classes you're taking, and
> your `menu` tibble. Print it. Notice how each part keeps its own type.*

<!-- ➕ insert a chunk here -->

**LM11 — matrix: all one type, rows × columns (2 min)** (brief):

```{r}
# LM11
matrix(1:6, nrow = 2)
```

> *Q6. Which structure could hold a column of names AND a column of numbers:
> a matrix or a data frame? Why?*
>
> *Your answer:*

# 3 · In-class activity: type it or contain it (15 min)

For each item, write **T** if the question is about a type and **S** if it's
about a structure, then answer it with code. ➕ Insert **one chunk per item**
(A through F) so each answer is its own runnable piece.

- **A.** What is `iris` — its structure?
- **B.** What is `iris$Sepal.Length` — its type?
- **C.** Is there any NA anywhere in `airquality$Solar.R`? How many?
- **D.** Build a tibble with 2 columns: your three favorite foods and a
  rating (1–10) for each.
- **E.** What is `airquality$Month` — its type? Should it be? (Months are
  numbers here — would a factor make more sense? One sentence.)
- **F.** Build a list holding your food tibble from D *and* a character
  vector of the three places you eat those foods.

<!-- ➕ insert chunks A–F here -->

> *A: ____ · B: ____ · C: ____ · D: ____ · E: ____ · F: ____*

# 4 · Exit question (5 min)

> *In your own words, one sentence each: what is a data type, and what is a
> data structure? Use one example of each from today — and at least one must
> be something YOU built.*
>
> *Your answer:*

Save as `fnd02_yourlastname.qmd` in your `R/` folder and **Render**. If a
chunk you inserted breaks the render, read the error: it names the chunk.
