---
title: "Foundations 02 — Data Types and Data Structures"
subtitle: "STAT/DATA 1810 · Week 2 Monday · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right-body
    toc-title: "Contents"
execute:
  warning: false
  message: false
---

<!--
CLO: 1.5, 1.6, 1.4 · 7.6 preview · 9 practiced (students insert chunks)
Timing: distinction+chunks 8' · types 22' · structures 30' · activity 15'
        · exit 5' = 80'
Emphasis per spec: vectors, data frames/tibbles, lists get practice;
matrix & tribble get one chunk each, no drills.
Checkpoint: §2 should start by the 30' mark. If behind, drop YT4 and E/F.
-->

```{r}
#| label: setup
library(tidyverse)
```

::: {.callout-warning title="Timing & emphasis"}
- **Budget:** distinction + chunks 8′ · types 22′ · structures 30′ · activity 15′ · exit 5′ = **80′**
- **Checkpoint:** §2 should start by the **30′ mark** — if behind, drop YT4 and activity items E/F
- Practice goes to **vectors, data frames/tibbles, lists**; matrix & tribble get one chunk each, no drills
:::

# 0 · The one distinction, and your own chunks [8 min]{.timing}

::: {.callout-important title="Teaching note — the sentence on the board"}
- Write it and leave it there all class: **type = one value's nature; structure = the container**
:::

::: {.callout-important title="Teaching note — inserting chunks"}
- Demonstrate the shortcut (**Cmd/Ctrl + Shift + I**) *once*, slowly, on the projector: cursor on a blank line → press it → fence appears → type `1 + 1` → run
- Then [YT0]{.yt}: walk the room for 2 minutes — every screen must show output under a student-made chunk before moving on
- This is the one thing today that can't be caught up on later; every later YT depends on it
- Students who type the fence by hand are fine — the skill is "I can make a chunk," not the shortcut
:::

```{r}
# YT0
1 + 1
```

# 1 · Data types [22 min]{.timing}

```{r}
# LM1 — expected: "numeric", "integer", "character", "logical"
class(4.75)
class(4L)
class("latte")
class(TRUE)
```

::: {.callout-note title="Expected — LM1 & Q1"}
- [LM1]{.lm} `"numeric"` · `"integer"` · `"character"` · `"logical"`
- [Q1]{.q} most students predict `"numeric"` for `4L` — good; the surprise is the point
- The `L` suffix is worth 30 seconds, no more
:::

```{r}
# LM2 — "double" vs "integer"
typeof(4.75)
typeof(4L)
```

```{r}
# LM3 — factor with 3 levels
class(iris$Species)
levels(iris$Species)
```

::: {.callout-tip title="Q2 answer"}
- [Q2]{.q} **3 levels** — setosa, versicolor, virginica
- The menu metaphor ("a factor is text with a fixed menu") pays rent in week 6 when forcats arrives
:::

```{r}
# LM4
today <- Sys.Date()
class(today)
right_now <- Sys.time()
class(right_now)
```

::: {.callout-note title="Expected — LM4"}
- [LM4]{.lm} `"Date"` and `"POSIXct" "POSIXt"`
- Say "date-time" and move on — lubridate handles the ugliness in week 4
:::

```{r}
# LM5
sum(is.na(airquality$Ozone))
```

::: {.callout-tip title="Q3 answer"}
- [Q3]{.q} **37 missing**
- Real 1973 NYC air data — the missing values are genuinely missing, not planted
- First appearance of the `is.na()` + `sum()` idiom; students use it in every cycle
:::

```{r}
# LM6 — any correct fills, e.g.
is.numeric(4.75)
is.character("latte")
is.logical(TRUE)
is.factor(iris$Species)
```

```{r}
# YT1 — expected: "character", "logical", "numeric", "Date"
class("TRUE")
class(3 > 2)
class(mtcars$cyl)
class(Sys.Date() + 7)
```

::: {.callout-note title="Expected — YT1"}
- [YT1]{.yt} `"character"` · `"logical"` · `"numeric"` · `"Date"`
- Surprise #1: `"TRUE"` — quotes make it text
- Surprise #2: `3 > 2` *produces* a logical — comparisons are how logicals get made in practice (`filter()` in two weeks)
- `mtcars$cyl` being numeric though it only takes 4/6/8 previews activity item E
:::

# 2 · Data structures [30 min]{.timing}

```{r}
# LM7
drinks <- c("latte", "drip", "mocha")
drinks
length(drinks)
class(drinks)
```

```{r}
# YT2
prices <- c(4.75, 2.75, 5.25)
sizes  <- c("small", "medium", "large")
iced   <- c(TRUE, FALSE, TRUE)
prices; sizes; iced
class(prices); class(sizes); class(iced)
c(1, "two", 3)
class(c(1, "two", 3))
```

::: {.callout-tip title="Q4 answer — coercion"}
- [Q4]{.q} **character**
- A vector holds *one* type, so R converts everything to the type that can hold all of it — text
- This is **coercion**, arriving by accident rather than lecture — say the word once
- They meet it again when a numeric column imports as `<chr>` because of one stray "N/A"
:::

```{r}
# LM8
menu_df  <- data.frame(drink = drinks, price = c(4.75, 2.75, 5.25))
menu_tbl <- tibble(drink = drinks, price = c(4.75, 2.75, 5.25))
menu_df
menu_tbl
```

::: {.callout-tip title="Q5 model answers (any one)"}
- [Q5]{.q} the tibble announces its dimensions (`# A tibble: 3 × 2`)
- the tibble shows column types (`<chr>`, `<dbl>`)
- the data frame shows row numbers without types
- **Framing:** a tibble IS a data frame with better manners — we'll say "data frame" for both
:::

```{r}
# YT3
menu <- tibble(drink = drinks, price = prices, size = sizes, iced = iced)
menu
nrow(menu); ncol(menu); class(menu)
```

```{r}
#| eval: false
# YT3 — the deliberate break
tibble(drink = drinks, price = c(4.75, 2.75))
#> Error: Tibble columns must have compatible sizes.
#> • Size 3: Existing data.
#> • Size 2: Column `price`.
```

::: {.callout-caution title="YT3 teaching note — a good error"}
- [YT3]{.yt} The error names the sizes *and* the column — have someone read it aloud
- "Every column the same length" is now something they've felt, not heard
- `class(menu)` returns three things (`tbl_df`, `tbl`, `data.frame`) — say "it's a data frame, with extra labels" and move on
:::

```{r}
# LM9 — tribble, briefly
tribble(
  ~drink,  ~price,
  "chai",   4.50,
  "espresso", 3.00
)
```

```{r}
# LM10 — list
my_list <- list(name = "week 2", nums = c(1, 2, 3), data = menu_tbl)
my_list
length(my_list)
```

::: {.callout-important title="LM10 — junk-drawer framing"}
- [LM10]{.lm} A list holds anything, including a whole tibble
- `length()` is 3 — three drawers, whatever's inside them
- Plant the flag: *"when we pull data from an API in week 8, it arrives as a list"*
:::

```{r}
# YT4 — e.g.
me <- list(first = "Eman", n_classes = 4, menu = menu)
me
```

```{r}
# LM11 — matrix, briefly
matrix(1:6, nrow = 2)
```

::: {.callout-tip title="Q6 answer"}
- [Q6]{.q} **a data frame**
- A matrix forces every cell to one type, so names + numbers would coerce the numbers to text — same coercion as Q4, now at the structure level
:::

# 3 · In-class activity [15 min]{.timing}

::: {.callout-important title="Teaching note — one chunk per item"}
- Check that students actually made **six chunks**
- The habit of "one idea, one chunk" is what makes ETV documents readable
:::

```{r}
# A — Structure
class(iris)          # "data.frame"
```

```{r}
# B — Type
class(iris$Sepal.Length)   # "numeric"
```

```{r}
# C — Type question (is.na tests values)
sum(is.na(airquality$Solar.R))   # 7
```

```{r}
# D — Structure built by the student, e.g.
foods_tbl <- tibble(
  food   = c("tacos", "pho", "pizza"),
  rating = c(10, 9, 8)
)
foods_tbl
```

```{r}
# E — Type; stored as integer, arguably should be a factor (or a Date)
class(airquality$Month)
```

```{r}
# F — Structure: a list holding a tibble and a vector
food_places <- list(
  foods  = foods_tbl,
  places = c("truck on Higuera", "home", "campus")
)
food_places
```

::: {.callout-tip title="Activity answers"}
- **A** = S (structure)
- **B** = T (type)
- **C** = T — 7 missing
- **D** = S
- **E** = T — `"integer"`; a factor would make sense since 5–9 are labels, not amounts
- **F** = S
- **E is the discussion item:** *the type R stores is not always the type the data means* — that's the whole reason `mutate()` + `factor()` exists
:::

# 4 · Exit question [5 min]{.timing}

::: {.callout-tip title="Model answer"}
- *"A data type is what one value is — `4.75` is numeric. A data structure is the container holding values — my `menu` tibble is a data frame whose `iced` column is logical."*
- Passes: any answer pairing one correct **type** example with one correct **structure** example
- At least one example should be from a student-built object
:::
