---
title: "Foundations 02 — Data Types and Data Structures"
subtitle: "STAT/DATA 1810 · Week 2 Monday · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right-body
    toc-title: "Contents"
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT
  CLO 1 (1.5 data types: numeric/character/factor/date · 1.6 containers:
         vector, data frame/tibble, list, matrix · 1.4 assignment reinforced)
  CLO 7 preview (7.6 is.na())
  CLO 9 practiced: students INSERT THEIR OWN CHUNKS today
Time: 80 min. Datasets: iris, airquality (built in) + tiny constructed examples.
Timing: distinction+chunks 8' · types 22' · structures 30' · activity 15'
        · exit 5' = 80'
Emphasis per spec: vectors, data frames/tibbles, lists get practice;
matrix & tribble get one chunk each, no drills.
Checkpoint: §2 should start by the 30' mark. If behind, drop YT4 and E/F.
Rule of the day: a TYPE is what one value is; a STRUCTURE is how values are held.
LM = learning moment. YT = your turn. Times in headers are the section budget.
-->

```{r}
#| label: setup
library(tidyverse)
```

::: {.callout-warning title="Timing & emphasis"}
- **Budget:** distinction + chunks 8′ · types 22′ · structures 30′ · activity 15′ · exit 5′ = **80′**
- **Checkpoint:** §2 should start by the **30′ mark** — if behind, drop YT4 and activity items E/F
- Practice goes to **vectors, data frames/tibbles, lists**; matrix & tribble get one chunk each, no drills
:::

# 0 · The one distinction, and your own chunks [8 min]{.timing}

> **Type** = what a single value *is* (number? text? TRUE/FALSE?).
> **Structure** = the container holding many values (vector? data frame? list?).

Everything today hangs off that sentence.

::: {.callout-important title="Teaching note — the sentence on the board"}
- Write it and leave it there all class: **type = one value's nature; structure = the container**
:::

**New skill today: you make the chunks.** Last week every chunk was provided.
From now on, when you see **"➕ insert a chunk"**, you create one yourself.
Three ways — pick one and practice it until it's automatic:

1. Keyboard: **Cmd + Shift + I** (Mac) / **Ctrl + Shift + I** (Windows).
2. Command Palette (Cmd/Ctrl + Shift + P) → type *Quarto: Insert Code Cell*.
3. Type it by hand: a line with ```` ```{r} ````, your code, then a line with
   ```` ``` ````.

::: {.callout-important title="Teaching note — inserting chunks"}
- Demonstrate the shortcut (**Cmd/Ctrl + Shift + I**) *once*, slowly, on the projector: cursor on a blank line → press it → fence appears → type `1 + 1` → run
- Then [YT0]{.yt}: walk the room for 2 minutes — every screen must show output under a student-made chunk before moving on
- This is the one thing today that can't be caught up on later; every later YT depends on it
- Students who type the fence by hand are fine — the skill is "I can make a chunk," not the shortcut
:::

> *YT0 (3 min). Directly below this line, insert a chunk. Inside it, write a
> comment `# YT0` and the line `1 + 1`. Run it. If output appears under your
> chunk, you're ready.*

<!-- ➕ insert a chunk here -->

```{r}
# YT0
1 + 1
```

# 1 · Data types [22 min]{.timing}

**LM1 — the four you'll live with (4 min).** Predict each `class()` before
running:

```{r}
# LM1: predict, then run
class(4.75)
class(4L)
class("latte")
class(TRUE)
```

> *Q1. Predictions: numeric · integer · character · logical*

::: {.callout-note title="Expected — LM1 & Q1"}
- [LM1]{.lm} `"numeric"` · `"integer"` · `"character"` · `"logical"`
- [Q1]{.q} most students predict `"numeric"` for `4L` — good; the surprise is the point
- The `L` suffix is worth 30 seconds, no more
:::

**LM2 — doubles vs integers (2 min).** Both are numeric; `typeof()` tells
them apart.

```{r}
# LM2
typeof(4.75)
typeof(4L)
```

**LM3 — factors (4 min).** A factor is text with a fixed menu of allowed
values (levels). `iris$Species` is built in:

```{r}
# LM3: character vs factor
class(iris$Species)
levels(iris$Species)
```

> *Q2. How many levels does Species have, and what are they?*
>
> *Your answer:* 3 levels — setosa, versicolor, virginica.

::: {.callout-tip title="Q2 answer"}
- [Q2]{.q} **3 levels** — setosa, versicolor, virginica
- The menu metaphor ("a factor is text with a fixed menu") pays rent in week 6 when forcats arrives
:::

**LM4 — dates and date-times (2 min).**

```{r}
# LM4
today <- Sys.Date()
class(today)

right_now <- Sys.time()
class(right_now)
```

::: {.callout-note title="Expected — LM4"}
- [LM4]{.lm} `"Date"` and `"POSIXct" "POSIXt"`
- Say "date-time" and move on — lubridate handles the ugliness in week 4
:::

**LM5 — NA, the missing value (3 min).** `airquality` (built in) has real
missing ozone readings:

```{r}
# LM5: count the missing values in Ozone
sum(is.na(airquality$Ozone))
```

> *Q3. How many Ozone values are missing? 37*

::: {.callout-tip title="Q3 answer"}
- [Q3]{.q} **37 missing**
- Real 1973 NYC air data — the missing values are genuinely missing, not planted
- First appearance of the `is.na()` + `sum()` idiom; students use it in every cycle
:::

**LM6 — the is.\* family (3 min).** Test types with functions that return
TRUE/FALSE:

```{r}
# LM6: any correct fills work, e.g.
is.numeric(4.75)
is.character("latte")
is.logical(TRUE)
is.factor(iris$Species)
```

> *YT1 (4 min). ➕ Insert a chunk below. In it, find the `class()` of each of
> the following, predicting first: `"TRUE"` (with quotes) · `3 > 2` ·
> `mtcars$cyl` · `Sys.Date() + 7`. Then in the comment, write which one
> surprised you.*

<!-- ➕ insert a chunk here -->

```{r}
# YT1 — surprise: "TRUE" in quotes is character, and 3 > 2 makes a logical
class("TRUE")
class(3 > 2)
class(mtcars$cyl)
class(Sys.Date() + 7)
```

> *Predictions: character · logical · numeric · Date*

::: {.callout-note title="Expected — YT1"}
- [YT1]{.yt} `"character"` · `"logical"` · `"numeric"` · `"Date"`
- Surprise #1: `"TRUE"` — quotes make it text
- Surprise #2: `3 > 2` *produces* a logical — comparisons are how logicals get made in practice (`filter()` in two weeks)
- `mtcars$cyl` being numeric though it only takes 4/6/8 previews activity item E
:::

# 2 · Data structures [30 min]{.timing}

**LM7 — vector: one type, many values (4 min).** Built with `c()`:

```{r}
# LM7: a vector
drinks <- c("latte", "drip", "mocha")
drinks
length(drinks)
class(drinks)
```

> *YT2 (4 min). ➕ Insert a chunk. Build THREE vectors and print each:*
> *(a) `prices` — three numbers, one price per drink above;*
> *(b) `sizes` — three sizes as text ("small", "medium", "large");*
> *(c) `iced` — three TRUE/FALSE values, whether each drink is iced.*
> *Then run `class()` on each. What happens if you run `c(1, "two", 3)`? Try
> it and look at the class.*

<!-- ➕ insert a chunk here -->

```{r}
# YT2
prices <- c(4.75, 2.75, 5.25)
sizes  <- c("small", "medium", "large")
iced   <- c(TRUE, FALSE, TRUE)
prices; sizes; iced
class(prices); class(sizes); class(iced)
c(1, "two", 3)
class(c(1, "two", 3))
```

> *Q4. `c(1, "two", 3)` has class character. In one phrase, why?*
>
> *Your answer:* A vector holds one type, so R coerces everything to text.

::: {.callout-tip title="Q4 answer — coercion"}
- [Q4]{.q} **character**
- A vector holds *one* type, so R converts everything to the type that can hold all of it — text
- This is **coercion**, arriving by accident rather than lecture — say the word once
- They meet it again when a numeric column imports as `<chr>` because of one stray "N/A"
:::

**LM8 — data frame & tibble: vectors standing side by side (6 min).** Each
column is a vector; every column has the same length.

```{r}
# LM8: same data, two containers
menu_df  <- data.frame(drink = drinks, price = c(4.75, 2.75, 5.25))
menu_tbl <- tibble(drink = drinks, price = c(4.75, 2.75, 5.25))

menu_df
menu_tbl
```

> *Q5. Name one visible difference between how the data frame and the tibble
> printed:*
>
> *Your answer:* The tibble announces its dimensions (`# A tibble: 3 × 2`) and
> shows column types (`<chr>`, `<dbl>`); the data frame shows neither.

::: {.callout-tip title="Q5 model answers (any one)"}
- [Q5]{.q} the tibble announces its dimensions (`# A tibble: 3 × 2`)
- the tibble shows column types (`<chr>`, `<dbl>`)
- the data frame shows row numbers without types
- **Framing:** a tibble IS a data frame with better manners — we'll say "data frame" for both
:::

> *YT3 (5 min). ➕ Insert a chunk. Build a tibble called `menu` from YOUR
> vectors from YT2 — columns `drink`, `price`, `size`, `iced`. Print it.
> Then run `nrow(menu)`, `ncol(menu)`, and `class(menu)`. What happens if one
> vector has only two values? Try it (make a broken one on purpose), read
> the error, then fix it.*

<!-- ➕ insert a chunk here -->

```{r}
# YT3
menu <- tibble(drink = drinks, price = prices, size = sizes, iced = iced)
menu
nrow(menu); ncol(menu); class(menu)
```

```{r}
#| eval: false
# YT3 — the deliberate break
tibble(drink = drinks, price = c(4.75, 2.75))
#> Error: Tibble columns must have compatible sizes.
#> • Size 3: Existing data.
#> • Size 2: Column `price`.
```

```{r}
# YT3 — the fix: give price all three values
tibble(drink = drinks, price = c(4.75, 2.75, 5.25))
```

::: {.callout-caution title="YT3 teaching note — a good error"}
- [YT3]{.yt} The error names the sizes *and* the column — have someone read it aloud
- "Every column the same length" is now something they've felt, not heard
- `class(menu)` returns three things (`tbl_df`, `tbl`, `data.frame`) — say "it's a data frame, with extra labels" and move on
:::

**LM9 — tribble: a tibble typed row by row (2 min)** (brief — you'll rarely
need it):

```{r}
# LM9
tribble(
  ~drink,  ~price,
  "chai",   4.50,
  "espresso", 3.00
)
```

**LM10 — list: the junk drawer (4 min).** A list holds *anything*, even other
structures:

```{r}
# LM10
my_list <- list(name = "week 2", nums = c(1, 2, 3), data = menu_tbl)
my_list
length(my_list)
```

::: {.callout-important title="LM10 — junk-drawer framing"}
- [LM10]{.lm} A list holds anything, including a whole tibble
- `length()` is 3 — three drawers, whatever's inside them
- Plant the flag: *"when we pull data from an API in week 8, it arrives as a list"*
:::

> *YT4 (3 min). ➕ Insert a chunk. Build a list called `me` with three named
> parts: your first name (text), the number of classes you're taking, and
> your `menu` tibble. Print it. Notice how each part keeps its own type.*

<!-- ➕ insert a chunk here -->

```{r}
# YT4 — e.g.
me <- list(first = "Eman", n_classes = 4, menu = menu)
me
```

**LM11 — matrix: all one type, rows × columns (2 min)** (brief):

```{r}
# LM11
matrix(1:6, nrow = 2)
```

> *Q6. Which structure could hold a column of names AND a column of numbers:
> a matrix or a data frame? Why?*
>
> *Your answer:* A data frame — a matrix forces every cell to one type, so the
> numbers would be coerced to text.

::: {.callout-tip title="Q6 answer"}
- [Q6]{.q} **a data frame**
- A matrix forces every cell to one type, so names + numbers would coerce the numbers to text — same coercion as Q4, now at the structure level
:::

# 3 · In-class activity: type it or contain it [15 min]{.timing}

For each item, write **T** if the question is about a type and **S** if it's
about a structure, then answer it with code. ➕ Insert **one chunk per item**
(A through F) so each answer is its own runnable piece.

- **A.** What is `iris` — its structure?
- **B.** What is `iris$Sepal.Length` — its type?
- **C.** Is there any NA anywhere in `airquality$Solar.R`? How many?
- **D.** Build a tibble with 2 columns: your three favorite foods and a
  rating (1–10) for each.
- **E.** What is `airquality$Month` — its type? Should it be? (Months are
  numbers here — would a factor make more sense? One sentence.)
- **F.** Build a list holding your food tibble from D *and* a character
  vector of the three places you eat those foods.

::: {.callout-important title="Teaching note — one chunk per item"}
- Check that students actually made **six chunks**
- The habit of "one idea, one chunk" is what makes ETV documents readable
:::

<!-- ➕ insert chunks A–F here -->

```{r}
# A — Structure
class(iris)          # "data.frame"
```

```{r}
# B — Type
class(iris$Sepal.Length)   # "numeric"
```

```{r}
# C — Type question (is.na tests values)
sum(is.na(airquality$Solar.R))   # 7
```

```{r}
# D — Structure built by the student, e.g.
foods_tbl <- tibble(
  food   = c("tacos", "pho", "pizza"),
  rating = c(10, 9, 8)
)
foods_tbl
```

```{r}
# E — Type; stored as integer, arguably should be a factor (or a Date)
class(airquality$Month)
```

```{r}
# F — Structure: a list holding a tibble and a vector
food_places <- list(
  foods  = foods_tbl,
  places = c("truck on Higuera", "home", "campus")
)
food_places
```

> *A: S · B: T · C: T (7 missing) · D: S · E: T — integer; a factor would make
> more sense since 5–9 are labels, not amounts · F: S*

::: {.callout-tip title="Activity answers"}
- **A** = S (structure)
- **B** = T (type)
- **C** = T — 7 missing
- **D** = S
- **E** = T — `"integer"`; a factor would make sense since 5–9 are labels, not amounts
- **F** = S
- **E is the discussion item:** *the type R stores is not always the type the data means* — that's the whole reason `mutate()` + `factor()` exists
:::

# 4 · Exit question [5 min]{.timing}

> *In your own words, one sentence each: what is a data type, and what is a
> data structure? Use one example of each from today — and at least one must
> be something YOU built.*
>
> *Your answer:* A data type is what one value is — `4.75` is numeric. A data
> structure is the container holding values — my `menu` tibble is a data frame
> whose `iced` column is logical.

::: {.callout-tip title="Model answer"}
- *"A data type is what one value is — `4.75` is numeric. A data structure is the container holding values — my `menu` tibble is a data frame whose `iced` column is logical."*
- Passes: any answer pairing one correct **type** example with one correct **structure** example
- At least one example should be from a student-built object
:::

Save as `fnd02_yourlastname.qmd` in your `R/` folder and **Render**. If a
chunk you inserted breaks the render, read the error: it names the chunk.
