---
title: "Foundations 02 — Data Types and Data Structures"
subtitle: "STAT/DATA 1810 · Week 2 Monday · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right-body
    toc-title: "Contents"
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT
  CLO 1 (1.5 data types: numeric/character/factor/date · 1.6 containers:
         vector, data frame/tibble, list, matrix · 1.4 assignment reinforced)
  CLO 7 preview (7.6 is.na())
  CLO 9 practiced: students insert their own chunks (taught in Foundations 01)
Time: 80 min. Datasets: tiny Cal Poly examples built in class + iris, airquality.
Timing: distinction 4' · types 28' · structures 24' · when things break 6'
        · activity 12' · exit 5' · save & render 1' = 80'
Checkpoint: §2 should start by the 32' mark. If behind, drop YT4 and E/F.
Carries the Fnd 01 conventions: Name/Date line, LM/YT legend, 👀 Variables
checks, a 'when things break' section, restart-then-render close.
Rule of the day: a TYPE is what one value is; a STRUCTURE is how values are held.
Packages: ONLY tibble today (tibble(), tribble()). No tidyverse — load what the
lesson needs, nothing more.
LM = learning moment. YT = your turn. Times in headers are the section budget.
-->

```{r}
#| label: setup
library(tibble)   # today we need tibble() and tribble() — that's it
```

> 👀 **Variables check.** Loading `tibble` added *nothing* to the Variables
> pane. `library()` opens a toolbox; only `<-` creates an object.

**Name:** Instructor key · **Date:** \_\_\_\_\_\_\_\_\_\_

> Before anything else, type your name and today's date on the line above (and put your name in the `author:` line at the very top of this file).

**How to read this document.** Two labels show up over and over:

- **LM = Learning Moment.** I demonstrate; you follow along and run the code.
- **YT = Your Turn.** You do it yourself — fill the blank or the empty chunk.
- **👀 Variables check.** Glance at the **Variables** pane (right side, under
  SESSION) and answer the question. It is your session's memory made visible.

::: {.callout-warning title="Timing & labels"}
- **Budget:** distinction 4′ · types 28′ · structures 24′ · when things break 6′ · activity 12′ · exit 5′ · save & render 1′ = **80′**
- **Checkpoint:** §2 should start by the **32′ mark** — if behind, drop YT4 and activity items E/F
- [LM]{.lm} **Learning Moment** — instructor demonstrates, students follow
- [YT]{.yt} **Your Turn** — students do it themselves
- 👀 **Variables check** — the Fnd 01 / Lab 01 habit; keep it alive: every 👀 is a 20-second look at the pane, not a discussion
- Say the LM/YT names out loud the first time each appears — students otherwise guess
- Practice goes to **vectors, data frames/tibbles, lists**; matrix & tribble get one chunk each, no drills
- **Packages:** only `tibble`. Say it out loud: *"we load what today needs, not the whole tidyverse"*
:::

# 0 · The one distinction [4 min]{.timing}

> **Type** = what a single value *is* (number? text? TRUE/FALSE?).
> **Structure** = the container holding many values (vector? data frame? list?).

Everything today hangs off that sentence.

You already know how to insert a chunk (**Cmd/Ctrl + Shift + I**, from last
week). Whenever you see **"➕ insert a chunk"**, make one — every YT today
depends on it.

::: {.callout-important title="Teaching note — the sentence on the board"}
- Write it and leave it there all class: **type = one value's nature; structure = the container**
- Chunk insertion was taught in Foundations 01 — do **not** re-teach it; if someone is stuck, a neighbor shows them the shortcut
- One line on the `setup` chunk: we load `tibble` only. Last week's `library(tidyverse)` loads nine packages; today needs one function family from one of them
:::

{{< vspace 1.5rem >}}

# 1 · Data types [28 min]{.timing}

## 1.1 · The types, defined first (before any code) [5 min]{.timing}

Every single value in R has a **type**. Here are the ones you'll live with,
each with a Cal Poly example:

| Type | What it is | Example value |
|---|---|---|
| **numeric** (double) | a number that can have decimals | `4.75` — a latte at Julian's |
| **integer** | a whole number (written with an `L`) | `38L` — people in the Rec Center at 7 am |
| **character** | text, in quotes | `"Scout Coffee"` |
| **logical** | `TRUE` or `FALSE` | `TRUE` — the drink is iced |
| **factor** | text from a *fixed menu* of allowed values | `"morning"` from {morning, afternoon, evening} |
| **Date** / **date-time** | a calendar date; a date *and* clock time | `2026-09-22` / `2026-09-22 07:00:00` |
| **NA** | a missing value — a hole, not a type of its own | the headcount nobody recorded |

::: {.callout-important title="Teaching note — definitions before class()"}
- Read the table row by row *before* anyone runs code. The order today is: **define the type → then discover it with `class()`**
- Ask: "If I write a latte price, which row?" (numeric) "The name of the shop?" (character) "Whether it's iced?" (logical)
- `NA` gets one sentence: it's a placeholder for "nobody knows," and it can sit inside any type
:::

## 1.2 · Finding the type with `class()` [7 min]{.timing}

**LM1 — `class()` tells you the type (4 min).** Three lattes, three prices.
Predict each `class()` from the table above, *then* run:

```{r}
# LM1: predict from the table, then run
class(4.75)            # Julian's latte
class(38L)             # people in the Rec Center at 7 am
class("Scout Coffee")  # where the latte came from
class(TRUE)            # iced?
```

> *Q1. Predictions: numeric · integer · character · logical*

::: {.callout-note title="Expected — LM1 & Q1"}
- [LM1]{.lm} `"numeric"` · `"integer"` · `"character"` · `"logical"`
- [Q1]{.q} most students predict `"numeric"` for `38L` — the `L` suffix is worth 30 seconds, no more: R stores whole numbers as integers when you ask, otherwise as doubles
:::

**LM2 — doubles vs integers (1 min).** Both are numeric; `typeof()` tells them
apart.

```{r}
# LM2
typeof(4.75)
typeof(38L)
```

**LM3 — dates and date-times (2 min).**

```{r}
# LM3
today <- Sys.Date()
class(today)

right_now <- Sys.time()
class(right_now)
```

::: {.callout-note title="Expected — LM3"}
- [LM3]{.lm} `"Date"` and `"POSIXct" "POSIXt"`
- Say "date-time" and move on — lubridate handles the ugliness in week 4
:::

## 1.3 · Three families: quantitative, categorical, other [6 min]{.timing}

R's types sort into three families you'll use to pick plots and summaries all
term:

| Family | Question it answers | R types | Cal Poly examples |
|---|---|---|---|
| **Quantitative** | *how much? how many?* — you can average it | numeric, integer | latte price (`4.75`), Rec Center headcount (`38`), miles from campus (`2.3`) |
| **Categorical** | *which kind? which group?* — you count it | character, factor, logical | coffee shop (`"Julian's"`, `"Starbucks"`, `"Scout"`), time slot (`"morning"`), iced (`TRUE`) |
| **Other** (time/date) | *when?* — ordered, but not a plain number | Date, date-time | the day you bought the latte, the clock time you swiped into the Rec |

> *Q2. Which family: (a) the number of people in the Rec Center at 5 pm ·
> (b) the shop that made your latte · (c) the date of the game vs. UCSB ·
> (d) whether you bought a latte today (yes/no)?*
>
> *Your answer:* (a) quantitative · (b) categorical · (c) other — a date ·
> (d) categorical — a logical is a two-level category.

::: {.callout-tip title="Q2 answer + why the families matter"}
- [Q2]{.q} (a) **quantitative** (b) **categorical** (c) **other / date** (d) **categorical** — logical is a yes/no category
- Why bother: quantitative → histograms and means; categorical → bar charts and counts; dates → line plots over time. Week 3's plots pick themselves once you name the family
- Trap to name now: a number can be categorical. Your student ID is digits, but averaging student IDs is nonsense
:::

**A small distinction — "string" vs "character".** Other languages call text a
*string*. R calls the *type* **character**, and one piece of text (`"Scout
Coffee"`) is a **string** — a character value of length one. So `"latte"` is a
string, its class is character, and `c("latte", "mocha")` is a character
*vector* of two strings. You'll hear both words; they are not in conflict.

## 1.4 · Factors vs characters — why it matters [6 min]{.timing}

**LM4 — the same text, two types (4 min).** Rec Center headcounts come with a
time slot. As character, the slots are just words; as a **factor**, they come
with a fixed menu of allowed values (**levels**) *in an order you choose*:

```{r}
# LM4: character vs factor
slot_chr <- c("evening", "morning", "afternoon", "morning", "evening", "evening")
slot_fct <- factor(slot_chr, levels = c("morning", "afternoon", "evening"))

class(slot_chr)
class(slot_fct)
levels(slot_fct)

table(slot_chr)   # alphabetical — afternoon first
table(slot_fct)   # the order YOU chose — morning first
```

> 👀 **Variables check.** `slot_chr` and `slot_fct` are both listed — same six
> values, but Variables describes one as *character* and one as *factor with
> 3 levels*. That description IS the type.

> *Q3. How many levels does `slot_fct` have, and what order did we give
> them? Why would that order matter on a bar chart?*
>
> *Your answer:* 3 levels — morning, afternoon, evening. A bar chart follows
> the levels, so the factor plots the day in time order; the character
> version plots alphabetically (afternoon, evening, morning), which reads as
> nonsense.

::: {.callout-tip title="Q3 answer — the two reasons factors exist"}
- [Q3]{.q} **3 levels**, in the order morning → afternoon → evening
- **Reason 1 — order.** Characters sort alphabetically; factors sort by *your* levels. Every plot and table of a categorical variable inherits that order
- **Reason 2 — a fixed menu.** A factor knows `"night"` is *not* allowed (it becomes `NA`); a character column will happily accept `"Morning"`, `"morning "`, and `"AM"` as three different groups. Factors are how you catch typos in categories
- Rule of thumb to leave them with: **character = free text (names, comments); factor = a category with a known set of values**
- The menu metaphor pays rent in week 6 when forcats arrives
:::

**LM5 — NA, the missing value (2 min).** `airquality` (built in) has real
missing ozone readings; `is.na()` finds them and `sum()` counts them:

```{r}
# LM5: count the missing values in Ozone
sum(is.na(airquality$Ozone))
```

> *Q4. How many Ozone values are missing? 37*

::: {.callout-tip title="Q4 answer"}
- [Q4]{.q} **37 missing**
- First appearance of the `is.na()` + `sum()` idiom; students use it in every cycle
:::

**LM6 — the is.\* family (2 min).** Test types with functions that return
TRUE/FALSE:

```{r}
# LM6: any correct fills work, e.g.
is.numeric(4.75)
is.character("Scout Coffee")
is.logical(TRUE)
is.factor(slot_fct)
```

> *YT1 (4 min). ➕ Insert a chunk below. In it, find the `class()` of each of
> the following, predicting first: `"TRUE"` (with quotes) · `4.75 > 4.25` ·
> `mtcars$cyl` · `Sys.Date() + 7`. Then in the comment, write which one
> surprised you.*

<!-- ➕ insert a chunk here -->

```{r}
# YT1 — surprise: "TRUE" in quotes is character, and a comparison makes a logical
class("TRUE")
class(4.75 > 4.25)
class(mtcars$cyl)
class(Sys.Date() + 7)
```

> *Predictions: character · logical · numeric · Date*

::: {.callout-note title="Expected — YT1"}
- [YT1]{.yt} `"character"` · `"logical"` · `"numeric"` · `"Date"`
- Surprise #1: `"TRUE"` — quotes make it text (a string!)
- Surprise #2: `4.75 > 4.25` *produces* a logical — comparisons are how logicals get made in practice (`filter()` in two weeks)
- `mtcars$cyl` being numeric though it only takes 4/6/8 previews activity item E: **the type R stores is not always the family the data means**
:::

{{< vspace 1.5rem >}}

# 2 · Data structures [24 min]{.timing}

**LM7 — vector: one type, many values (4 min).** Built with `c()`. Three
places to get a latte near campus:

```{r}
# LM7: a vector
shops <- c("Julian's", "Starbucks", "Scout Coffee")
shops
length(shops)
class(shops)
```

> *YT2 (4 min). ➕ Insert a chunk. Build THREE vectors and print each:*
> *(a) `price` — what a latte costs at each shop above (Julian's on campus,
> the Starbucks in the UU, Scout downtown — use real prices if you know
> them, e.g. 4.75, 5.45, 5.95);*
> *(b) `size` — the size you'd order at each, as text ("small", "medium",
> "large");*
> *(c) `iced` — three TRUE/FALSE values, whether you'd get it iced.*
> *Then run `class()` on each. What happens if you run `c(4.75, "five", 5.95)`?
> Try it and look at the class.*

<!-- ➕ insert a chunk here -->

```{r}
# YT2
price <- c(4.75, 5.45, 5.95)
size  <- c("small", "medium", "large")
iced  <- c(TRUE, FALSE, TRUE)
price; size; iced
class(price); class(size); class(iced)
c(4.75, "five", 5.95)
class(c(4.75, "five", 5.95))
```

> *Q5. `c(4.75, "five", 5.95)` has class character. In one phrase, why?*
>
> *Your answer:* A vector holds one type, so R coerces everything to text.

::: {.callout-tip title="Q5 answer — coercion"}
- [Q5]{.q} **character**
- A vector holds *one* type, so R converts everything to the type that can hold all of it — text
- This is **coercion**, arriving by accident rather than lecture — say the word once
- They meet it again when a price column imports as `<chr>` because one row says "free"
:::

**LM8 — data frame & tibble: vectors standing side by side (6 min).** Each
column is a vector; every column has the same length.

```{r}
# LM8: same data, two containers
latte_df  <- data.frame(shop = shops, price = c(4.75, 5.45, 5.95))
latte_tbl <- tibble(shop = shops, price = c(4.75, 5.45, 5.95))

latte_df
latte_tbl
```

> *Q6. Name one visible difference between how the data frame and the tibble
> printed:*
>
> *Your answer:* The tibble announces its dimensions (`# A tibble: 3 × 2`) and
> shows column types (`<chr>`, `<dbl>`); the data frame shows neither.

::: {.callout-tip title="Q6 model answers (any one)"}
- [Q6]{.q} the tibble announces its dimensions (`# A tibble: 3 × 2`)
- the tibble shows column types (`<chr>`, `<dbl>`) — **the families from §1.3, printed for you**
- the data frame shows row numbers without types
- **Framing:** a tibble IS a data frame with better manners — we'll say "data frame" for both
:::

> *YT3 (5 min). ➕ Insert a chunk. Build a tibble called `lattes` from YOUR
> vectors from YT2 — columns `shop`, `price`, `size`, `iced`. Print it.
> Then run `nrow(lattes)`, `ncol(lattes)`, and `class(lattes)`. What happens if
> one vector has only two values? Try it (make a broken one on purpose), read
> the error, then fix it.*

<!-- ➕ insert a chunk here -->

```{r}
# YT3
lattes <- tibble(shop = shops, price = price, size = size, iced = iced)
lattes
nrow(lattes); ncol(lattes); class(lattes)
```

> 👀 **Variables check.** `lattes` gets an expand arrow that `price` did not.
> Click it: four columns, each with its own type and length 3. A data frame
> is a container of vectors — and the pane shows you exactly that.

```{r}
#| eval: false
# YT3 — the deliberate break
tibble(shop = shops, price = c(4.75, 5.45))
#> Error: Tibble columns must have compatible sizes.
#> • Size 3: Existing data.
#> • Size 2: Column `price`.
```

```{r}
# YT3 — the fix: give price all three values
tibble(shop = shops, price = c(4.75, 5.45, 5.95))
```

::: {.callout-caution title="YT3 teaching note — a good error"}
- [YT3]{.yt} The error names the sizes *and* the column — have someone read it aloud
- "Every column the same length" is now something they've felt, not heard
- `class(lattes)` returns three things (`tbl_df`, `tbl`, `data.frame`) — say "it's a data frame, with extra labels" and move on
:::

**LM9 — tribble: a tibble typed row by row, with a factor inside (3 min).**
Rec Center headcounts, three slots. Note the `slot` column is made a factor
*with the order we want*:

```{r}
# LM9: Rec Center headcounts — a tribble, then a factor column
rec <- tribble(
  ~slot,       ~hour, ~people,
  "morning",     7,     38,
  "afternoon",  13,    112,
  "evening",    18,    164
)
rec$slot <- factor(rec$slot, levels = c("morning", "afternoon", "evening"))
rec
```

::: {.callout-note title="LM9 — read the type tags"}
- `slot` prints as `<fct>`, `hour` and `people` as `<dbl>` — categorical, quantitative, quantitative. The tibble is telling you the families
- `rec$slot <- factor(...)` is the first time they *change* a column's type — one sentence: "we told R this text is a category with an order"
:::

**LM10 — list: the junk drawer (4 min).** A list holds *anything*, even other
structures:

```{r}
# LM10
my_list <- list(name = "week 2", nums = c(1, 2, 3), data = latte_tbl)
my_list
length(my_list)
```

::: {.callout-important title="LM10 — junk-drawer framing"}
- [LM10]{.lm} A list holds anything, including a whole tibble
- `length()` is 3 — three drawers, whatever's inside them
- Plant the flag: *"when we pull data from an API in week 8, it arrives as a list"*
:::

> *YT4 (3 min). ➕ Insert a chunk. Build a list called `me` with three named
> parts: your first name (text), the number of classes you're taking this
> quarter, and your `lattes` tibble. Print it. Notice how each part keeps its
> own type.*

<!-- ➕ insert a chunk here -->

```{r}
# YT4 — e.g.
me <- list(first = "Eman", n_classes = 4, lattes = lattes)
me
```

**LM11 — matrix: all one type, rows × columns (2 min)** (brief):

```{r}
# LM11
matrix(1:6, nrow = 2)
```

> *Q7. Which structure could hold a column of shop names AND a column of
> prices: a matrix or a data frame? Why?*
>
> *Your answer:* A data frame — a matrix forces every cell to one type, so the
> prices would be coerced to text.

::: {.callout-tip title="Q7 answer"}
- [Q7]{.q} **a data frame**
- A matrix forces every cell to one type, so names + prices would coerce the prices to text — same coercion as Q5, now at the structure level
:::

{{< vspace 1.5rem >}}

# 2b · When things break [6 min]{.timing}

Today's errors are *type* and *structure* errors. Same rule as last week:
read the message slowly, name the family, then fix it.

> *YT5 (6 min). Each chunk below has ONE bug. Read the error (or the silent
> wrong answer!), fix it in place, and write the error family in the comment.*

```{r}
#| eval: false
# YT5a — error family: ____
price <- c(4.75, "5.45", 5.95)
mean(price)
```

```{r}
# YT5a — error family: coercion — one quoted number turned the whole vector to character; mean() warns and returns NA
price <- c(4.75, 5.45, 5.95)
mean(price)
```

```{r}
#| eval: false
# YT5b — error family: ____
tibble(shop = c("Julian's", "Starbucks", "Scout Coffee"), price = c(4.75, 5.45))
```

```{r}
# YT5b — error family: column lengths don't match (size 3 vs size 2) — every column must be the same length
tibble(shop = c("Julian's", "Starbucks", "Scout Coffee"), price = c(4.75, 5.45, 5.95))
```

```{r}
#| eval: false
# YT5c — error family: ____
factor(c("morning", "Morning", "evening"), levels = c("morning", "afternoon", "evening"))
```

```{r}
# YT5c — error family: a value not on the factor's menu becomes NA (no error!) — "Morning" ≠ "morning"
factor(c("morning", "morning", "evening"), levels = c("morning", "afternoon", "evening"))
```

::: {.callout-caution title="YT5 answers — three error families for types & structures"}
- [YT5a]{.yt} `c(4.75, "5.45", 5.95)` → the vector is **character**; `mean()` warns *"argument is not numeric or logical: returning NA"*
  - family: **coercion** — one text value drags the whole vector to text. The tell is the quotes in the printout
  - fix: remove the quotes (later: `as.numeric()`)
- [YT5b]{.yt} `Tibble columns must have compatible sizes` — **size 3 vs size 2**
  - family: **mismatched column lengths** — the error names the sizes *and* the column
  - fix: give every column the same number of values
- [YT5c]{.yt} no error at all — `"Morning"` prints as `<NA>`
  - family: **value not in the factor's levels** → silently becomes `NA` (the sneakiest of the three; the menu metaphor from §1.4 explains it)
  - fix: match the level exactly (`"morning"`) — R is case-sensitive
- Two of the three are *silent*: no red text. Say it: **the absence of an error is not the presence of a right answer** — print your objects
:::

{{< vspace 1.5rem >}}

# 3 · In-class activity: type it or contain it [12 min]{.timing}

For each item, write **T** if the question is about a type and **S** if it's
about a structure, then answer it with code. ➕ Insert **one chunk per item**
(A through F) so each answer is its own runnable piece.

- **A.** What is `rec` (from LM9) — its structure?
- **B.** What is `rec$people` — its type? Which family?
- **C.** Is there any NA anywhere in `airquality$Solar.R`? How many?
- **D.** Build a tibble with 2 columns: three places you eat near campus and
  a rating (1–10) for each.
- **E.** What is `airquality$Month` — its type? Should it be? (Months are
  stored as numbers here — would a factor make more sense? One sentence.)
- **F.** Build a list holding your food tibble from D *and* a character
  vector of what you order at each place.

::: {.callout-important title="Teaching note — one chunk per item"}
- Check that students actually made **six chunks**
- The habit of "one idea, one chunk" is what makes ETV documents readable
:::

<!-- ➕ insert chunks A–F here -->

```{r}
# A — Structure
class(rec)          # a tibble (tbl_df / tbl / data.frame)
```

```{r}
# B — Type
class(rec$people)   # "numeric" — quantitative
```

```{r}
# C — Type question (is.na tests values)
sum(is.na(airquality$Solar.R))   # 7
```

```{r}
# D — Structure built by the student, e.g.
eats <- tibble(
  place  = c("Woodstock's", "Firestone", "Campus Market"),
  rating = c(9, 10, 6)
)
eats
```

```{r}
# E — Type; stored as integer, arguably should be a factor (or a Date)
class(airquality$Month)
```

```{r}
# F — Structure: a list holding a tibble and a vector
eats_list <- list(
  places = eats,
  orders = c("pizza", "tri-tip sandwich", "burrito")
)
eats_list
```

> *A: S · B: T (numeric, quantitative) · C: T (7 missing) · D: S · E: T —
> integer; a factor would make more sense since 5–9 are labels, not amounts ·
> F: S*

::: {.callout-tip title="Activity answers"}
- **A** = S (structure — a tibble)
- **B** = T — `"numeric"`, quantitative
- **C** = T — 7 missing
- **D** = S
- **E** = T — `"integer"`; a factor would make sense since 5–9 are labels, not amounts
- **F** = S
- **E is the discussion item:** *the type R stores is not always the family the data means* — that's the whole reason `mutate()` + `factor()` exists, and it's §1.4 again
:::

{{< vspace 1.5rem >}}

# 4 · Exit question [5 min]{.timing}

> *In your own words, one sentence each: what is a data type, and what is a
> data structure? Use one example of each from today — and at least one must
> be something YOU built.*
>
> *Your answer:* A data type is what one value is — `4.75` (a Julian's
> latte) is numeric. A data structure is the container holding values — my
> `lattes` tibble is a data frame whose `iced` column is logical.

::: {.callout-tip title="Model answer"}
- *"A data type is what one value is — `4.75` is numeric. A data structure is the container holding values — my `lattes` tibble is a data frame whose `iced` column is logical."*
- Passes: any answer pairing one correct **type** example with one correct **structure** example
- At least one example should be from a student-built object
- Bonus signal: a student who names the *family* (quantitative/categorical) alongside the type has the whole lesson
:::

# 5 · Save & render [1 min]{.timing}

Save this file into your `R/Notes/` folder as `fnd02_yourlastname.qmd`. Then
**Restart R** (top-right session menu → Restart) — watch Variables empty out —
and press **Render** (Cmd/Ctrl + Shift + K). If a chunk you inserted breaks
the render, read the error: it names the chunk. Restarting first is how you
prove that everything the document needs is *in* the document.

::: {.callout-important title="Teaching note — restart, then render"}
- Same close as Fnd 01 and Lab 01: **restart → Variables empties → render**. The 10-second rebuild is the argument for code-in-the-file
- A render that fails on a student-inserted chunk is a *good* failure: the error names the chunk — have them read it aloud
:::
