---
title: "Foundations 02 — Data Types and Data Structures"
subtitle: "STAT/DATA 1810 · Week 2 Monday · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

<!--
CLO: 1.5, 1.6, 1.4 · 7.6 preview · 9 practiced (students insert chunks)
Timing: distinction+chunks 8' · types 22' · structures 30' · activity 15'
        · exit 5' = 80'
Emphasis per spec: vectors, data frames/tibbles, lists get practice;
matrix & tribble get one chunk each, no drills.
Checkpoint: §2 should start by the 30' mark. If behind, drop YT4 and E/F.
-->

```{r}
#| label: setup
library(tidyverse)
```

# 0 · The one distinction, and your own chunks (8 min)

Write the sentence on the board and leave it there all class:
**type = one value's nature; structure = the container.**

**Teaching note — inserting chunks.** Demonstrate the shortcut
(Cmd/Ctrl + Shift + I) *once*, slowly, on the projector: cursor on a blank
line, press it, the fence appears, type `1 + 1`, run. Then YT0 — walk the
room for 2 minutes and make sure every screen shows output under a
student-made chunk before moving on. This is the one thing today that can't
be caught up on later; every later YT depends on it. Students who type the
fence by hand are fine — the skill is "I can make a chunk," not the shortcut.

```{r}
# YT0
1 + 1
```

# 1 · Data types (22 min)

```{r}
# LM1 — expected: "numeric", "integer", "character", "logical"
class(4.75)
class(4L)
class("latte")
class(TRUE)
```

**Q1 note:** most students predict `"numeric"` for `4L` — good; the surprise
is the point. The `L` suffix is worth 30 seconds, no more.

```{r}
# LM2 — "double" vs "integer"
typeof(4.75)
typeof(4L)
```

```{r}
# LM3 — factor with 3 levels
class(iris$Species)
levels(iris$Species)
```

**Q2:** 3 levels — setosa, versicolor, virginica. **Teaching note:** the menu
metaphor ("a factor is text with a fixed menu") pays rent in week 6 when
forcats arrives.

```{r}
# LM4
today <- Sys.Date()
class(today)
right_now <- Sys.time()
class(right_now)
```

Expected classes: `"Date"` and `"POSIXct" "POSIXt"` — say "date-time" and
move on; lubridate handles the ugliness in week 4.

```{r}
# LM5
sum(is.na(airquality$Ozone))
```

**Q3: 37 missing.** Real 1973 NYC air data — the missing values are genuinely
missing, not planted. First appearance of the `is.na()` + `sum()` idiom;
students will use it in every cycle.

```{r}
# LM6 — any correct fills, e.g.
is.numeric(4.75)
is.character("latte")
is.logical(TRUE)
is.factor(iris$Species)
```

```{r}
# YT1 — expected: "character", "logical", "numeric", "Date"
class("TRUE")
class(3 > 2)
class(mtcars$cyl)
class(Sys.Date() + 7)
```

**YT1 teaching note:** the surprise is almost always `"TRUE"` — quotes make
it text. Second surprise: `3 > 2` *produces* a logical; comparisons are how
logicals get made in practice (filter() in two weeks). `mtcars$cyl` being
numeric though it only takes 4/6/8 is a preview of activity item E.

# 2 · Data structures (30 min)

```{r}
# LM7
drinks <- c("latte", "drip", "mocha")
drinks
length(drinks)
class(drinks)
```

```{r}
# YT2
prices <- c(4.75, 2.75, 5.25)
sizes  <- c("small", "medium", "large")
iced   <- c(TRUE, FALSE, TRUE)
prices; sizes; iced
class(prices); class(sizes); class(iced)
c(1, "two", 3)
class(c(1, "two", 3))
```

**Q4: character.** A vector holds *one* type, so R converts everything to
the type that can hold all of it — text. This is **coercion**, arriving by
accident rather than lecture. Say the word once; students meet it again
when a numeric column imports as `<chr>` because of one stray "N/A".

```{r}
# LM8
menu_df  <- data.frame(drink = drinks, price = c(4.75, 2.75, 5.25))
menu_tbl <- tibble(drink = drinks, price = c(4.75, 2.75, 5.25))
menu_df
menu_tbl
```

**Q5 model answers (any one):** the tibble announces its dimensions
(`# A tibble: 3 × 2`); the tibble shows column types (`<chr>`, `<dbl>`); the
data frame shows row numbers without types. **Framing:** a tibble IS a data
frame with better manners — we'll say "data frame" for both.

```{r}
# YT3
menu <- tibble(drink = drinks, price = prices, size = sizes, iced = iced)
menu
nrow(menu); ncol(menu); class(menu)
```

```{r}
#| eval: false
# YT3 — the deliberate break
tibble(drink = drinks, price = c(4.75, 2.75))
#> Error: Tibble columns must have compatible sizes.
#> • Size 3: Existing data.
#> • Size 2: Column `price`.
```

**YT3 teaching note:** the error is a *good* error — it names the sizes and
the column. Have someone read it aloud. "Every column the same length" is
now something they've felt, not heard. (`class(menu)` returns three things —
`tbl_df`, `tbl`, `data.frame` — say "it's a data frame, with extra labels"
and move on.)

```{r}
# LM9 — tribble, briefly
tribble(
  ~drink,  ~price,
  "chai",   4.50,
  "espresso", 3.00
)
```

```{r}
# LM10 — list
my_list <- list(name = "week 2", nums = c(1, 2, 3), data = menu_tbl)
my_list
length(my_list)
```

**Junk-drawer framing:** a list holds anything, including a whole tibble.
`length()` is 3 — three drawers, whatever's inside them. Plant the flag:
*"when we pull data from an API in week 8, it arrives as a list."*

```{r}
# YT4 — e.g.
me <- list(first = "Eman", n_classes = 4, menu = menu)
me
```

```{r}
# LM11 — matrix, briefly
matrix(1:6, nrow = 2)
```

**Q6:** a data frame — a matrix forces every cell to one type, so names +
numbers would coerce the numbers to text (same coercion as Q4, now at the
structure level).

# 3 · In-class activity (15 min)

One chunk per item — check that students actually made six chunks; the
habit of "one idea, one chunk" is what makes ETV documents readable.

```{r}
# A — Structure
class(iris)          # "data.frame"
```

```{r}
# B — Type
class(iris$Sepal.Length)   # "numeric"
```

```{r}
# C — Type question (is.na tests values)
sum(is.na(airquality$Solar.R))   # 7
```

```{r}
# D — Structure built by the student, e.g.
foods_tbl <- tibble(
  food   = c("tacos", "pho", "pizza"),
  rating = c(10, 9, 8)
)
foods_tbl
```

```{r}
# E — Type; stored as integer, arguably should be a factor (or a Date)
class(airquality$Month)
```

```{r}
# F — Structure: a list holding a tibble and a vector
food_places <- list(
  foods  = foods_tbl,
  places = c("truck on Higuera", "home", "campus")
)
food_places
```

**Answers: A = S · B = T · C = T (7 missing) · D = S · E = T ("integer";
a factor would make sense since 5–9 are labels, not amounts) · F = S.**
E is the discussion item: *the type R stores is not always the type the
data means* — that's the whole reason `mutate()` + `factor()` exists.

# 4 · Exit question (5 min)

**Model answer:** "A data type is what one value is — `4.75` is numeric. A
data structure is the container holding values — my `menu` tibble is a data
frame whose `iced` column is logical." Any answer pairing one correct type
example with one correct structure example passes; at least one example
should be from a student-built object.
