---
title: "Building the California NIBRS 2025 Teaching Dataset"
subtitle: "Instructor build script — run once per year, not in class"
author: "STAT/DATA 1810"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
execute:
  eval: false
  warning: false
  message: false
---

## What this file does and does not do

This is a **build script**, not a lesson. It runs once, on your machine, and
produces two `.rda` files that students load in a single line. Students never
see the 500 MB ZIP, never touch SQL, and never wait ten minutes for a database
to build.

| Stage | Who runs it | Runtime | Output |
|---|---|---|---|
| 0. Download master file | you | ~15 min | `NIBRS/nibrs-2025.zip` (~500 MB) |
| 1. Build SQLite database | you | ~20–40 min | `NIBRS/nibrs2025.db` (~5 GB) |
| 2. Extract California | you | ~2–5 min | data frames in memory |
| 3. Transform + save | you | ~1 min | `data/ca_crime_2025.rda` (teaching) |
| 4. Document | you | seconds | `ca_crime_2025_dictionary.csv` |

The pipeline follows Greg Ridgeway's *R4crim* Chapter 5
(<https://gregridgeway.github.io/R4crim/05_Working_with_NIBRS_data.html>),
which is CC-BY-SA licensed. Stages 1 and 2 are adapted from it; stages 3 and 4
are new.

::: {.callout-important}
## Three things to know before you commit to this dataset

1. **Recency ceiling.** The FBI accepts NIBRS submissions through March of the
   following year. The 2025 master file is the most recent complete year
   available, and it was finalized in March 2026. If a cycle needs *this
   month's* crime, NIBRS is the wrong source — use a city open-data portal
   (LA, SF, San Diego all publish weekly).

2. **California coverage is incomplete.** Ridgeway's Chapter 5 reports roughly
   80% of California's population covered by NIBRS-reporting agencies, one of
   the lowest coverage rates in the country. This is a limitation *and* the
   single best CLO 12 teaching hook in your entire dataset spine — see
   §7 below.

3. **Arrest data is right-censored.** A January offense had 14 months to
   produce a recorded arrest; a December offense had 3. Any raw "clearance
   rate by month" chart will slope downward for reasons that have nothing to
   do with policing. The fixed 60-day window in §5 handles this.
:::

## 0. Setup

```{r}
#| label: setup

library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(RSQLite)

# ---- configuration knobs -------------------------------------------------

nibrs_dir  <- "NIBRS"                        # where the zip + lookups live
db_path    <- file.path(nibrs_dir, "nibrs2025.db")
out_dir    <- "data"

rebuild_db <- FALSE   # TRUE only the first time, or after a new data year
data_year  <- 2025
ca_code    <- 4       # numeric state code for California in NIBRS

dir.create(out_dir, showWarnings = FALSE)
```

### 0a. Fetch the lookup tables (automatic)

The eight lookup CSVs live in the `NIBRS/` folder of
<https://github.com/gregridgeway/R4crim>. Filenames and column headers below
were verified against the repo. This chunk downloads them; it skips anything
already present, so it is safe to re-run.

```{r}
#| label: fetch-lookups
#| eval: true

dir.create(nibrs_dir, showWarnings = FALSE)

r4crim_raw <- "https://raw.githubusercontent.com/gregridgeway/R4crim/master/NIBRS"

lookup_files <- c(
  "nibrs_format.csv",              # column widths/types for all 8 segments
  "nibrs_offense_lookup.csv",      # ucr_code, crime_cat, crime
  "nibrs_state_lookup.csv",        # state_code, state_abbr, state_name
  "nibrs_location_lookup.csv",     # location_code, location_type
  "nibrs_weapon_lookup.csv",       # weapon_code, weapon_type
  "nibrs_property_lookup.csv",     # property_code, property_type
  "nibrs_drug_lookup.csv",         # drug_code, drug_type
  "nibrs_circumstance_lookup.csv"  # circumstance_code, circumstance
)

for (f in lookup_files) {
  dest <- file.path(nibrs_dir, f)
  if (!file.exists(dest)) {
    message("Downloading ", f)
    download.file(file.path(r4crim_raw, f), destfile = dest, quiet = TRUE)
  }
}

# confirm they parsed
map_dfr(lookup_files, function(f) {
  d <- read.csv(file.path(nibrs_dir, f))
  tibble(file = f, rows = nrow(d), cols = paste(names(d), collapse = ", "))
})
```

Expected output — if any row disagrees, Ridgeway has revised the notes and the
`load_lookup()` calls in §1 need their column names updated to match:

| file | rows | cols |
|---|---|---|
| `nibrs_format.csv` | 235 | col_name, col_type, col_width, segment |
| `nibrs_offense_lookup.csv` | 63 | ucr_code, crime_cat, crime |
| `nibrs_state_lookup.csv` | 56 | state_code, state_abbr, state_name |
| `nibrs_location_lookup.csv` | 45 | location_code, location_type |
| `nibrs_weapon_lookup.csv` | 17 | weapon_code, weapon_type |
| `nibrs_property_lookup.csv` | 68 | property_code, property_type |
| `nibrs_drug_lookup.csv` | 18 | drug_code, drug_type |
| `nibrs_circumstance_lookup.csv` | 17 | circumstance_code, circumstance |

### 0b. Download the master file (manual, ~500 MB)

This one cannot be scripted — the FBI serves it behind a JavaScript download
page with no stable direct link.

1. Go to the Crime Data Explorer downloads page:
   <https://cde.ucr.cjis.gov/LATEST/webapp/#/pages/downloads>
2. Scroll to **Master File Downloads**.
3. Choose **National Incident-Based Reporting System (NIBRS)** from the
   dropdown, select **2025**, download.
4. Put `nibrs-2025.zip` in your `NIBRS/` folder. **Leave it zipped** — the
   loader reads the compressed file directly via `unz()`.

```{r}
#| label: check-master
#| eval: true

zip_path <- file.path(nibrs_dir, "nibrs-2025.zip")

if (!file.exists(zip_path)) {
  stop("Still need the master file.\n",
       "  Download nibrs-2025.zip from https://cde.ucr.cjis.gov/LATEST/webapp/#/pages/downloads\n",
       "  (Master File Downloads -> NIBRS -> 2025)\n",
       "  and save it to: ", normalizePath(nibrs_dir, mustWork = FALSE))
}

# confirm the inner filename matches what the loader expects
unzip(zip_path, list = TRUE)
```

::: {.callout-warning}
## Check the inner filename

The loader in §1 hardcodes `2025_NIBRS_NATIONAL_MASTER_FILE.txt`. The FBI has
changed this naming between years. If `unzip(..., list = TRUE)` shows
something different, update `master_txt` below to match.
:::

```{r}
#| label: master-txt

master_txt <- "2025_NIBRS_NATIONAL_MASTER_FILE.txt"
```

## 1. Build the SQLite database

Adapted from R4crim §5–6. Reads the 5.8 GB fixed-width master file one million
lines at a time, splits each batch by segment code, and appends to eight
tables. The whole load happens inside one transaction — without
`dbBegin()`/`dbCommit()` this takes an order of magnitude longer.

```{r}
#| label: build-db
#| eval: false

nibrs_format <- read.csv(file.path(nibrs_dir, "nibrs_format.csv"))

con <- dbConnect(SQLite(), dbname = db_path)

seg_tables <- list("01" = "admin",    "02" = "offense",  "03" = "property",
                   "04" = "victim",   "05" = "offender", "06" = "arrestee",
                   "07" = "groupb_arrest", "BH" = "batch_header")

if (rebuild_db) {

  for (tbl in unlist(seg_tables)) {
    if (dbExistsTable(con, tbl)) dbRemoveTable(con, tbl)
  }

  infile <- unz(zip_path, master_txt, open = "r")

  dbBegin(con)
  lines_read <- 0

  while (length(a <- readLines(infile, n = 1e6)) > 0) {

    lines_read <- lines_read + length(a)
    message("Lines read: ", format(lines_read, big.mark = ","))

    d_split <- split(a, substring(a, 1, 2))

    # window segments are partial reports; fold them into their parent segment
    d_split[["01"]] <- c(d_split[["01"]], d_split[["W1"]])
    d_split[["03"]] <- c(d_split[["03"]], d_split[["W3"]])
    d_split[["06"]] <- c(d_split[["06"]], d_split[["W6"]])
    d_split[c("W1", "W3", "W6")] <- NULL

    for (seg in names(d_split)) {
      fmt <- nibrs_format |> filter(segment == seg)

      df <- read_fwf(I(d_split[[seg]]),
                     col_positions = fwf_widths(fmt$col_width),
                     col_type      = paste(fmt$col_type, collapse = ""),
                     show_col_types = FALSE) |>
        rename_with(~ fmt$col_name) |>
        rename(ori     = originating_agency_identifier_ori,
               segment = segment_level,
               state   = numeric_state_code)

      if (seg == "03") {
        df <- df |> mutate(value_of_property = as.numeric(value_of_property))
      }
      if (seg == "BH") {
        df <- df |> mutate(total_pop = rowSums(across(starts_with("current_pop")),
                                               na.rm = TRUE))
      }

      dbWriteTable(con, seg_tables[[seg]], df, append = TRUE)
    }
  }

  dbCommit(con)
  close(infile)
}

dbListTables(con)
```

### Load the lookup tables

```{r}
#| label: lookups
#| eval: false

load_lookup <- function(con, table, file, key, cols) {
  dbExecute(con, paste0("DROP TABLE IF EXISTS ", table))
  a <- read.csv(file.path(nibrs_dir, file), colClasses = "character")
  dbExecute(con, sprintf(
    "CREATE TABLE %s (%s TEXT NOT NULL PRIMARY KEY, %s)",
    table, key, paste(paste(cols, "TEXT"), collapse = ", ")))
  dbAppendTable(con, table, a)
  nrow(a)
}

if (rebuild_db) {
  load_lookup(con, "offense_lookup", "nibrs_offense_lookup.csv",
              "ucr_code", c("crime_cat", "crime"))
  load_lookup(con, "location_lookup", "nibrs_location_lookup.csv",
              "location_code", "location_type")
  load_lookup(con, "weapon_lookup", "nibrs_weapon_lookup.csv",
              "weapon_code", "weapon_type")
  load_lookup(con, "property_lookup", "nibrs_property_lookup.csv",
              "property_code", "property_type")
  load_lookup(con, "circumstance_lookup", "nibrs_circumstance_lookup.csv",
              "circumstance_code", "circumstance")

  # state lookup keyed on an integer
  dbExecute(con, "DROP TABLE IF EXISTS state_lookup")
  a <- read.csv(file.path(nibrs_dir, "nibrs_state_lookup.csv"),
                colClasses = "character")
  dbExecute(con, "CREATE TABLE state_lookup (
                    state_code INTEGER NOT NULL PRIMARY KEY,
                    state_abbr TEXT, state_name TEXT)")
  dbAppendTable(con, "state_lookup", a)
}
```

### Indexes

Without these the California extract scans millions of rows per join. Build
once; they persist with the database file.

```{r}
#| label: indexes
#| eval: false

dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_offense_state ON offense(state)")
dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_victim_state  ON victim(state)")
dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_arrestee_state ON arrestee(state)")
dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_offense  ON offense(ori, incident_number)")
dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_victim   ON victim(ori, incident_number)")
dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_arrestee ON arrestee(ori, incident_number)")
dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_bh_ori   ON batch_header(ori)")
```

Confirm California is code 4 before relying on it:

::: {.callout-warning}
## `dbConnect()` creates an empty database if the file is missing

This is SQLite behavior, not a bug — but it means a stray `dbConnect()` call
leaves behind a 0-byte `nibrs2025.db` that looks like a built database and
isn't. Every query against it then fails with `no such table: offense`, which
reads like a code error rather than a missing download. This chunk checks
before trusting it.
:::

```{r}
#| label: confirm-ca

con <- dbConnect(SQLite(), dbname = db_path)

tabs <- dbListTables(con)

if (length(tabs) == 0) {
  dbDisconnect(con)
  stop("'", db_path, "' exists but is empty -- it was created by a stray ",
       "dbConnect() call, not by the loader.\n",
       "  Delete it:  file.remove('", db_path, "')\n",
       "  Then get the master file and run §1 with rebuild_db <- TRUE.")
}

if (!all(c("offense", "victim", "arrestee", "batch_header") %in% tabs)) {
  stop("Database is missing segment tables. Found: ",
       paste(tabs, collapse = ", "), "\n  Re-run §1 with rebuild_db <- TRUE.")
}

if (!"offense_lookup" %in% tabs) {
  stop("Segment tables loaded but lookup tables are missing. ",
       "Re-run the 'lookups' chunk in §1.")
}

dbGetQuery(con, "SELECT * FROM state_lookup WHERE state_abbr = 'CA'")
```

## 2. Extract — the E in ETV

Everything below is one `SELECT`. This is the point worth making to students
when you eventually show them where the data came from: **the extraction step
filtered 70 million records down to California before R ever allocated
memory.** That is a different mental model from `read_csv()`.

### 2a. Agencies

One row per California law enforcement agency. Small, and the key to any
per-capita comparison.

```{r}
#| label: extract-agency

bh_fields <- dbListFields(con, "batch_header")

ca_agency_raw <- dbGetQuery(con, "
  SELECT ori,
         city_name,
         state_abbreviation,
         total_pop,
         agency_nibrs_flag,
         number_of_months_reported,
         covered_by_ori
  FROM batch_header
  WHERE state_abbreviation = 'CA'")

nrow(ca_agency_raw)
```

### 2b. Offenses

One row per **offense record** — not per incident. An incident with three
offenses contributes three rows. Say this out loud in class; it is the single
most common misreading of NIBRS.

`arr` is a subquery collapsing the arrestee table to the earliest arrest per
incident, so an incident with ten arrests does not multiply the offense rows.

```{r}
#| label: extract-offense

ca_offense_raw <- dbGetQuery(con, "
  SELECT o.ori,
         o.incident_number,
         o.incident_date,
         o.ucr_offense_code,
         ol.crime_cat,
         ol.crime,
         o.offense_attempted_completed,
         o.location_type          AS location_code,
         ll.location_type         AS location,
         o.weapon_force_1,
         o.weapon_force_2,
         o.weapon_force_3,
         o.bias_motivation,
         o.suspected_of_using_1,
         bh.city_name,
         arr.first_arrest
  FROM offense AS o
  LEFT JOIN offense_lookup  AS ol ON o.ucr_offense_code = ol.ucr_code
  LEFT JOIN location_lookup AS ll ON o.location_type    = ll.location_code
  LEFT JOIN batch_header    AS bh ON o.ori              = bh.ori
  LEFT JOIN (SELECT ori, incident_number, MIN(arrest_date) AS first_arrest
             FROM arrestee
             GROUP BY ori, incident_number) AS arr
         ON o.ori = arr.ori AND o.incident_number = arr.incident_number
  WHERE o.state = 4")

nrow(ca_offense_raw)
```

::: {.callout-note}
## Why `LEFT JOIN` everywhere

An `INNER JOIN` silently drops offenses whose code is missing from a lookup
table, or whose agency is missing from `batch_header`. You would never see the
loss. `LEFT JOIN` keeps the row and gives you an `NA` you can count. Students
meet this again in the ETV cycle when they tabulate `is.na(crime)`.
:::

### 2c. Individual victims

```{r}
#| label: extract-victim

ca_victim_raw <- dbGetQuery(con, "
  SELECT v.ori,
         v.incident_number,
         v.victim_sequence_number,
         v.type_of_victim,
         v.age_of_victim,
         v.sex_of_victim,
         v.race_of_victim,
         v.ethnicity_of_victim,
         v.ucr_offense_code_1,
         v.relationship_of_victim_to_offender
  FROM victim AS v
  WHERE v.state = 4")

nrow(ca_victim_raw)
```

### 2d. Arrestees

```{r}
#| label: extract-arrestee

ca_arrestee_raw <- dbGetQuery(con, "
  SELECT a.ori,
         a.incident_number,
         a.arrest_date,
         a.type_of_arrest,
         a.ucr_arrest_offense_code,
         ol.crime           AS arrest_crime,
         a.age_of_arrestee,
         a.sex_of_arrestee,
         a.race_of_arrestee,
         a.ethnicity_of_arrestee
  FROM arrestee AS a
  LEFT JOIN offense_lookup AS ol ON a.ucr_arrest_offense_code = ol.ucr_code
  WHERE a.state = 4")

nrow(ca_arrestee_raw)
```

```{r}
#| label: weapon-lookup-to-r

weapon_lookup <- dbGetQuery(con, "SELECT * FROM weapon_lookup")
dbDisconnect(con)
```

## 3. Transform — the T in ETV

Six distinct transformation moves. Each one maps to something already in your
CLO sequence, so this dataset can carry a whole cycle without new machinery.

| Move | What it fixes | CLO |
|---|---|---|
| Parse `YYYYMMDD` → `Date` | dates arrive as text | 3 |
| Recode single-letter codes → labels | `M`/`F`/`U` are unreadable | 3 |
| Collapse three weapon columns → one flag | wide-to-long reasoning | 3 |
| Sentinel missing → `NA` | `99`, `999`, `1` mean "unknown" | 3, 12 |
| Derive `days_to_arrest`, `cleared_60` | censoring-aware measure | 3, 10 |
| Character → factor | storage, ordering, plotting | 1, 3 |

```{r}
#| label: helpers

# NIBRS dates arrive as YYYYMMDD, but Ridgeway's chapter rewrites some of them
# in place to YYYY-MM-DD. Strip hyphens first so this works either way.
parse_nibrs_date <- function(x) {
  as.Date(gsub("-", "", x, fixed = TRUE), format = "%Y%m%d")
}

# Age fields are character and use non-numeric codes for infants/unknowns.
parse_nibrs_age <- function(x) {
  suppressWarnings(as.numeric(ifelse(grepl("^[0-9]{1,3}$", trimws(x)),
                                     trimws(x), NA)))
}

# Firearm weapon codes: 11 (type not stated), 12 (handgun), 13 (rifle),
# 14 (shotgun), 15 (other firearm).
#
# CAREFUL: weapon_force_1 is 2 characters wide, but weapon_force_2 and
# weapon_force_3 are 3 — the trailing character carries the automatic-weapon
# indicator. An exact match on "12" therefore MISSES an automatic handgun
# recorded as "12A" in the second or third slot. Match on the leading two
# digits instead.
firearm_codes <- c("11", "12", "13", "14", "15")

is_firearm <- function(x) {
  !is.na(x) & str_sub(str_trim(x), 1, 2) %in% firearm_codes
}

label_sex <- function(x) {
  factor(dplyr::recode(x, "M" = "Male", "F" = "Female", .default = NA_character_),
         levels = c("Female", "Male"))
}

label_race <- function(x) {
  factor(dplyr::recode(x,
    "W" = "White", "B" = "Black", "I" = "American Indian/Alaska Native",
    "A" = "Asian", "P" = "Native Hawaiian/Pacific Islander",
    "U" = "Unknown", .default = NA_character_))
}

label_ethnicity <- function(x) {
  factor(dplyr::recode(x,
    "H" = "Hispanic or Latino", "N" = "Not Hispanic or Latino",
    "U" = "Unknown", .default = NA_character_))
}
```

### 3a. Agencies

```{r}
#| label: transform-agency

ca_agency <- ca_agency_raw |>
  filter(is.na(covered_by_ori)) |>          # drop agencies reporting through another
  transmute(
    ori,
    agency        = str_to_title(str_trim(city_name)),
    population    = na_if(as.numeric(total_pop), 0),
    reports_nibrs = !is.na(agency_nibrs_flag) & agency_nibrs_flag == "A",
    months_reported = suppressWarnings(as.integer(number_of_months_reported)),
    months_reported = if_else(months_reported %in% 0:12, months_reported,
                              NA_integer_),
    full_year     = !is.na(months_reported) & months_reported == 12
  ) |>
  arrange(desc(population))

glimpse(ca_agency)
```

### 3b. Offenses

```{r}
#| label: transform-offense

ca_offense <- ca_offense_raw |>
  mutate(
    incident_date = parse_nibrs_date(incident_date),
    first_arrest  = parse_nibrs_date(first_arrest),

    agency   = str_to_title(str_trim(city_name)),
    crime    = factor(crime),
    crime_cat = factor(crime_cat),
    location = factor(location),

    completed = factor(
      dplyr::recode(offense_attempted_completed,
                    "C" = "Completed", "A" = "Attempted",
                    .default = NA_character_),
      levels = c("Attempted", "Completed")),

    # three weapon columns collapse to one flag; a shooting is an offense
    # with a firearm code in ANY of the three (see is_firearm() above --
    # prefix match, because slots 2 and 3 are 3 chars wide)
    firearm = is_firearm(weapon_force_1) |
              is_firearm(weapon_force_2) |
              is_firearm(weapon_force_3),

    # bias_motivation 88 = "none"; treat as no bias rather than a category
    bias = !is.na(bias_motivation) &
           !bias_motivation %in% c("88", "99", "  "),

    arrest_made    = !is.na(first_arrest),
    days_to_arrest = as.numeric(first_arrest - incident_date),
    days_to_arrest = if_else(days_to_arrest < 0, NA_real_, days_to_arrest),

    # censoring-safe clearance indicator: same 60-day window for every offense
    cleared_60 = !is.na(days_to_arrest) & days_to_arrest <= 60,

    month = factor(format(incident_date, "%b"), levels = month.abb),
    weekday = factor(weekdays(incident_date),
                     levels = c("Monday","Tuesday","Wednesday","Thursday",
                                "Friday","Saturday","Sunday"))
  ) |>
  # a handful of records carry dates outside the data year
  filter(!is.na(incident_date),
         format(incident_date, "%Y") == as.character(data_year)) |>
  select(ori, agency, incident_number, incident_date, month, weekday,
         crime_cat, crime, ucr_offense_code, completed, location,
         firearm, bias, arrest_made, days_to_arrest, cleared_60)

glimpse(ca_offense)
```

### 3c. Victims

```{r}
#| label: transform-victim

ca_victim <- ca_victim_raw |>
  mutate(
    victim_type = factor(dplyr::recode(type_of_victim,
      "I" = "Individual", "B" = "Business", "F" = "Financial institution",
      "G" = "Government", "R" = "Religious organization", "S" = "Society/Public",
      "L" = "Law enforcement officer", "O" = "Other", "U" = "Unknown",
      .default = NA_character_)),
    age       = parse_nibrs_age(age_of_victim),
    sex       = label_sex(sex_of_victim),
    race      = label_race(race_of_victim),
    ethnicity = label_ethnicity(ethnicity_of_victim),
    victim_seq = suppressWarnings(as.integer(victim_sequence_number))
  ) |>
  select(ori, incident_number, victim_seq, victim_type,
         age, sex, race, ethnicity,
         ucr_offense_code = ucr_offense_code_1,
         relationship = relationship_of_victim_to_offender)

glimpse(ca_victim)
```

### 3d. Arrestees

```{r}
#| label: transform-arrestee

ca_arrestee <- ca_arrestee_raw |>
  mutate(
    arrest_date = parse_nibrs_date(arrest_date),
    arrest_type = factor(dplyr::recode(type_of_arrest,
      "O" = "On-view", "S" = "Summoned/cited", "T" = "Taken into custody",
      .default = NA_character_)),
    arrest_crime = factor(arrest_crime),
    age       = parse_nibrs_age(age_of_arrestee),
    sex       = label_sex(sex_of_arrestee),
    race      = label_race(race_of_arrestee),
    ethnicity = label_ethnicity(ethnicity_of_arrestee)
  ) |>
  select(ori, incident_number, arrest_date, arrest_type,
         ucr_arrest_offense_code, arrest_crime,
         age, sex, race, ethnicity)

glimpse(ca_arrestee)
```

## 4. Size control

The full California offense table is on the order of 1.4 million rows. That is
fine on your machine and painful on 105 student laptops over campus wifi. Two
products:

- `ca_crime_2025_full.rda` — everything, for your own use, **not distributed**
- `ca_crime_2025.rda` — teaching subset, target under ~10 MB

The subset keeps every offense from a named list of agencies. Adjust
`teaching_agencies` to taste — the SLO-area regex is there so the dataset
connects to the SLO crime log cycle students already know.

```{r}
#| label: subset

# top agencies by offense volume, plus anything on the Central Coast
top_agencies <- ca_offense |>
  count(ori, agency, sort = TRUE) |>
  slice_head(n = 12)

central_coast <- ca_agency |>
  filter(str_detect(agency, regex("San Luis Obispo|Paso Robles|Atascadero|Pismo|Arroyo Grande|Morro Bay|Santa Maria|Santa Barbara", ignore_case = TRUE))) |>
  select(ori, agency)

teaching_agencies <- bind_rows(select(top_agencies, ori, agency),
                               central_coast) |>
  distinct(ori, .keep_all = TRUE)

teaching_agencies
```

```{r}
#| label: build-subset

keep_ori <- teaching_agencies$ori

ca_offense_sub  <- ca_offense  |> filter(ori %in% keep_ori) |> droplevels()
ca_victim_sub   <- ca_victim   |> semi_join(ca_offense_sub,
                                            by = c("ori", "incident_number")) |>
                                  droplevels()
ca_arrestee_sub <- ca_arrestee |> semi_join(ca_offense_sub,
                                            by = c("ori", "incident_number")) |>
                                  droplevels()
ca_agency_sub   <- ca_agency   |> filter(ori %in% keep_ori)

tibble(
  object = c("offense", "victim", "arrestee", "agency"),
  rows   = c(nrow(ca_offense_sub), nrow(ca_victim_sub),
             nrow(ca_arrestee_sub), nrow(ca_agency_sub))
)
```

```{r}
#| label: save

# full archive — keep locally, do not push to Canvas or GitHub
save(ca_offense, ca_victim, ca_arrestee, ca_agency,
     file = file.path(out_dir, "ca_crime_2025_full.rda"),
     compress = "xz")

# teaching subset — this is the file students download
ca_offense  <- ca_offense_sub
ca_victim   <- ca_victim_sub
ca_arrestee <- ca_arrestee_sub
ca_agency   <- ca_agency_sub

save(ca_offense, ca_victim, ca_arrestee, ca_agency,
     file = file.path(out_dir, "ca_crime_2025.rda"),
     compress = "xz")

file.size(file.path(out_dir, "ca_crime_2025.rda")) / 1024^2  # MB
```

::: {.callout-tip}
## If the teaching file is still too big

In rough order of what to try:

1. Cut `teaching_agencies` from 12 down to 5 or 6.
2. Drop `ca_victim` — it is the largest table and most cycles do not need it.
3. Drop `incident_number` from `ca_offense` (kills the joins, halves the size).
4. Subsample: `slice_sample(ca_offense, n = 250000)` with a fixed seed.

Do **not** switch to `compress = "gzip"` to save build time. The `xz` pass
costs you a minute once and costs 105 students nothing.
:::

## 5. Verify before you ship

Never distribute a file you have not interrogated. These are the checks worth
running every year.

```{r}
#| label: verify

# 1. Did the lookup joins actually match?
ca_offense |> summarise(missing_crime = mean(is.na(crime)),
                        missing_location = mean(is.na(location)))

# 2. Are dates inside the data year?
range(ca_offense$incident_date)

# 3. Does the seasonal pattern look plausible?
ca_offense |> count(month)

# 4. Censoring check — raw arrest rate should decay late in the year,
#    the 60-day version should not.
ca_offense |>
  group_by(month) |>
  summarise(raw_arrest = mean(arrest_made),
            cleared_60 = mean(cleared_60))

# 4b. Firearm flag — did the prefix match catch anything exact match missed?
#     If these two disagree, the 3-char weapon slots carry automatic
#     indicators and the prefix match is doing real work.
ca_offense_raw |>
  summarise(
    exact_match = sum(weapon_force_2 %in% firearm_codes |
                      weapon_force_3 %in% firearm_codes, na.rm = TRUE),
    prefix_match = sum(is_firearm(weapon_force_2) |
                       is_firearm(weapon_force_3), na.rm = TRUE)
  )

# 5. Coverage — what share of CA population is even in here?
ca_agency |>
  summarise(covered = sum(population[reports_nibrs], na.rm = TRUE),
            not_covered = sum(population[!reports_nibrs], na.rm = TRUE)) |>
  mutate(pct = 100 * covered / (covered + not_covered))
```

If check 4 shows the raw rate sloping down and `cleared_60` roughly flat, the
censoring correction is doing its job — and you now have a ready-made class
demo of why the naive statistic misleads.

## 6. Generate the data dictionary

Auto-generated from the object itself, so it can never drift from the data.
Uses `purrr::map_*`, which lands squarely in your Weeks 10–12 block — this
chunk doubles as an instructor-side example of why `map()` beats a loop.

```{r}
#| label: dictionary

describe_df <- function(df, object_name) {
  tibble(
    object     = object_name,
    variable   = names(df),
    type       = map_chr(df, ~ class(.x)[1]),
    n_missing  = map_int(df, ~ sum(is.na(.x))),
    pct_missing = map_dbl(df, ~ round(100 * mean(is.na(.x)), 1)),
    n_distinct = map_int(df, ~ dplyr::n_distinct(.x, na.rm = TRUE)),
    example    = map_chr(df, ~ {
      v <- .x[!is.na(.x)]
      if (length(v) == 0) NA_character_ else as.character(v[1])
    })
  )
}

data_dictionary <- bind_rows(
  describe_df(ca_offense,  "ca_offense"),
  describe_df(ca_victim,   "ca_victim"),
  describe_df(ca_arrestee, "ca_arrestee"),
  describe_df(ca_agency,   "ca_agency")
)

write_csv(data_dictionary, file.path(out_dir, "ca_crime_2025_dictionary.csv"))

print(data_dictionary, n = Inf)
```

## 7. The CLO 12 material this dataset hands you

Four documented, defensible limitations. Each is a real property of the data,
not a hypothetical — which is what makes them teachable rather than
moralizing.

**Coverage.** Roughly 80% of California's population lives in a
NIBRS-reporting jurisdiction, per Ridgeway's Chapter 5, among the lowest rates
nationally. A statement like "California had *N* robberies in 2025" is
straightforwardly false; the honest version names the covered population as
the denominator. Students can compute the covered share themselves from
`ca_agency` — the ethics point arrives as a `summarise()` result, not a
lecture.

**Right-censoring.** The March 2026 submission deadline means late-year
offenses had less time to produce a recorded arrest. Ask students to plot both
`arrest_made` and `cleared_60` by month and explain the divergence. This is
one of the cleanest "the artifact is in the collection process, not the world"
examples available at this level.

**Reported crime only.** NIBRS captures crimes reported to police. Reporting
rates vary sharply by offense type and victim characteristics, so comparing
raw counts across crime types compares reporting propensity as much as
incidence.

**Sentinel values.** The property segment uses `1` for unknown value, and
several agencies use strings of 9s as informal missing codes. Ridgeway's
chapter documents a stolen automobile recorded at \$729 million. Any mean that
includes these is meaningless; medians and explicit `NA` recoding are the fix.
If you extend this build to include property records, keep that example — an
absurd number is more memorable than a rule.

## 8. Rebuilding next year

1. Download `nibrs-2026.zip` when the FBI posts it (expect ~September 2027).
2. Change `data_year <- 2026`, update the zip filename and inner `.txt` name.
3. Set `rebuild_db <- TRUE`.
4. Re-pull the R4crim lookup CSVs — offense codes get added between years.
5. Run §5 verification and diff the dictionary against last year's.

Step 5 matters most. A silent schema change in the master file shows up as a
spike in `missing_crime` and nowhere else.
