---
title: "ETV Cycle: California Crime, 2025"
subtitle: "Extraction — Transformation — Visualization"
author: "Your name here"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
```

## The question

Every ETV cycle starts with a question, not with data. Ours:

> **Across California's largest police departments, which crimes get solved
> quickly, and which do not?**

By the end you will have an answer, a chart, and — more useful than either — a
clear sense of why the obvious version of the answer is wrong.

## E — Extraction

The data come from the FBI's National Incident-Based Reporting System (NIBRS),
2025. Every crime reported to a participating police department in the United
States, one row per offense.

The national file has about 70 million records and does not fit in R's memory.
Someone else already did the hard extraction: filtered to California, attached
readable labels to the numeric codes, and saved the result as an `.rda` file.
That is a real extraction step even though it happens in one line here — the
work was pulling a specific slice out of something far too large to open.

```{r}
#| label: extract

load("data/ca_crime_2025.rda")
```

`load()` is different from every import function you have used so far. It does
not return anything you assign. It drops objects into your environment by the
names they had when they were saved. Check what arrived:

```{r}
#| label: whats-here

ls()
```

Four data frames:

| Object | One row is... |
|---|---|
| `ca_offense` | one **offense** reported to police |
| `ca_victim` | one victim connected to an incident |
| `ca_arrestee` | one person arrested |
| `ca_agency` | one police department |

```{r}
#| label: glimpse-offense

glimpse(ca_offense)
```

::: {.callout-warning}
## One row is one offense, not one incident

A single incident can involve several offenses. Someone who breaks into a car
and steals a laptop generates two offense rows sharing one `incident_number`.
So `nrow(ca_offense)` counts **offenses**, not crimes-as-events. Get this
backwards and every number downstream is wrong.
:::

```{r}
#| label: offense-vs-incident

ca_offense |>
  summarise(
    offense_records = n(),
    incidents       = n_distinct(paste(ori, incident_number)),
    offenses_per_incident = round(offense_records / incidents, 2)
  )
```

### Your turn — E

1. How many police departments are in `ca_agency`? Which has the largest
   `population`?
2. What is the date range of `ca_offense$incident_date`?
3. Use `count()` to find the five most common values of `crime`.

## T — Transformation

The extraction handed you clean types already — dates are `Date`, categories
are `factor`. Transformation here means **reshaping toward the question**, not
repairing damage.

### Filter to what the question is about

Attempted crimes clear differently from completed ones, and offenses with a
missing crime label cannot be grouped. Both get dropped, and both drops get
counted, because a filter you cannot report is a filter you should not trust.

```{r}
#| label: transform-filter

nrow_before <- nrow(ca_offense)

offenses <- ca_offense |>
  filter(completed == "Completed", !is.na(crime))

tibble(before = nrow_before,
       after  = nrow(offenses),
       dropped_pct = round(100 * (1 - nrow(offenses) / nrow_before), 1))
```

### Group and summarise

```{r}
#| label: transform-summarise

clearance <- offenses |>
  group_by(crime) |>
  summarise(
    n_offenses  = n(),
    pct_arrest  = round(100 * mean(arrest_made), 1),
    pct_60day   = round(100 * mean(cleared_60), 1),
    median_days = median(days_to_arrest, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_offenses >= 500) |>
  arrange(desc(pct_60day))

clearance
```

Two clearance columns, and they disagree. That disagreement is the most
interesting thing in this document.

- `pct_arrest` — did an arrest **ever** get recorded for this incident?
- `pct_60day` — did an arrest get recorded **within 60 days** of the offense?

### Why the two columns differ

The FBI accepted 2025 submissions through March 2026. A January offense had
fourteen months for an arrest to appear in the data. A December offense had
three. Nothing about December policing is worse — the *observation window* is
shorter.

```{r}
#| label: censoring

censoring <- offenses |>
  group_by(month) |>
  summarise(
    `Ever arrested`      = 100 * mean(arrest_made),
    `Arrested within 60 days` = 100 * mean(cleared_60),
    .groups = "drop"
  ) |>
  pivot_longer(-month, names_to = "measure", values_to = "pct")

censoring
```

`pivot_longer()` stacks the two measures into one column so a single
`ggplot()` call can draw both. Reshaping data to fit the *plot* you want, not
the other way around, is most of what transformation is for.

### Your turn — T

1. Rebuild `clearance` grouped by `crime_cat` instead of `crime`. Does the
   ranking hold?
2. Add a `firearm` column to the summary: what fraction of each crime type
   involved a firearm?
3. Filter to a single agency and rerun. Does that department look like the
   state?

## V — Visualization

### Censoring, made visible

```{r}
#| label: plot-censoring
#| fig-width: 8
#| fig-height: 4.5

ggplot(censoring, aes(x = month, y = pct, colour = measure, group = measure)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c("Ever arrested" = "#b45f3f",
                                 "Arrested within 60 days" = "#1d3557")) +
  labs(
    title = "The obvious measure slopes down. The fair one doesn't.",
    subtitle = "California offenses reported to NIBRS, 2025",
    x = NULL, y = "Percent of offenses",
    colour = NULL,
    caption = "Late-year offenses had less time to produce a recorded arrest before the March 2026 cutoff."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold"))
```

The red line falls through the autumn. The navy line does not. The red line is
not measuring police performance; it is measuring how long the FBI waited
before closing the file.

### Which crimes get solved

```{r}
#| label: plot-clearance
#| fig-width: 8
#| fig-height: 7

clearance |>
  slice_max(n_offenses, n = 20) |>
  ggplot(aes(x = pct_60day, y = fct_reorder(crime, pct_60day))) +
  geom_col(fill = "#1d3557") +
  geom_text(aes(label = paste0(pct_60day, "%")),
            hjust = -0.15, size = 3.2) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "60-day clearance rate by offense type",
    subtitle = "20 most common completed offenses, California, 2025",
    x = "Percent arrested within 60 days", y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
```

### Volume against clearance

```{r}
#| label: plot-scatter
#| fig-width: 8
#| fig-height: 5.5

ggplot(clearance, aes(x = n_offenses, y = pct_60day)) +
  geom_point(alpha = 0.7, size = 2.5, colour = "#1d3557") +
  ggrepel::geom_text_repel(aes(label = crime), size = 2.8,
                           max.overlaps = 12, colour = "grey30") +
  scale_x_log10(labels = scales::comma) +
  labs(
    title = "Common crimes are rarely the ones that get solved",
    x = "Number of offense records (log scale)",
    y = "Percent arrested within 60 days"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### Your turn — V

1. Facet the clearance chart by `crime_cat`.
2. Build a heatmap of offense counts by `weekday` and `crime_cat` using
   `geom_tile()`.
3. Join `ca_offense` to `ca_agency` and plot offenses per 1,000 residents by
   department. Which agency's rate surprises you, and what would you need to
   check before believing it?

## What this data cannot tell you

Four limits. Not boilerplate — each one changes what conclusions are
available.

**Not all of California is here.** Roughly a fifth of the state's population
lives in a jurisdiction that did not report to NIBRS in 2025. Every number in
this document describes the reporting jurisdictions, and nothing else. Check
it yourself:

```{r}
#| label: coverage

ca_agency |>
  summarise(
    covered     = sum(population[reports_nibrs], na.rm = TRUE),
    not_covered = sum(population[!reports_nibrs], na.rm = TRUE)
  ) |>
  mutate(pct_covered = round(100 * covered / (covered + not_covered), 1))
```

**Only reported crime.** NIBRS records crimes reported to police. Reporting
rates differ sharply by offense — vehicle theft gets reported because
insurance requires it; sexual assault often does not. Comparing raw counts
across crime types partly compares willingness to report.

**Arrest is not guilt.** `arrest_made` records that someone was arrested. It
says nothing about charging, conviction, or whether the right person was
arrested.

**Agencies count differently.** Two departments facing identical events can
produce different records depending on local recording practice. Cross-agency
comparisons need more care than cross-time comparisons within one agency.

## Wrap-up

Write two or three sentences answering the original question — which crimes
get solved quickly in California, and which do not — and name the one
limitation above that most constrains your answer.

Then state which measure you would report to a city council, `pct_arrest` or
`pct_60day`, and why.
