---
title: "Lab 01 — Foundations: Positron, Objects, and Data Structures"
subtitle: "STAT/DATA 1810 · Take-home (AI permitted) · Student version"
author: "Your name here"
date: today
format:
  html:
    toc: true
    embed-resources: true
  docx: default
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT (Foundations block; see course website → PLOs & CLOs)
  CLO 1  1.2 Positron · 1.3 file management · 1.4 assignment ·
         1.5 data types · 1.6 containers · 1.7 packages vs libraries ·
         1.12 storage magnitudes
  CLO 7  7.4 comparison operators (intro) · 7.6 is.na() (intro)
  CLO 11 11.1 ?function help
  CLO 9  practiced, not graded (this IS a rendered Quarto document)
Dataset: mtcars (built in). Target time: 60–90 min. 15 questions.
HOW TO SUCCEED: fill every blank (____), complete every empty chunk, and
answer every "> Your answer:" prompt. The lab must RENDER cleanly to submit.
-->

# Welcome to your first lab

This lab reviews everything from the first two weeks: Positron, objects,
data types, data structures, and how to look inside data. You may use AI
tools — one question below even requires it — but **you must be able to
render this document yourself**, and the written answers must be yours.

We use `mtcars`, a dataset built into R (32 cars from a 1974 magazine). You
do not need to import anything — it is already there.

```{r}
#| label: setup
library(tidyverse)
```

---

# Part A · Positron & the environment (Q1–Q4)

**Q1.** Fill in the blanks about the Positron interface.

> - The pane where R answers one command at a time is the ________.
> - The pane listing every object you have created this session is the
>   ________ pane.
> - Code you want to keep lives in the ________ (this `.qmd` file).

**Q2.** Assignment. Create an object called `my_name` holding your name as
text, then print it.

```{r}
# Q2: create my_name and print it

```

**Q3.** Predict, then run. First write your prediction, then complete the
chunk and check.

> Prediction for `wheels`: ________

```{r}
# Q3: run this and compare to your prediction
cars_counted <- 4
wheels <- cars_counted * 4
wheels
```

**Q4.** Packages. Fill in the blanks:

> - You run `install.packages("tidyverse")` ________ per computer.
> - You run `library(tidyverse)` ________ you start a new session.

---

# Part B · Data types (Q5–Q8)

**Q5.** Run `class()` on three columns of `mtcars` and report what each is.

```{r}
# Q5: check the class of mpg, cyl, and (using $) one more column you choose
class(mtcars$mpg)
class(mtcars$cyl)
class(____)
```

> Your three answers: ________, ________, ________

**Q6.** Use `help()` **or** `?` to look up a function. Open the help page for
`mean`, then answer from what you read.

```{r}
#| eval: false
# Q6: open the help page (this line does not render output — that's fine)
?mean
```

> From the help page, what does the `na.rm` argument do (one sentence)?
>
> Your answer:

**Q7.** Missing values. `mtcars` has none, so we make a small vector that
does. Use `is.na()` and `sum()` to count the missing values.

```{r}
# Q7: count the NAs in this vector
test_scores <- c(88, 92, NA, 75, NA, 100)
sum(is.na(____))
```

> How many values are missing? ________

**Q8.** Comparison operators return TRUE/FALSE. Complete each so it returns
`TRUE`.

```{r}
# Q8: fill each blank so the result is TRUE
10 > ____
"cat" == ____
5 <= ____
```

---

# Part C · Data structures from scratch (Q9–Q12)

**Q9.** Build a **vector** with `c()`: three of your favorite numbers.

```{r}
# Q9: create a numeric vector called fav_nums

```

**Q10.** Build a **data frame from scratch** using `tibble()`. Make three
drinks and their prices.

```{r}
# Q10: create a tibble called menu with columns `drink` and `price`
menu <- tibble(
  drink = c("____", "____", "____"),
  price = c(____, ____, ____)
)
menu
```

**Q11.** Build a **list** with `list()` holding a name, a number, and your
`menu` tibble from Q10.

```{r}
# Q11: create a list called my_stuff

```

**Q12.** In one sentence, explain the difference between a **data type** and
a **data structure**, using one example of each from this lab.

> Your answer:

---

# Part D · Inspecting data & storage (Q13–Q14)

**Q13.** Inspect `mtcars`. Run each function and answer.

```{r}
# Q13: inspect the dataset
glimpse(mtcars)
dim(mtcars)
```

> - How many rows (cars)? ________
> - How many columns (variables)? ________
> - What does ONE row represent? ________

**Q14.** Storage. Every object takes up memory. Check how big `mtcars` is.

```{r}
# Q14: how many bytes does mtcars occupy?
object.size(mtcars)
```

> About how many kilobytes (KB) is that? ________
> Why does the size of a dataset matter before you import it? (one sentence)
>
> Your answer:

---

# Part E · Working with AI (Q15)

**Q15.** This question requires an AI assistant (ChatGPT, Claude, Positron's
assistant, etc.). You will ask it to write code you have **not** learned yet,
run that code, and explain what it does. The goal is not to understand every
detail — it is to get comfortable producing and reading code you will fully
understand later in the course.

**Step 1 — copy this exact prompt to your AI:**

> *"In R, using the built-in mtcars dataset and the dplyr package, write code
> that groups the cars by their number of cylinders (cyl) and reports the
> average miles per gallon (mpg) for each group. Then add one ggplot2 bar
> chart of the result. Explain nothing — just give me the code."*

**Step 2 — paste the AI's code into the chunk below and run it.** If it errors,
tell the AI the error and ask it to fix it, then paste the working version.

```{r}
# Q15: paste the AI-generated code here and run it

```

**Step 3 — explain what the code did.** For **each chunk of code the AI gave
you**, write one short sentence in plain English describing what that line or
block does. You are not expected to know these functions yet — do your best
from the output and the AI's help.

> Your explanations (one line per piece of code):
>
> 1.
> 2.
> 3.
>
> One thing this code did that you do not yet understand:

---

# Finish: render your lab (required)

Your lab is not complete until it **renders cleanly**. Do this:

1. Save the file.
2. Click **Render**. Render it **twice**: once to **HTML** and once to **Word
   (docx)** — the header at the top of this file is already set up for both.
3. If rendering fails, R will point you to the chunk that broke. Fix it and
   render again. A common cause: an empty required chunk or a leftover `____`.

**Why rendering matters:** a document that renders is a document anyone can
reproduce — your future self, a classmate, a grader. Code that only works on
your screen is not yet science. Rendering cleanly is the habit this whole
course is built on.

**Then:** go to Canvas and take the **Lab 01 Quiz**. You will only be able to
answer it if you completed this lab — the quiz asks for values and answers you
produced above.
