---
title: "Lab 01 — Foundations: Positron, Objects, and Data Structures"
subtitle: "STAT/DATA 1810 · Take-home (AI permitted) · Student version"
author: "Your name here"
date: today
format:
  html:
    toc: true
    embed-resources: true
  docx: default
execute:
  warning: false
  message: false
---

```{=html}
<!--
CLO ALIGNMENT (Foundations block; see course website → PLOs & CLOs)
  CLO 1  1.2 Positron · 1.3 file management · 1.4 assignment ·
         1.5 data types · 1.6 containers · 1.7 packages vs libraries ·
         1.12 storage magnitudes
  CLO 7  7.4 comparison operators (intro) · 7.6 is.na() (intro)
  CLO 11 11.1 ?function help
  CLO 9  practiced, not graded (this IS a rendered Quarto document)
Dataset: mtcars (built in). Target time: 60–90 min. 16 questions.
HOW TO SUCCEED: fill every blank (____), complete every empty chunk, and
answer every "> Your answer:" prompt. The lab must RENDER cleanly to submit.
-->
```

# Welcome to your first lab

This lab reviews everything from the first two weeks: Positron, objects, data types, data structures, and how to look inside data. You may use AI tools — one question below even requires it — but **you must be able to render this document yourself**, and the written answers must be yours.

We use `mtcars`, a dataset built into R (32 cars from a 1974 magazine). You do not need to import anything — it is already there.

**A habit to build this week.** Every time you run a chunk, glance at the **Variables** pane (right side, under **SESSION**). It is your R session's memory made visible: what just appeared, what changed, what did *not* change. You will see a 👀 reminder after several chunks below — do not skip them. By the end of the lab the habit should be automatic.

Before you run anything: look at Variables now. It should say *No variables have been created* (if it doesn't, restart R so you start from a clean session). Remember what empty looks like.

```{r}
#| label: setup
library(MASS) ## Base Package
```

> 👀 **Variables check.** Loading a package did *not* add anything to the Variables pane. Why not? Think about what `library()` does versus what `<-` does. (You'll need this idea in Q4.)

------------------------------------------------------------------------

# Part A · Positron & the environment (Q1–Q4)

**Q1.** Fill in the blanks about the Positron interface.

> -   The pane where R answers one command at a time is the \_\_\_\_\_\_\_\_.
> -   The pane listing every object you have created this session is the \_\_\_\_\_\_\_\_ pane.
> -   Code you want to keep lives in the \_\_\_\_\_\_\_\_ (this `.qmd` file).

**Q2.** Assignment. Create an object called `my_name` holding your name as text, then print it.

```{r}
# Q2: create my_name and print it

```

> 👀 **Variables check.** `my_name` should now be listed. If it isn't, one of two things went wrong — either the chunk did not actually run, or you did not use the assignment arrow. Which was it?

**Q3.** Predict, then run. First write your prediction, then complete the chunk and check.

> Prediction for `wheels`: \_\_\_\_\_\_\_\_

```{r}
# Q3: run this and compare to your prediction
cars_counted <- 4
wheels <- cars_counted * 4
wheels
```

> 👀 **Variables check.** How many objects are listed now? Now change `cars_counted` to `10` in the Console (not in the chunk), and look at Variables again. Did `wheels` change? Explain in one sentence *when* `wheels` got its value.
>
> Your answer:

**Q4.** Packages. Suppose you close Positron and reopen it tomorrow. You try to use a `tidyverse` function and R says `could not find function`. Only one of the two lines below needs to be run again — and the other one does not.

``` r
install.packages("tidyverse")
library(tidyverse)
```

> Which line needs re-running, and why does the other one *not* need it? Reason from where each one leaves its result: on your computer's disk, or in the session's memory (the thing that disappeared when you closed Positron).
>
> Your answer:

------------------------------------------------------------------------

# Part B · Data types (Q5–Q8)

**Q5.** Run `class()` on three columns of `mtcars` and report what each is.

```{r}
# Q5: check the class of mpg, cyl, and (using $) one more column you choose
class(mtcars$mpg)
class(mtcars$cyl)
class(____)
```

> Your three answers: \_\_\_\_\_\_\_\_, \_\_\_\_\_\_\_\_, \_\_\_\_\_\_\_\_
>
> `cyl` only ever takes the values 4, 6, or 8 — it *behaves* like a category. Did R store it that way? What does that tell you about how R decides a column's type?
>
> Your answer:

**Q6.** Use `help()` **or** `?` to look up a function. Open the help page for `mean`, then answer from what you read.

```{r}
#| eval: false
# Q6: open the help page (this line does not render output — that's fine)
?mean
```

> From the help page, what does the `na.rm` argument do (one sentence)? What is its default, and what would `mean()` return with that default if one value were `NA`?
>
> Your answer:

**Q7.** Missing values. `mtcars` has none, so we make a small vector that does. Run the first line and look carefully at what it returns — one result *per value*. Then think: if you added those results up, what would R count? Use that idea to complete the second line so it returns the number of missing values.

```{r}
# Q7: count the NAs in this vector
test_scores <- c(88, 92, NA, 75, NA, 100)
is.na(test_scores)
____(is.na(test_scores))
```

> How many values are missing? \_\_\_\_\_\_\_\_ What is R treating `TRUE` as when it adds them? \_\_\_\_\_\_\_\_

**Q8.** Comparison operators return TRUE/FALSE. Complete each so it returns `TRUE`. Before you run, predict which of the three has *more than one* correct answer.

```{r}
# Q8: fill each blank so the result is TRUE
10 > ____
"cat" == ____
5 <= ____
```

------------------------------------------------------------------------

# Part C · Data structures from scratch (Q9–Q13)

**Q9.** Build a **vector** with `c()`: three of your favorite numbers.

```{r}
# Q9: create a numeric vector called fav_nums

```

> 👀 **Variables check.** Look at how Variables describes `fav_nums` — it tells you the type *and* the length. Is that the same kind of description it gave `my_name`? What's different?

**Q10.** Build a **data frame from scratch** using `tibble()`. Make three drinks and their prices.

```{r}
# Q10: create a tibble called menu with columns `drink` and `price`
menu <- tibble(
  drink = c("____", "____", "____"),
  price = c(____, ____, ____)
)
menu
```

> 👀 **Variables check.** `menu` gets a little expand arrow in Variables that `fav_nums` did not. Click it. What is it showing you, and why does a tibble get one when a vector doesn't?

**Q11.** `data.frame` vs `tibble`. R has had `data.frame()` since the beginning; `tibble()` comes from the tidyverse. Build the *same* menu the old way, then print both and **compare the printouts line by line**.

```{r}
# Q11: same menu, built with data.frame() — copy your three drinks and prices from Q10
menu_df <- data.frame(
  drink = c("____", "____", "____"),
  price = c(____, ____, ____)
)
menu_df
menu
class(menu_df)
class(menu)
```

> Look at the very first line of each printout, and at the line just under the column names. Then look at the two `class()` results.
>
> -   What does the tibble print that the data.frame does not? (two things) \_\_\_\_\_\_\_\_
> -   `class(menu)` lists more than one thing. Is `"data.frame"` in that list? What does that tell you about the relationship between a tibble and a data frame? \_\_\_\_\_\_\_\_
> -   In your own words, describe the difference between a `data.frame` and a `tibble`, and say which one you would rather look at when you meet a dataset with 50 columns — and why.
>
> Your answer:

**Q12.** Build a **list** with `list()` holding a name, a number, and your `menu` tibble from Q10.

```{r}
# Q12: create a list called my_stuff

```

> 👀 **Variables check.** Expand `my_stuff`. You should see three things of three different kinds inside one object. Could `c()` have done that? Try `c("a", 1)` in the Console and look at what happened to the `1`.

**Q13.** In one sentence, explain the difference between a **data type** and a **data structure**, using one example of each from this lab.

> Your answer:

------------------------------------------------------------------------

# Part D · Inspecting data & storage (Q14–Q15)

**Q14.** Inspect `mtcars`. Run each function and answer.

```{r}
# Q14: inspect the dataset
glimpse(mtcars)
dim(mtcars)
```

> -   How many rows (cars)? \_\_\_\_\_\_\_\_
> -   How many columns (variables)? \_\_\_\_\_\_\_\_
> -   What does ONE row represent? \_\_\_\_\_\_\_\_
> -   `glimpse()` and `dim()` gave you the same two numbers. Which one tells you more, and what is the extra information?
>
> 👀 **Variables check.** `mtcars` is *not* in your Variables pane even though you just used it. Where does it live, then? (Hint: think about what `library()` did at the top of this lab, and what package `mtcars` might come with.)

**Q15.** Storage. Every object takes up memory. Check how big `mtcars` is.

```{r}
# Q15: how many bytes does mtcars occupy?
object.size(mtcars)
```

> About how many kilobytes (KB) is that? \_\_\_\_\_\_\_\_
>
> Now reason: this session's Variables pane is where objects live while you work — that is your computer's memory. `mtcars` is 32 rows. Imagine a dataset with 32 *million* rows of the same columns. Roughly how big would it be? Why might it be worth knowing that *before* you try to import it?
>
> Your answer:

------------------------------------------------------------------------

# Part E · Working with AI (Q16)

**Q16.** This question requires an AI assistant (ChatGPT, Claude, Positron's assistant, etc.). You will ask it to write code you have **not** learned yet, run that code, and explain what it does. The goal is not to understand every detail — it is to get comfortable producing and reading code you will fully understand later in the course.

**Step 1 — copy this exact prompt to your AI:**

> *"In R, using the built-in mtcars dataset and the dplyr package, write code that groups the cars by their number of cylinders (cyl) and reports the average miles per gallon (mpg) for each group. Then add one ggplot2 bar chart of the result. Explain nothing — just give me the code."*

**Step 2 — paste the AI's code into the chunk below and run it.** If it errors, tell the AI the error and ask it to fix it, then paste the working version.

```{r}
# Q16: paste the AI-generated code here and run it

```

> 👀 **Variables check.** Did the AI's code create any new objects, or did it print results without saving anything? Look at Variables to decide. If nothing was saved, how would you change the code so the summary table *was* saved as an object? (You know the operator.)

**Step 3 — explain what the code did.** For **each chunk of code the AI gave you**, write one short sentence in plain English describing what that line or block does. You are not expected to know these functions yet — do your best from the output and the AI's help.

> Your explanations (one line per piece of code):
>
> 1.  
>
> 2.  
>
> 3.  
>
> One thing this code did that you do not yet understand:

------------------------------------------------------------------------

# Finish: render your lab (required)

Your lab is not complete until it **renders cleanly**. Do this:

1.  Save the file.
2.  **Restart R** first (top-right session menu → Restart). Watch the Variables pane empty out. Now click **Render**. Render it **twice**: once to **HTML** and once to **Word (docx)** — the header at the top of this file is already set up for both.
3.  If rendering fails, R will point you to the chunk that broke. Fix it and render again. A common cause: an empty required chunk, a leftover `____`, or code that only worked because of something you typed in the Console (which Render never sees).

**Why rendering matters:** a document that renders is a document anyone can reproduce — your future self, a classmate, a grader. Code that only works on your screen is not yet science. Rendering cleanly is the habit this whole course is built on. Restarting *before* rendering is how you prove to yourself that everything the document needs is *in* the document.

**Then:** go to Canvas and take the **Lab 01 Quiz**. You will only be able to answer it if you completed this lab — the quiz asks for values and answers you produced above.