---
title: "Lab 01 — Foundations: Positron, Objects, and Data Structures"
subtitle: "STAT/DATA 1810 · Take-home (AI permitted) · SOLUTION"
author: "Instructor key"
date: today
format:
  html:
    toc: true
    embed-resources: true
  docx: default
execute:
  warning: false
  message: false
---

<!--
CLO: 1.2–1.7, 1.12 · 7.4, 7.6 (intro) · 11.1 · 9 (practiced)
Grading: this .qmd is completion/receipt (rendered cleanly = credit). The
POINTS come from the Canvas quiz, whose answers can only be produced by
completing this file. Values below feed the quiz answer key.
-->

```{r}
#| label: setup
library(tidyverse)
```

# Part A · Positron & the environment

**Q1.** Console · Environment · Source editor.

**Q2.**
```{r}
my_name <- "Immanuel"
my_name
```

**Q3.** Prediction: 16. `wheels` = 16.
```{r}
cars_counted <- 4
wheels <- cars_counted * 4
wheels
```

**Q4.** install.packages() = **once** per computer; library() = **every** time
you start a new session.

# Part B · Data types

**Q5.**
```{r}
class(mtcars$mpg)   # "numeric"
class(mtcars$cyl)   # "numeric"
class(mtcars$hp)    # "numeric" (any column accepted; all mtcars cols are numeric)
```
Accept any third column. Teaching note: every `mtcars` column is numeric —
a fine moment to note that "looks like a category" (cyl = 4/6/8) is still
stored as numeric until we say otherwise.

**Q6.** `na.rm = TRUE` removes `NA` values before computing the mean; the
default `FALSE` returns `NA` if any value is missing.

**Q7.**
```{r}
test_scores <- c(88, 92, NA, 75, NA, 100)
sum(is.na(test_scores))
```
**2 missing.**

**Q8.** Any values satisfying each; e.g.
```{r}
10 > 3
"cat" == "cat"
5 <= 5
```

# Part C · Data structures from scratch

**Q9.**
```{r}
fav_nums <- c(7, 11, 24)
fav_nums
```

**Q10.**
```{r}
menu <- tibble(
  drink = c("latte", "drip", "mocha"),
  price = c(4.75, 2.75, 5.25)
)
menu
```

**Q11.**
```{r}
my_stuff <- list(name = "Immanuel", number = 42, table = menu)
my_stuff
```

**Q12.** Model: "A data type is what one value is (`mtcars$mpg` is numeric); a
data structure is the container holding values (`menu` is a tibble)." Any
answer pairing a correct type example with a correct structure example.

# Part D · Inspecting data & storage

**Q13.**
```{r}
glimpse(mtcars)
dim(mtcars)
```
Rows: **32** · Columns: **11** · One row = **one car model**.

**Q14.**
```{r}
object.size(mtcars)
```
About **7 KB** (~7,208 bytes). Model sentence: file size matters because R
loads data into memory, so a file too large for your computer's memory
cannot be imported (or needs different tools).

# Part E · Working with AI

**Q15.** Representative working answer the AI should produce:

```{r}
mtcars |>
  group_by(cyl) |>
  summarise(avg_mpg = mean(mpg)) |>
  ggplot(aes(x = factor(cyl), y = avg_mpg)) +
  geom_col(fill = "#12263A") +
  labs(x = "Cylinders", y = "Average MPG",
       title = "Fewer cylinders, better gas mileage")
```

Expected result: 4-cyl ≈ 26.7 mpg, 6-cyl ≈ 19.7, 8-cyl ≈ 15.1 — a clear
downward staircase.

**What to look for in their explanations (not exact wording):**
- `group_by(cyl)` — splits the cars into groups by cylinder count
- `summarise(avg_mpg = mean(mpg))` — computes one average mpg per group
- `ggplot(...) + geom_col()` — draws the averages as bars

**Grading Q15:** credit is for *engaging honestly* — running the AI code,
getting it to work, and attempting a plain-English read of each piece. A
correct "one thing I don't understand yet" (e.g., the pipe `|>`, or `factor()`)
is a sign the question worked as intended. This is CLO 11 (adapt via
resources) in its earliest, gentlest form — do not grade for CS mastery.

# Rendering & Canvas

Confirm the student rendered to **both** HTML and docx (the YAML supports
both). The Canvas quiz is the graded artifact; this file is the gate.
