---
title: "Lab 01 — Foundations: Positron, Objects, and Data Structures"
subtitle: "STAT/DATA 1810 · Take-home (AI permitted) · SOLUTION"
author: "Instructor key"
date: today
format:
  html:
    toc: true
    embed-resources: true
  docx: default
execute:
  warning: false
  message: false
---

<!--
CLO: 1.2–1.7, 1.12 · 7.4, 7.6 (intro) · 11.1 · 9 (practiced)
Grading: this .qmd is completion/receipt (rendered cleanly = credit). The
POINTS come from the Canvas quiz, whose answers can only be produced by
completing this file. Values below feed the quiz answer key. 16 questions.
The 👀 "Variables check" prompts are habit-building; grade them for a genuine
attempt, not exact wording. Model responses are given for each.
-->

```{r}
#| label: setup
library(tidyverse)
```

**Setup 👀 check.** `library()` loads functions into the session but creates no
object — nothing is assigned with `<-`, so nothing appears in Variables. (Sets
up the disk-vs-memory idea for Q4.)

# Part A · Positron & the environment

**Q1.** Console · Variables (accept Environment) · Editor (accept Source editor).

**Q2.**
```{r}
my_name <- "Immanuel"
my_name
```
👀 `my_name` appears. If not: chunk never ran (no output either) or `=`/no
arrow was used and it printed instead of assigning.

**Q3.** Prediction: 16. `wheels` = 16.
```{r}
cars_counted <- 4
wheels <- cars_counted * 4
wheels
```
👀 Two (then three, with `my_name`) objects. After `cars_counted <- 10` in the
Console, `wheels` is **still 16** — it got its value at the moment its `<-`
line ran; R objects store values, not formulas. (Same idea as Fnd 01 Q4.)

**Q4.** `library(tidyverse)` needs re-running. `install.packages()` wrote the
package to **disk**, which persists when Positron closes; `library()` loads it
into the **session's memory**, which is wiped every time the session ends.
Once per computer / once per session.

# Part B · Data types

**Q5.**
```{r}
class(mtcars$mpg)   # "numeric"
class(mtcars$cyl)   # "numeric"
class(mtcars$hp)    # "numeric" (any column accepted; all mtcars cols are numeric)
```
Accept any third column. `cyl` is stored as **numeric** even though it acts
like a category — R types a column by *what the values are* (numbers), not by
how we intend to use them. Making it a category is our job (`factor()`, later).

**Q6.** `na.rm = TRUE` removes `NA` values before computing the mean. Default is
`FALSE`, under which a single `NA` makes the whole result `NA`.

**Q7.**
```{r}
test_scores <- c(88, 92, NA, 75, NA, 100)
is.na(test_scores)
sum(is.na(test_scores))
```
**2 missing.** `is.na()` returns one TRUE/FALSE per value; `sum()` treats
`TRUE` as **1** and `FALSE` as 0, so the sum is the count of NAs.

**Q8.** Any values satisfying each; e.g.
```{r}
10 > 3
"cat" == "cat"
5 <= 5
```
The `==` line has exactly one answer (`"cat"`); the other two have infinitely
many.

# Part C · Data structures from scratch

**Q9.**
```{r}
fav_nums <- c(7, 11, 24)
fav_nums
```
👀 Variables shows `fav_nums` as numeric **of length 3**; `my_name` was
character of length 1. Same kind of description (type + length) — a single
value is just a vector of length 1.

**Q10.**
```{r}
menu <- tibble(
  drink = c("latte", "drip", "mocha"),
  price = c(4.75, 2.75, 5.25)
)
menu
```
👀 The expand arrow reveals the **columns** inside `menu` (each with its own
type and length). A tibble is a container of vectors; a plain vector has
nothing inside to expand.

**Q11.** `data.frame` vs `tibble`.
```{r}
menu_df <- data.frame(
  drink = c("latte", "drip", "mocha"),
  price = c(4.75, 2.75, 5.25)
)
menu_df
menu
class(menu_df)
class(menu)
```
- The tibble prints (1) its **dimensions** — `# A tibble: 3 × 2` — and (2) the
  **type of every column** under the column names (`<chr>`, `<dbl>`). The
  data.frame prints neither.
- `class(menu)` is `"tbl_df" "tbl" "data.frame"` — `"data.frame"` **is** in the
  list. A tibble is a data frame (everything that works on a data frame works
  on a tibble) with extra behaviour layered on top.
- Model description: "A tibble is the tidyverse's modern data frame. It holds
  the same thing — columns of equal length — but it prints politely: it shows
  the dimensions and column types, only the first 10 rows, and only the columns
  that fit on screen, whereas a data.frame prints everything. It also never
  silently changes your text to factors or your column names. With 50 columns
  I'd want the tibble — the data.frame would spew every column across the
  console and I'd have no idea what type anything was."

  Credit for: (a) tibble shows dims + column types when printed, (b) a tibble
  *is* a data.frame, (c) a sensible preference with a reason. Bonus
  recognition: tibble prints only 10 rows / fitting columns; no
  stringsAsFactors surprise; no partial matching with `$`. Do not require these.

**Q12.**
```{r}
my_stuff <- list(name = "Immanuel", number = 42, table = menu)
my_stuff
```
👀 Three elements of three different kinds inside one object. `c()` could
**not** do that: `c("a", 1)` gives `"a" "1"` — the number was coerced to text
because a vector holds one type only. A list is the container for mixed types.

**Q13.** Model: "A data type is what one value is (`mtcars$mpg` is numeric); a
data structure is the container holding values (`menu` is a tibble)." Any
answer pairing a correct type example with a correct structure example.

# Part D · Inspecting data & storage

**Q14.**
```{r}
glimpse(mtcars)
dim(mtcars)
```
Rows: **32** · Columns: **11** · One row = **one car model**. `glimpse()` tells
more: every column's **name, type, and first values**; `dim()` is just the two
numbers.

👀 `mtcars` is not in Variables because it was not created in this session — it
ships inside a package (`datasets`, loaded automatically with base R) and is
looked up from there when you use it. Variables shows only what *this session*
created.

**Q15.**
```{r}
object.size(mtcars)
```
About **7 KB** (~7,208 bytes). Scaling reasoning: 32 million rows is one
million × 32 rows → roughly 7 KB × 1,000,000 ≈ **7 GB** (any order-of-magnitude
estimate in the GB range is fine). Objects live in memory (that's what
Variables shows), so a dataset larger than the computer's memory cannot be
imported as-is — knowing the size first tells you whether `read_csv()` will
work or whether you need a different tool.

# Part E · Working with AI

**Q16.** Representative working answer the AI should produce:

```{r}
mtcars |>
  group_by(cyl) |>
  summarise(avg_mpg = mean(mpg)) |>
  ggplot(aes(x = factor(cyl), y = avg_mpg)) +
  geom_col(fill = "#12263A") +
  labs(x = "Cylinders", y = "Average MPG",
       title = "Fewer cylinders, better gas mileage")
```

Expected result: 4-cyl ≈ 26.7 mpg, 6-cyl ≈ 19.7, 8-cyl ≈ 15.1 — a clear
downward staircase.

👀 Most AI answers pipe straight to the plot and save **nothing** — Variables
does not change. Fix: assign the summary, e.g.
`mpg_by_cyl <- mtcars |> group_by(cyl) |> summarise(avg_mpg = mean(mpg))`.
Some AIs already do this; either observation is correct if it matches what
Variables shows.

**What to look for in their explanations (not exact wording):**
- `group_by(cyl)` — splits the cars into groups by cylinder count
- `summarise(avg_mpg = mean(mpg))` — computes one average mpg per group
- `ggplot(...) + geom_col()` — draws the averages as bars

**Grading Q16:** credit is for *engaging honestly* — running the AI code,
getting it to work, and attempting a plain-English read of each piece. A
correct "one thing I don't understand yet" (e.g., the pipe `|>`, or `factor()`)
is a sign the question worked as intended. This is CLO 11 (adapt via
resources) in its earliest, gentlest form — do not grade for CS mastery.

# Rendering & Canvas

Confirm the student rendered to **both** HTML and docx (the YAML supports
both). Students are told to **Restart R before rendering** — that is the
check that nothing depends on Console-only work. The Canvas quiz is the graded
artifact; this file is the gate.
