---
title: "ETV Cycle 03 — [Crime Data: Title TBD]"
subtitle: "STAT/DATA 1810 · In-class activity · Student version (TEMPLATE)"
author: "Your name here"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
  eval: false
---

<!--
TEMPLATE STATUS: structural template only. Instructor fills every [bracketed]
placeholder once a specific crime dataset is chosen. Chunks are eval: false
until the dataset exists; flip execute.eval to true at that point.

CLO ALIGNMENT (planned)
  CLO 2 (2.7 extraction from .RData — a NEW file type: load(), not read_*())
  CLO 3 (3.1 dplyr: select, filter, mutate/recode, group_by, summarize, arrange)
  CLO 4 (4.1 ggplot + geom_point; 4.3 counts; 4.5 defensible claim)
  CLO 12 ([select 4–5 of the categories below])
Time budget: 80 minutes.
-->

# 0 · Context & ethics (10 min)

Crime data are about people at some of the worst moments of their lives, and
about neighborhoods that live with the numbers afterward. Before any code,
discuss **[instructor: keep 4–5]**:

- **Privacy & confidentiality:** [Can individuals — victims or accused — be
  identified from location, time, and offense together?]
- **Geographic stigmatization:** [What happens to a neighborhood when its dot
  density becomes its reputation?]
- **Underreporting / inconsistent reporting:** [Which offenses are least
  reported, and how does that bend every count we compute?]
- **Bias in policing or data collection:** [Do these records measure crime, or
  measure *enforcement*?]
- **Responsible communication:** [What must accompany a crime chart before it
  is shared outside this room?]

> *Write one sentence on the consideration that most changes how you'd present
> this analysis:*

# 1 · The central question

> **[Insert ONE focused research question about the crime data — e.g., "How
> does the count of [offense type] vary by [hour of day / month / category]?"]**

# 2 · E — Extraction (12 min)

**LM1 — a new file type.** This dataset arrives as an **`.RData` file** — an R
object saved to disk. Unlike `read_csv()` or `read_excel()`, you don't assign
the result: `load()` places the saved object(s) directly into your environment.

```{r}
library(tidyverse)

# LM1: load the .RData file
load("Data/[FILENAME].RData")

# What just appeared? List the environment:
ls()
```

**LM2 — inspect.** The loaded object is named `[OBJECT_NAME]`.

```{r}
# LM2: inspect
glimpse([OBJECT_NAME])

# [Instructor: one line noting rows, date range, and unit of observation]
```

# 3 · T — Transformation (28 min)

**LM3 — select.** [Keep the variables needed for the question: e.g.,
`offense`, `date_time`, `neighborhood` … ]

```{r}
# LM3: select the relevant variables
crime_small_df <- [OBJECT_NAME] |>
  select([ ... ])
```

**LM4 — filter.** [Keep the relevant observations: offense type(s), date
window, records with usable values. State the rule you applied.]

```{r}
# LM4: filter the relevant observations
crime_clean_df <- crime_small_df |>
  filter([ ... ])
```

> *How many rows removed, and why is that rule defensible? (Underreporting
> beat returns here.)*

**LM5 — mutate / recode.** [Create or recode one variable the question needs —
e.g., extract hour from the timestamp, or collapse offense categories.]

```{r}
# LM5: create or recode

```

**LM6 — group + summarize + arrange.** [Group by the question's key variable,
count with `n()` or summarize, arrange descending.]

```{r}
# LM6: group_by() |> summarize() |> arrange()
crime_summary_df <- crime_clean_df |>
  group_by([ ... ]) |>
  summarize(n = n()) |>
  arrange(desc(n))

crime_summary_df
```

# 4 · V — Visualization & summary (20 min)

**LM7 — the summary table.** [Present `crime_summary_df` — this table should
already half-answer the question.]

**LM8 — geom_point().** [Plot the summarized counts as points — e.g., count by
hour of day. Add honest labels; per the ethics discussion, the title must not
overclaim.]

```{r}
# LM8: ggplot(...) + geom_point()
ggplot(crime_summary_df, aes(x = [ ... ], y = n)) +
  geom_point(size = 3) +
  labs(
    x = "[ ... ]",
    y = "Number of reported incidents",
    title = "[Pattern, stated carefully — 'reported', not 'occurred']"
  )
```

> *[Describe the important pattern in one sentence — and one sentence on what
> the pattern might reflect besides crime itself.]*

# 5 · Interpretation & final answer (10 min)

> **[Use evidence from the table and the plot to answer the central question
> in 2–3 sentences. Required trio: the pattern, its size, and one caveat drawn
> from section 0 — most likely reporting or enforcement bias.]**
