---
title: "Foundations 03 — Inspecting and Accessing Data"
subtitle: "STAT/DATA 1810 · Week 2 Wednesday · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right-body
    toc-title: "Contents"
execute:
  warning: false
  message: false
---

<!--
CLO: 1.6, 1.12, accessing ($ / [ ] / [[ ]]) · 7.6 · 9 practiced
Timing: job 2' · inspecting 22' · accessing 33' · storage 10' · ETV preview 5'
        · exit 8' = 80'
Checkpoint: §2 starts at 24'. If behind, cut YT5 and shorten YT4 to (a)–(c).
The ETV preview closes foundations: run Cycle-0-style coffee import→plot live,
narrating nothing but the three stage names. Monday they drive C1.
-->

```{r}
#| label: setup
library(tidyverse)
```

::: {.callout-warning title="Timing & checkpoints"}
- **Budget:** job 2′ · inspecting 22′ · accessing 33′ · storage 10′ · ETV preview 5′ · exit 8′ = **80′**
- **Checkpoint:** §2 starts at **24′** — if behind, cut YT5 and shorten YT4 to (a)–(c)
- The ETV preview closes foundations: run a Cycle-0-style coffee import → plot live, narrating nothing but the three stage names — Monday they drive C1
:::

# 1 · Inspecting data [22 min]{.timing}

```{r}
head(airquality); tail(airquality)
dim(airquality); nrow(airquality); ncol(airquality)
```

::: {.callout-tip title="Q1 answer"}
- [Q1]{.q} **153 rows × 6 columns**
- One row = **one day** of measurements (May–Sep 1973)
- The observation question matters more than the numbers — every dataset this term starts with "what is a row?"
:::

```{r}
names(airquality)
str(airquality)
glimpse(airquality)
```

::: {.callout-tip title="Q2 model answer"}
- [Q2]{.q} `glimpse()` is easier — one line per variable, fits any width
- The tags are the **types from Monday**: `int` integers, `dbl` doubles
- Point made: Monday's vocabulary is now reading equipment
:::

```{r}
summary(airquality)
```

::: {.callout-tip title="Q3 answer"}
- [Q3]{.q} **Ozone (37 NA)** and **Solar.R (7 NA)**
- `summary()` prints the NA counts in-column — students should find them without help
:::

```{r}
# YT1 — mtcars, unseen
dim(mtcars)
head(mtcars)
glimpse(mtcars)
summary(mtcars)
```

::: {.callout-note title="Expected — YT1"}
- [YT1]{.yt} **(a)** 32 × 11
- **(b)** one row = one car model — the row *names* are the cars (first time they've seen row names carry meaning)
- **(c)** all numeric — `dbl` everywhere, even `cyl`, `am`, `gear`, which are really categories (connect to Monday's item E)
- **(d)** none — `summary()` shows no NA row
- **(e)** median mpg = **19.2**
- Teaching note: the point of YT1 is that they choose the functions themselves — don't prescribe which
:::

# 2 · Accessing data [33 min]{.timing}

```{r}
mean(airquality$Temp)
```

::: {.callout-note title="Expected"}
- **77.88°F**
:::

```{r}
# YT2
mean(airquality$Wind)
max(airquality$Temp)
mean(airquality$Ozone)
mean(airquality$Ozone, na.rm = TRUE)
sum(airquality$Temp > 90)
```

::: {.callout-note title="Expected — YT2"}
- [YT2]{.yt} **(a)** 9.96 mph
- **(b)** 97
- **(c)** `NA`, then **42.13** with `na.rm = TRUE` — the callback to Foundations 01 Q5; let someone remember it before you say it
- **(d)** **14 days** — the first *logical vector doing arithmetic* (TRUE = 1); say it once, it returns with `filter()` Monday
:::

```{r}
airquality[1, ]
airquality[, 4]
airquality[1:5, c("Ozone", "Temp")]
```

::: {.callout-important title="LM5 — predictions worth pausing on"}
- [LM5]{.lm} `airquality[, 4]` returns a *vector* — a single column drops the frame in base R
- Don't belabor it; Q4 lands it
:::

```{r}
# YT3
airquality[10:15, c("Month", "Day")]
airquality[nrow(airquality), ]
airquality[1:3, 1:3]
head(airquality["Temp"])
```

::: {.callout-tip title="YT3 notes"}
- [YT3]{.yt} **(b)** is the one to check: `nrow()` *inside* the bracket — ask "what if tomorrow the file has 200 rows?" — that's why we don't type 153
- **(d)** previews Q4: no comma → column selection → still a data frame
:::

```{r}
airquality[[1, "Temp"]]
my_list <- list(nums = c(1, 2, 3), word = "hello")
my_list[["nums"]]
my_list$word
```

```{r}
head(airquality$Temp)
head(airquality["Temp"])
head(airquality[["Temp"]])
```

::: {.callout-tip title="Q4 answer — rule of thumb"}
- [Q4]{.q} **the middle one — `airquality["Temp"]`** keeps the data frame wrapping (the printed header gives it away)
- Rule of thumb to leave them with:
  - `$` → a column you'll compute on
  - `[rows, cols]` → a rectangle
  - `[[ ]]` → dig one thing out of a container
- Deeper indexing rules are drill material (M4), not lecture material
:::

```{r}
which(is.na(airquality$Ozone)) |> head()
```

::: {.callout-note title="Expected — first |> sighting"}
- **rows 5, 10, 25, 26, 27, 32**
- Name the pipe ("read it as *then*"), promise it becomes the backbone Monday, move on
:::

```{r}
# YT4 — access challenge on iris
length(iris$Species)                       # (a) 150
iris[51:53, ]                              # (b) versicolor
iris[[100, "Petal.Width"]]                 # (c) 1.3
mean(iris$Sepal.Length[1:50])              # (d) 5.006
my_list2 <- list(flowers = iris, n = nrow(iris))
my_list2[["n"]]                            # (e) 150
my_list2$n                                 # (e) 150 — same
```

::: {.callout-tip title="YT4 notes"}
- [YT4]{.yt} **(a)** 150 · **(b)** versicolor · **(c)** 1.3 · **(d)** 5.006 · **(e)** 150
- **(d)** is the sneaky one — `$` then `[1:50]` on the *vector*
- Some students write `mean(iris[1:50, "Sepal.Length"])` — also correct; show both on the projector, same answer, say the rule of thumb again
- **Stretch:** `$` wins, by a lot — that's the honest answer and it's why the tidyverse is built around column names
:::

# 3 · How big is data? [10 min]{.timing}

```{r}
object.size(airquality)
object.size(iris)
object.size(mtcars)
```

::: {.callout-tip title="Q5 & Q6 answers"}
- [Q5]{.q} actually run it: airquality ≈ 5.6 KB, mtcars ≈ 7.2 KB, iris ≈ 7.8 KB → **airquality < mtcars < iris**
  - mtcars is small in rows but its row *names* cost bytes — a fun 20-second aside if asked; otherwise just confirm the order
- [Q6]{.q} **about 16 copies** — 8 GB ≈ 8,000 MB; 8,000 / 500 = 16
  - honest version: overhead means fewer in practice
  - the point is *order-of-magnitude* reasoning, which is sub-outcome 1.12's whole job
:::

```{r}
# YT5
object.size(iris)
object.size(iris$Sepal.Length)
```

::: {.callout-tip title="YT5 answer"}
- [YT5]{.yt} one column ≈ 1.2 KB of ≈ 7.8 KB — about **15%**, less than 1/5
- Why: the data frame carries overhead (names, row labels, the factor's level table) beyond its raw numbers
- Full credit: any answer noticing "less than 1/5, because the container itself costs something"
:::

# 4 · Where this is going [5 min]{.timing}

::: {.callout-important title="Teaching note — the ETV preview"}
- Run the coffee import → `filter` → histogram in under three minutes
- Say only: *"Extraction. Transformation. Visualization. Monday, this is yours."*
- No function explanations — the mystique is the motivation
:::

# 5 · Exit question [8 min]{.timing}

::: {.callout-tip title="Model answer"}
```{r}
#| eval: false
glimpse(mysterydata)      # structure: rows, columns, names, types
summary(mysterydata)      # ranges + WHERE the NAs are
head(mysterydata)         # what the values actually look like

mysterydata$score         # one column, as a vector
```

- Passes: any three of {`glimpse`/`str`, `summary`, `head`/`tail`, `dim`/`nrow`+`ncol`, `names`} with correct one-line justifications
- The access line must be `$score` or `[["score"]]` (a vector) — `["score"]` is a data frame: half credit with the explanation
- The habit being graded: **inspect before you compute** — the E stage's second half, forever
:::
