---
title: "Foundations 03 — Inspecting and Accessing Data"
subtitle: "STAT/DATA 1810 · Week 2 Wednesday · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right-body
    toc-title: "Contents"
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT
  CLO 1 (1.6 containers in use · 1.12 storage magnitudes, object.size/file.size)
  CLO 1 accessing: $, [ ], [[ ]]  (kept light — the tidyverse verbs take over Monday)
  CLO 7 (7.6 is.na in practice)
  CLO 9 practiced: you insert your own chunks (➕)
Time: 80 min. Datasets: airquality, iris, mtcars (built in).
Timing: job 2' · inspecting 20' · accessing 20' · the pipe 15' · storage 13'
        · ETV preview 5' · exit 5' = 80'
Checkpoint: §3 (the pipe) starts by 42'. If behind, cut YT3(c) and YT5.
Packages: tibble only (glimpse()). The pipe |> is base R — no package needed.
Ends with a 5-minute live preview of a full ETV cycle — where all of this is going.
LM = learning moment. YT = your turn. Times in headers are the section budget.
-->

**Name:** Instructor key

```{r}
#| label: setup
library(tibble)   # for glimpse(). The pipe |> is built into R — nothing to load.
```

::: {.callout-warning title="Timing & checkpoints"}
- **Budget:** job 2′ · inspecting 20′ · accessing 20′ · the pipe 15′ · storage 13′ · ETV preview 5′ · exit 5′ = **80′**
- **Checkpoint:** §3 (the pipe) starts by **42′** — if behind, cut YT3(c) and YT5
- `$` / `[ ]` are deliberately lighter than before: enough to read code and grab a column, not a drill. The pipe and "how big is data" got the time instead
- The ETV preview closes foundations: run a Cycle-0-style coffee import → plot live, narrating nothing but the three stage names — Monday they drive C1
:::

# 0 · Today's job [2 min]{.timing}

You get handed a dataset. Before any analysis, you must answer three
questions: **What is this? How big is it? What's inside?** Today is those
three questions, as functions — then how to *reach in and take a piece* —
and then the one symbol that will carry every line of code you write for
the rest of the quarter.

{{< vspace 1.5rem >}}

# 1 · Inspecting data [20 min]{.timing}

We'll interrogate `airquality` — 1973 New York air measurements, built in.

**LM1 — the first look (5 min).** Run each; note what question each answers:

```{r}
# LM1: five inspection functions
head(airquality)
tail(airquality)
dim(airquality)
nrow(airquality)
ncol(airquality)
```

> *Q1. Rows: 153 · Columns: 6 · What is one OBSERVATION here (what does
> one row represent)?*
>
> *Your answer:* One row is **one day** of air measurements in New York
> (May–September 1973): that day's Ozone, Solar.R, Wind, Temp, Month, Day.

::: {.callout-tip title="Q1 answer"}
- [Q1]{.q} **153 rows × 6 columns**
- One row = **one day** of measurements (May–Sep 1973)
- The observation question matters more than the numbers — every dataset this term starts with "what is a row?"
:::

**LM2 — names and structure (4 min).**

```{r}
# LM2
names(airquality)
str(airquality)
glimpse(airquality)
```

> *Q2. `str()` and `glimpse()` answer the same question. Which is easier to
> read, and what do the `<int>` / `int` tags tell you?*
>
> *Your answer:* `glimpse()` — one line per variable, and it fits any window
> width. The `int` / `<int>` tags are the **data types from Monday**: every
> column here is stored as integers (a `dbl` would mean doubles/decimals) —
> all quantitative, even though Month is really a category.

::: {.callout-tip title="Q2 model answer"}
- [Q2]{.q} `glimpse()` is easier — one line per variable, fits any width
- The tags are the **types from Monday**: `int` integers, `dbl` doubles, `chr` character, `fct` factor
- Point made: Monday's vocabulary is now reading equipment. Ask which *family* each column is — Month is stored quantitative but *means* categorical (Monday's item E)
:::

**LM3 — the numeric overview (3 min).**

```{r}
# LM3
summary(airquality)
```

> *Q3. Which TWO variables contain missing values, and how many each?
> (summary() tells you directly.)*
>
> *Your answer:* **Ozone (37 NA)** and **Solar.R (7 NA)** — `summary()`
> prints an `NA's` line under each column that has them.

::: {.callout-tip title="Q3 answer"}
- [Q3]{.q} **Ozone (37 NA)** and **Solar.R (7 NA)**
- `summary()` prints the NA counts in-column — students should find them without help
:::

*(`View(airquality)` opens a spreadsheet view — try it in the Console; it
doesn't belong in a rendered document.)*

> *YT1 (8 min). ➕ Insert a chunk and inspect a dataset you have NOT seen:
> `mtcars`. Use whichever of the functions above you need to answer:*
> *(a) how many rows and columns?*
> *(b) what is one observation (one row)? Hint: run `head()` and look at the
> row labels on the left.*
> *(c) which columns are NOT numeric — or are they all numeric?*
> *(d) any missing values? How do you know?*
> *(e) what is the median `mpg`?*

<!-- ➕ insert a chunk here -->

```{r}
# YT1 — mtcars, unseen
dim(mtcars)
head(mtcars)
glimpse(mtcars)
summary(mtcars)
```

> *(a) 32 rows × 11 columns · (b) one row = one car model (the row names are
> the cars) · (c) all numeric — every column is `dbl` · (d) none —
> `summary()` shows no `NA's` line for any column · (e) median mpg = 19.2*

::: {.callout-note title="Expected — YT1"}
- [YT1]{.yt} **(a)** 32 × 11
- **(b)** one row = one car model — the row *names* are the cars (first time they've seen row names carry meaning)
- **(c)** all numeric — `dbl` everywhere, even `cyl`, `am`, `gear`, which are really categories (connect to Monday's item E)
- **(d)** none — `summary()` shows no NA row
- **(e)** median mpg = **19.2**
- Teaching note: the point of YT1 is that they choose the functions themselves — don't prescribe which
:::

{{< vspace 1.5rem >}}

# 2 · Accessing data — just enough [20 min]{.timing}

Three tools, three jobs. You need to *read* all three; you'll *write* mostly
the first one:

| Tool | Job | Returns |
|---|---|---|
| `$` | one column, **by name** | a vector |
| `[rows, cols]` | a rectangle, by position/name | usually a data frame |
| `[[ ]]` | one element, dug out | the thing itself |

**LM4 — `$`, the workhorse (5 min).**

```{r}
# LM4: a column is a vector
airquality$Temp
mean(airquality$Temp)
```

::: {.callout-note title="Expected"}
- **77.88°F**
:::

> *YT2 (5 min). ➕ Insert a chunk. Using `$`, find:*
> *(a) the mean of Wind;*
> *(b) the max Temp;*
> *(c) the mean of Ozone — it comes back `NA`. Fix it with the argument you
> looked up in the help page last week;*
> *(d) how many days had Temp above 90? (Hint: `sum(airquality$Temp > 90)` —
> TRUE counts as 1.)*

<!-- ➕ insert a chunk here -->

```{r}
# YT2
mean(airquality$Wind)
max(airquality$Temp)
mean(airquality$Ozone)
mean(airquality$Ozone, na.rm = TRUE)
sum(airquality$Temp > 90)
```

> *(a) 9.96 mph · (b) 97 · (c) `NA`, then 42.13 with `na.rm = TRUE` ·
> (d) 14 days*

::: {.callout-note title="Expected — YT2"}
- [YT2]{.yt} **(a)** 9.96 mph
- **(b)** 97
- **(c)** `NA`, then **42.13** with `na.rm = TRUE` — the callback to Foundations 01 Q5; let someone remember it before you say it
- **(d)** **14 days** — the first *logical vector doing arithmetic* (TRUE = 1); say it once, it returns with `filter()` Monday
:::

**LM5 — `[rows, columns]` and `[[ ]]`, in one pass (5 min).** Predict each
before running. Square brackets take a *position or a name*; double brackets
dig out *one thing*:

```{r}
# LM5: predict, then run
airquality[1, ]                        # first row, all columns
airquality[1:5, c("Ozone", "Temp")]    # a rectangle
airquality[[1, "Temp"]]                # one cell, dug out

# the same three tools on a list (Monday's junk drawer):
my_list <- list(nums = c(1, 2, 3), word = "hello")
my_list$word
my_list[["nums"]]
```

> *Q4. `airquality$Temp` and `airquality[["Temp"]]` give the same vector.
> `airquality["Temp"]` does not — what is it instead? A one-column data
> frame (the printed column header gives it away).*

::: {.callout-tip title="Q4 answer — rule of thumb, and stop there"}
- [Q4]{.q} `airquality["Temp"]` is still a **data frame** (one column); `$` and `[[ ]]` return the bare vector
- Rule of thumb to leave them with, and *no more than this today*:
  - `$` → a column you'll compute on (90% of what you'll write)
  - `[rows, cols]` → a rectangle (you'll read it in other people's code; Monday `filter()`/`select()` replace it)
  - `[[ ]]` → dig one thing out of a container (lists, week 8)
- Deeper indexing rules are drill material, not lecture material
:::

> *YT3 (5 min). ➕ Insert a chunk. Switch to `iris`. Write ONE line for each:*
> *(a) the Species column as a vector — how long is it?*
> *(b) rows 51–53, all columns — what species are they?*
> *(c) the last row of iris (use `nrow()` inside the brackets — no
> hard-coded 150).*

<!-- ➕ insert a chunk here -->

```{r}
# YT3
length(iris$Species)          # (a) 150
iris[51:53, ]                 # (b) versicolor
iris[nrow(iris), ]            # (c) row 150 — virginica
```

::: {.callout-tip title="YT3 notes"}
- [YT3]{.yt} **(a)** 150 · **(b)** versicolor · **(c)** row 150, virginica
- **(c)** is the one to check: `nrow()` *inside* the bracket — ask "what if tomorrow the file has 200 rows?" — that's why we don't type 150
:::

{{< vspace 1.5rem >}}

# 3 · The pipe `|>` — the symbol you'll type more than any other [15 min]{.timing}

::: {.callout-important title="Read this one out loud"}
**`|>` means "and then."** Take the thing on the left, hand it to the function
on the right. That's it. Everything from Monday on — every ETV cycle, every
plot, every summary — is a chain of *and then*s.
:::

**LM6 — the same code, two ways (5 min).** Question: *on average, how warm
were the July days?* Without the pipe, you read this **inside-out**, from the
deepest parentheses to the outermost:

```{r}
# LM6a: nested — read from the INSIDE out
round(mean(airquality$Temp[airquality$Month == 7]), 1)
```

With the pipe, you read it **top-to-bottom, like a recipe**:

```{r}
# LM6b: piped — read "and then", top to bottom
airquality$Temp[airquality$Month == 7] |>
  mean() |>
  round(1)
```

Same answer, 83.9°F. But the second one you can *say*: "take July's
temperatures, **and then** average them, **and then** round to one decimal."

::: {.callout-note title="How the pipe works — the one rule"}
- `x |> f()` is exactly `f(x)`. The left side becomes the **first argument** of the function on the right
- `x |> f(y)` is `f(x, y)` — extra arguments go where they always did
- So `airquality$Temp |> mean(na.rm = TRUE)` is `mean(airquality$Temp, na.rm = TRUE)`
- Keyboard: **Cmd/Ctrl + Shift + M** types `|>` for you in Positron (if it types `%>%`, that's the older tidyverse pipe — same idea; set the *native pipe* option in Positron's settings)
- Style: one step per line, indent the continuation. Your eye follows the recipe
:::

**LM7 — a real chain (4 min).** Which days are missing Ozone? Read each
line as "and then":

```{r}
# LM7: WHICH rows are missing Ozone? — read it as a sentence
airquality$Ozone |>
  is.na() |>
  which() |>
  head()
```

::: {.callout-note title="Expected — LM7"}
- **rows 5, 10, 25, 26, 27, 32**
- Have a student read the chain aloud: "take Ozone, and then flag the NAs, and then find their positions, and then show the first six." If they can say it, they can write it
- The nested version is `head(which(is.na(airquality$Ozone)))` — put it on the projector next to the pipe. Nobody prefers the nested one
:::

> *YT4 (6 min). ➕ Insert a chunk. Rewrite each nested line as a pipe, one
> step per line, and confirm the answer doesn't change:*
> *(a) `sqrt(mean(airquality$Wind))`*
> *(b) `round(mean(airquality$Ozone, na.rm = TRUE), 2)`*
> *(c) `length(unique(iris$Species))`*
> *Then write ONE pipe of your own: three steps, any data, and a comment
> reading it as a sentence.*

<!-- ➕ insert a chunk here -->

```{r}
# YT4
airquality$Wind |> mean() |> sqrt()                    # (a) 3.156
airquality$Ozone |> mean(na.rm = TRUE) |> round(2)     # (b) 42.13
iris$Species |> unique() |> length()                   # (c) 3

# my own: "take Temp, and then keep the values above 90, and then count them"
airquality$Temp[airquality$Temp > 90] |> length()
```

::: {.callout-tip title="YT4 notes"}
- [YT4]{.yt} **(a)** 3.156 · **(b)** 42.13 · **(c)** 3
- **(b)** is the check: `na.rm = TRUE` stays *inside* `mean()` — the pipe only fills the first argument
- The "own pipe" is the deliverable: walk the room, ask three students to read theirs aloud. Monday's `coffee |> filter(...) |> ggplot(...)` is the same grammar with better verbs
:::

{{< vspace 1.5rem >}}

# 4 · How big is data? [13 min]{.timing}

**LM8 — the ladder (4 min).** Data size is measured in **bytes** — one byte
is roughly one letter. Every step up the ladder multiplies by a thousand:

![Each step is ×1,000. A byte is a letter; a kilobyte is a paragraph; a megabyte is a photo; a gigabyte is an hour of video; a terabyte is a whole laptop.](/assets/byte_ladder.svg){fig-alt="Ladder of storage units from byte to terabyte, each step ×1,000, with an everyday example on each step." width="100%"}

**LM9 — where real things land (4 min).** The bars below are on a *log*
scale — each gridline is another ×1,000. Notice where **your laptop's
memory** sits, because that's the ceiling for what R can hold:

![Real-world sizes on a log scale, from one letter to a national census. The orange "laptop memory" bar is the amount of data R can hold at once.](/assets/how_big_is_it.svg){fig-alt="Horizontal bar chart on a log scale showing sizes of everyday things: a text message (bytes), a survey CSV (KB), a song (MB), an hour of HD video (GB), a laptop's RAM (8–16 GB), a laptop drive (512 GB), and a census extract (1 TB)." width="100%"}

| Unit | ≈ bytes | Something you own that's about that size |
|---|---|---|
| **byte** | 1 | one letter of a text message |
| **kilobyte (KB)** | 1 thousand | a long email; this class's survey as a `.csv`; `airquality` (5.6 KB) |
| **megabyte (MB)** | 1 million | one phone photo (3 MB); one song (4 MB); a year of CA crime data (~100 MB) |
| **gigabyte (GB)** | 1 billion | an hour of HD Netflix (3 GB); your laptop's **memory** (8–16 GB) |
| **terabyte (TB)** | 1 trillion | two of your laptop's drives; a full US census extract |
| **petabyte (PB)** | 1 quadrillion | Netflix's entire catalog; the data behind a weather forecast |

Now measure **our** objects:

```{r}
# LM9: how big are OUR objects?
object.size(airquality)
object.size(iris)
object.size(mtcars)
```

> *Q5. Order the three from smallest to largest, and name the unit they're
> in: airquality (≈ 5.6 KB) < mtcars (≈ 7.2 KB) < iris (≈ 7.8 KB) — all
> kilobytes, i.e. a few paragraphs each.*

**Why it matters (3 min):** R holds data **in memory** — the orange bar, not
the red one. A 2 GB file on an 8 GB laptop imports (barely); a 200 GB file
does not — you'd need different tools (a database, or a bigger machine). File
size is the first thing to check *before* importing, which is exactly where
next week begins.

> *Q6. Your laptop has 8 GB of memory. Roughly how many copies of a 500 MB
> dataset could fit at once (ignore overhead)? About 16 — 8 GB ≈ 8,000 MB,
> and 8,000 / 500 = 16. How many hours of HD video is that? Not even three.*

::: {.callout-tip title="Q5 & Q6 answers"}
- [Q5]{.q} actually run it: airquality ≈ 5.6 KB, mtcars ≈ 7.2 KB, iris ≈ 7.8 KB → **airquality < mtcars < iris**, all in KB
  - mtcars is small in rows but its row *names* cost bytes — a fun 20-second aside if asked; otherwise just confirm the order
- [Q6]{.q} **about 16 copies** — 8 GB ≈ 8,000 MB; 8,000 / 500 = 16; that whole 8 GB is under three hours of HD video
  - honest version: overhead means fewer in practice; R also copies objects when you modify them
  - the point is *order-of-magnitude* reasoning, which is sub-outcome 1.12's whole job
- Land the contrast: their **phone camera roll** (tens of GB) is bigger than any dataset they'll open in this course — data science files are not the big files in their life
:::

> *YT5 (2 min). ➕ Insert a chunk. Run `object.size(iris)` and then
> `object.size(iris$Sepal.Length)`. Roughly what fraction of the whole is
> one column? Does that match "1 of 5 columns"? One sentence on why or why
> not.*

<!-- ➕ insert a chunk here -->

```{r}
# YT5
object.size(iris)
object.size(iris$Sepal.Length)
```

> *One column is ≈ 1.2 KB of ≈ 7.8 KB — about 15%, which is less than 1/5.
> It doesn't match because the data frame itself costs something: column
> names, row labels, and the Species factor's level table are stored on top
> of the raw numbers.*

::: {.callout-tip title="YT5 answer"}
- [YT5]{.yt} one column ≈ 1.2 KB of ≈ 7.8 KB — about **15%**, less than 1/5
- Why: the data frame carries overhead (names, row labels, the factor's level table) beyond its raw numbers
- Full credit: any answer noticing "less than 1/5, because the container itself costs something"
:::

{{< vspace 1.5rem >}}

# 5 · Where this is going [5 min — watch]{.timing}

Instructor runs one **complete ETV cycle live**, start to finish, no pauses:
import a CSV → transform → plot. Every function you just learned appears in
the first thirty seconds — and every line is joined by `|>`. Monday, you
drive.

::: {.callout-important title="Teaching note — the ETV preview"}
- Run the coffee import → `filter` → histogram in under three minutes
- Say only: *"Extraction. Transformation. Visualization. Monday, this is yours."* — and point at each `|>` as you go: "and then"
- No function explanations — the mystique is the motivation
:::

# 6 · Exit question [5 min]{.timing}

> *You're handed `mysterydata`, a data frame you've never seen. Write the
> three lines you'd run FIRST, in order, and say what each tells you. Then
> write ONE pipe that takes the column `score`, removes NAs from its mean,
> and rounds to one decimal.*

```{r}
#| eval: false
# your three inspection lines:
glimpse(mysterydata)      # structure: rows, columns, names, types
summary(mysterydata)      # ranges + WHERE the NAs are
head(mysterydata)         # what the values actually look like

# your one pipe:
mysterydata$score |>
  mean(na.rm = TRUE) |>
  round(1)
```

> *Your explanation:* `glimpse()` first tells me the shape (rows, columns),
> the names, and the type of every column. `summary()` then gives the range
> of each numeric column and, crucially, where the `NA`s are. `head()` shows
> what the values actually look like so I can sanity-check the types. Then
> the pipe reads "take score, and then average it ignoring NAs, and then
> round to one decimal."

::: {.callout-tip title="Model answer"}
```{r}
#| eval: false
glimpse(mysterydata)      # structure: rows, columns, names, types
summary(mysterydata)      # ranges + WHERE the NAs are
head(mysterydata)         # what the values actually look like

mysterydata$score |> mean(na.rm = TRUE) |> round(1)
```

- Passes: any three of {`glimpse`/`str`, `summary`, `head`/`tail`, `dim`/`nrow`+`ncol`, `names`} with correct one-line justifications
- The pipe must start from `mysterydata$score` (or `[["score"]]`) and keep `na.rm = TRUE` *inside* `mean()`; `mysterydata["score"]` is a data frame — half credit with the explanation
- The habits being graded: **inspect before you compute**, and **write it as a recipe**
:::

Save as `fnd03_yourlastname.qmd` in your `R/` folder and **Render**.
