---
title: "Foundations 03 — Inspecting and Accessing Data"
subtitle: "STAT/DATA 1810 · Week 2 Wednesday · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right-body
    toc-title: "Contents"
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT
  CLO 1 (1.6 containers in use · 1.12 storage magnitudes, object.size/file.size)
  CLO 1 accessing: $, [ ], [[ ]]
  CLO 7 (7.6 is.na in practice)
  CLO 9 practiced: you insert your own chunks (➕)
Time: 80 min. Datasets: airquality, iris, mtcars (built in).
Timing: job 2' · inspecting 22' · accessing 33' · storage 10' · ETV preview 5'
        · exit 8' = 80'
Checkpoint: §2 starts at 24'. If behind, cut YT5 and shorten YT4 to (a)–(c).
Ends with a 5-minute live preview of a full ETV cycle — where all of this is going.
The ETV preview closes foundations: run Cycle-0-style coffee import→plot live,
narrating nothing but the three stage names. Monday they drive C1.
LM = learning moment. YT = your turn. Times in headers are the section budget.
-->

**Name:** Instructor key

```{r}
#| label: setup
library(tidyverse)
```

::: {.callout-warning title="Timing & checkpoints"}
- **Budget:** job 2′ · inspecting 22′ · accessing 33′ · storage 10′ · ETV preview 5′ · exit 8′ = **80′**
- **Checkpoint:** §2 starts at **24′** — if behind, cut YT5 and shorten YT4 to (a)–(c)
- The ETV preview closes foundations: run a Cycle-0-style coffee import → plot live, narrating nothing but the three stage names — Monday they drive C1
:::

# 0 · Today's job [2 min]{.timing}

You get handed a dataset. Before any analysis, you must answer three
questions: **What is this? How big is it? What's inside?** Today is those
three questions, as functions — and then, once you know what's inside,
how to *reach in and take a piece*.

# 1 · Inspecting data [22 min]{.timing}

We'll interrogate `airquality` — 1973 New York air measurements, built in.

**LM1 — the first look (5 min).** Run each; note what question each answers:

```{r}
# LM1: five inspection functions
head(airquality)
tail(airquality)
dim(airquality)
nrow(airquality)
ncol(airquality)
```

> *Q1. Rows: 153 · Columns: 6 · What is one OBSERVATION here (what does
> one row represent)?*
>
> *Your answer:* One row is **one day** of air measurements in New York
> (May–September 1973): that day's Ozone, Solar.R, Wind, Temp, Month, Day.

::: {.callout-tip title="Q1 answer"}
- [Q1]{.q} **153 rows × 6 columns**
- One row = **one day** of measurements (May–Sep 1973)
- The observation question matters more than the numbers — every dataset this term starts with "what is a row?"
:::

**LM2 — names and structure (5 min).**

```{r}
# LM2
names(airquality)
str(airquality)
glimpse(airquality)
```

> *Q2. `str()` and `glimpse()` answer the same question. Which is easier to
> read, and what do the `<int>` / `int` tags tell you?*
>
> *Your answer:* `glimpse()` — one line per variable, and it fits any window
> width. The `int` / `<int>` tags are the **data types from Monday**: every
> column here is stored as integers (a `dbl` would mean doubles/decimals).

::: {.callout-tip title="Q2 model answer"}
- [Q2]{.q} `glimpse()` is easier — one line per variable, fits any width
- The tags are the **types from Monday**: `int` integers, `dbl` doubles
- Point made: Monday's vocabulary is now reading equipment
:::

**LM3 — the numeric overview (4 min).**

```{r}
# LM3
summary(airquality)
```

> *Q3. Which TWO variables contain missing values, and how many each?
> (summary() tells you directly.)*
>
> *Your answer:* **Ozone (37 NA)** and **Solar.R (7 NA)** — `summary()`
> prints an `NA's` line under each column that has them.

::: {.callout-tip title="Q3 answer"}
- [Q3]{.q} **Ozone (37 NA)** and **Solar.R (7 NA)**
- `summary()` prints the NA counts in-column — students should find them without help
:::

*(`View(airquality)` opens a spreadsheet view — try it in the Console; it
doesn't belong in a rendered document.)*

> *YT1 (8 min). ➕ Insert a chunk and inspect a dataset you have NOT seen:
> `mtcars`. Use whichever of the functions above you need to answer:*
> *(a) how many rows and columns?*
> *(b) what is one observation (one row)? Hint: run `head()` and look at the
> row labels on the left.*
> *(c) which columns are NOT numeric — or are they all numeric?*
> *(d) any missing values? How do you know?*
> *(e) what is the median `mpg`?*

<!-- ➕ insert a chunk here -->

```{r}
# YT1 — mtcars, unseen
dim(mtcars)
head(mtcars)
glimpse(mtcars)
summary(mtcars)
```

> *(a) 32 rows × 11 columns · (b) one row = one car model (the row names are
> the cars) · (c) all numeric — every column is `dbl` · (d) none —
> `summary()` shows no `NA's` line for any column · (e) median mpg = 19.2*

::: {.callout-note title="Expected — YT1"}
- [YT1]{.yt} **(a)** 32 × 11
- **(b)** one row = one car model — the row *names* are the cars (first time they've seen row names carry meaning)
- **(c)** all numeric — `dbl` everywhere, even `cyl`, `am`, `gear`, which are really categories (connect to Monday's item E)
- **(d)** none — `summary()` shows no NA row
- **(e)** median mpg = **19.2**
- Teaching note: the point of YT1 is that they choose the functions themselves — don't prescribe which
:::

# 2 · Accessing data [33 min]{.timing}

Three tools, three jobs:

| Tool | Job | Returns |
|---|---|---|
| `$` | one column, **by name** | a vector |
| `[rows, cols]` | a rectangle, by position/name | usually a data frame |
| `[[ ]]` | one element, dug out | the thing itself |

**LM4 — `$`, the workhorse (5 min).**

```{r}
# LM4: a column is a vector
airquality$Temp
mean(airquality$Temp)
```

::: {.callout-note title="Expected"}
- **77.88°F**
:::

> *YT2 (4 min). ➕ Insert a chunk. Using `$`, find:*
> *(a) the mean of Wind;*
> *(b) the max Temp;*
> *(c) the mean of Ozone — it comes back `NA`. Fix it with the argument you
> looked up in the help page last week;*
> *(d) how many days had Temp above 90? (Hint: `sum(airquality$Temp > 90)` —
> TRUE counts as 1.)*

<!-- ➕ insert a chunk here -->

```{r}
# YT2
mean(airquality$Wind)
max(airquality$Temp)
mean(airquality$Ozone)
mean(airquality$Ozone, na.rm = TRUE)
sum(airquality$Temp > 90)
```

> *(a) 9.96 mph · (b) 97 · (c) `NA`, then 42.13 with `na.rm = TRUE` ·
> (d) 14 days*

::: {.callout-note title="Expected — YT2"}
- [YT2]{.yt} **(a)** 9.96 mph
- **(b)** 97
- **(c)** `NA`, then **42.13** with `na.rm = TRUE` — the callback to Foundations 01 Q5; let someone remember it before you say it
- **(d)** **14 days** — the first *logical vector doing arithmetic* (TRUE = 1); say it once, it returns with `filter()` Monday
:::

**LM5 — `[rows, columns]` (6 min).** Predict each before running:

```{r}
# LM5: predict, then run
airquality[1, ]        # first row, all columns
airquality[, 4]        # all rows, 4th column
airquality[1:5, c("Ozone", "Temp")]
```

::: {.callout-important title="LM5 — predictions worth pausing on"}
- [LM5]{.lm} `airquality[, 4]` returns a *vector* — a single column drops the frame in base R
- Don't belabor it; Q4 lands it
:::

> *YT3 (5 min). ➕ Insert a chunk. Write ONE line for each:*
> *(a) rows 10 through 15 of ONLY the Month and Day columns;*
> *(b) the last row of airquality (use `nrow()` inside the brackets — no
> hard-coded 153);*
> *(c) rows 1–3, columns 1–3, by position;*
> *(d) all rows of the Temp column, but as a data frame not a vector (hint:
> leave out the comma).*

<!-- ➕ insert a chunk here -->

```{r}
# YT3
airquality[10:15, c("Month", "Day")]       # (a)
airquality[nrow(airquality), ]             # (b)
airquality[1:3, 1:3]                       # (c)
head(airquality["Temp"])                   # (d) — head() only to keep the output short
```

::: {.callout-tip title="YT3 notes"}
- [YT3]{.yt} **(b)** is the one to check: `nrow()` *inside* the bracket — ask "what if tomorrow the file has 200 rows?" — that's why we don't type 153
- **(d)** previews Q4: no comma → column selection → still a data frame
:::

**LM6 — `[[ ]]`, digging one thing out (4 min).**

```{r}
# LM6: single element — row 1's Temp
airquality[[1, "Temp"]]

# and on a list:
my_list <- list(nums = c(1, 2, 3), word = "hello")
my_list[["nums"]]
my_list$word
```

**LM7 — the comparison (4 min).** Run all three; look closely at what each
returns:

```{r}
# LM7: three ways at the same column
head(airquality$Temp)          # vector
head(airquality["Temp"])       # still a data frame (note the header!)
head(airquality[["Temp"]])     # vector again
```

> *Q4. Which of the three kept the data frame wrapping? The middle one —
> `airquality["Temp"]` (the printed column header gives it away).*

::: {.callout-tip title="Q4 answer — rule of thumb"}
- [Q4]{.q} **the middle one — `airquality["Temp"]`** keeps the data frame wrapping (the printed header gives it away)
- Rule of thumb to leave them with:
  - `$` → a column you'll compute on
  - `[rows, cols]` → a rectangle
  - `[[ ]]` → dig one thing out of a container
- Deeper indexing rules are drill material (M4), not lecture material
:::

**LM8 — missing values, located (2 min).**

```{r}
# LM8: WHICH rows are missing Ozone?
which(is.na(airquality$Ozone)) |> head()
```

::: {.callout-note title="Expected — first |> sighting"}
- **rows 5, 10, 25, 26, 27, 32**
- Name the pipe ("read it as *then*"), promise it becomes the backbone Monday, move on
:::

> *YT4 — access challenge (7 min). ➕ Insert a chunk. Switch to `iris`.*
> *(a) the Species column, as a vector — how long is it?*
> *(b) rows 51–53, all columns — what species are they?*
> *(c) the single value in row 100, column Petal.Width;*
> *(d) the mean Sepal.Length of ONLY the first 50 rows (setosa);*
> *(e) build `my_list2 <- list(flowers = iris, n = nrow(iris))` and pull `n`
> back out with `[[ ]]`, then with `$`. Same answer?*
> *Stretch: which of the three tools did you use most? Why?*

<!-- ➕ insert a chunk here -->

```{r}
# YT4 — access challenge on iris
length(iris$Species)                       # (a) 150
iris[51:53, ]                              # (b) versicolor
iris[[100, "Petal.Width"]]                 # (c) 1.3
mean(iris$Sepal.Length[1:50])              # (d) 5.006
my_list2 <- list(flowers = iris, n = nrow(iris))
my_list2[["n"]]                            # (e) 150
my_list2$n                                 # (e) 150 — same
```

> *(a) 150 · (b) versicolor · (c) 1.3 · (d) 5.006 · (e) 150 both ways — same
> answer. Stretch: `$`, by a lot — most questions are "give me this column by
> name", which is exactly what `$` does.*

::: {.callout-tip title="YT4 notes"}
- [YT4]{.yt} **(a)** 150 · **(b)** versicolor · **(c)** 1.3 · **(d)** 5.006 · **(e)** 150
- **(d)** is the sneaky one — `$` then `[1:50]` on the *vector*
- Some students write `mean(iris[1:50, "Sepal.Length"])` — also correct; show both on the projector, same answer, say the rule of thumb again
- **Stretch:** `$` wins, by a lot — that's the honest answer and it's why the tidyverse is built around column names
:::

# 3 · How big is data? [10 min]{.timing}

**LM9 — storage units (4 min).** Data size is measured in bytes:
**byte → KB (thousand) → MB (million) → GB (billion) → TB (trillion)**.
Rough anchors: this class's survey ≈ a few KB · a semester of sensor logs ≈
MBs · a music library ≈ GBs · a national census extract ≈ TBs.

```{r}
# LM9: how big are OUR objects?
object.size(airquality)
object.size(iris)
object.size(mtcars)
```

> *Q5. Order the three from smallest to largest: airquality (≈ 5.6 KB) <
> mtcars (≈ 7.2 KB) < iris (≈ 7.8 KB)*

**Why it matters:** R holds data **in memory**. A 2 GB file on a 8 GB laptop
imports; a 200 GB file does not — you'd need different tools. File size is
the first thing to check *before* importing, which is exactly where next
week begins.

> *Q6. Your laptop has 8 GB of memory. Roughly how many copies of a 500 MB
> dataset could fit at once (ignore overhead)? About 16 — 8 GB ≈ 8,000 MB,
> and 8,000 / 500 = 16.*

::: {.callout-tip title="Q5 & Q6 answers"}
- [Q5]{.q} actually run it: airquality ≈ 5.6 KB, mtcars ≈ 7.2 KB, iris ≈ 7.8 KB → **airquality < mtcars < iris**
  - mtcars is small in rows but its row *names* cost bytes — a fun 20-second aside if asked; otherwise just confirm the order
- [Q6]{.q} **about 16 copies** — 8 GB ≈ 8,000 MB; 8,000 / 500 = 16
  - honest version: overhead means fewer in practice
  - the point is *order-of-magnitude* reasoning, which is sub-outcome 1.12's whole job
:::

> *YT5 (3 min). ➕ Insert a chunk. Run `object.size(iris)` and then
> `object.size(iris$Sepal.Length)`. Roughly what fraction of the whole is
> one column? Does that match "1 of 5 columns"? One sentence on why or why
> not.*

<!-- ➕ insert a chunk here -->

```{r}
# YT5
object.size(iris)
object.size(iris$Sepal.Length)
```

> *One column is ≈ 1.2 KB of ≈ 7.8 KB — about 15%, which is less than 1/5.
> It doesn't match because the data frame itself costs something: column
> names, row labels, and the Species factor's level table are stored on top
> of the raw numbers.*

::: {.callout-tip title="YT5 answer"}
- [YT5]{.yt} one column ≈ 1.2 KB of ≈ 7.8 KB — about **15%**, less than 1/5
- Why: the data frame carries overhead (names, row labels, the factor's level table) beyond its raw numbers
- Full credit: any answer noticing "less than 1/5, because the container itself costs something"
:::

# 4 · Where this is going [5 min — watch]{.timing}

Instructor runs one **complete ETV cycle live**, start to finish, no pauses:
import a CSV → transform → plot. Every function you just learned appears in
the first thirty seconds. Monday, you drive.

::: {.callout-important title="Teaching note — the ETV preview"}
- Run the coffee import → `filter` → histogram in under three minutes
- Say only: *"Extraction. Transformation. Visualization. Monday, this is yours."*
- No function explanations — the mystique is the motivation
:::

# 5 · Exit question [8 min]{.timing}

> *You're handed `mysterydata`, a data frame you've never seen. Write the
> three lines you'd run FIRST, in order, and say what each tells you. Then
> write ONE more line that pulls out a single column named `score` as a
> vector.*

```{r}
#| eval: false
# your three inspection lines:
glimpse(mysterydata)      # structure: rows, columns, names, types
summary(mysterydata)      # ranges + WHERE the NAs are
head(mysterydata)         # what the values actually look like

# your one access line:
mysterydata$score         # one column, as a vector
```

> *Your explanation:* `glimpse()` first tells me the shape (rows, columns),
> the names, and the type of every column. `summary()` then gives the range
> of each numeric column and, crucially, where the `NA`s are. `head()` shows
> what the values actually look like so I can sanity-check the types. Then
> `mysterydata$score` pulls the one column out as a vector (so would
> `mysterydata[["score"]]`; `mysterydata["score"]` would still be a data
> frame).

::: {.callout-tip title="Model answer"}
```{r}
#| eval: false
glimpse(mysterydata)      # structure: rows, columns, names, types
summary(mysterydata)      # ranges + WHERE the NAs are
head(mysterydata)         # what the values actually look like

mysterydata$score         # one column, as a vector
```

- Passes: any three of {`glimpse`/`str`, `summary`, `head`/`tail`, `dim`/`nrow`+`ncol`, `names`} with correct one-line justifications
- The access line must be `$score` or `[["score"]]` (a vector) — `["score"]` is a data frame: half credit with the explanation
- The habit being graded: **inspect before you compute** — the E stage's second half, forever
:::

Save as `fnd03_yourlastname.qmd` in your `R/` folder and **Render**.
