---
title: "Foundations 03 — Inspecting and Accessing Data"
subtitle: "STAT/DATA 1810 · Week 2 Wednesday · Student version"
author: "Your name here"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT
  CLO 1 (1.6 containers in use · 1.12 storage magnitudes, object.size/file.size)
  CLO 1 accessing: $, [ ], [[ ]]  (kept light — the tidyverse verbs take over Monday)
  CLO 7 (7.6 is.na in practice)
  CLO 9 practiced: you insert your own chunks (➕)
Time: 80 min. Datasets: airquality, iris, mtcars (built in).
Packages: tibble only (glimpse()). The pipe |> is base R — no package needed.
Ends with a 5-minute live preview of a full ETV cycle — where all of this is going.
LM = learning moment. YT = your turn. Times in headers are the section budget.
-->

```{r}
#| label: setup
library(tibble)   # for glimpse(). The pipe |> is built into R — nothing to load.
```

# 0 · Today's job (2 min)

You get handed a dataset. Before any analysis, you must answer three
questions: **What is this? How big is it? What's inside?** Today is those
three questions, as functions — then how to *reach in and take a piece* —
and then the one symbol that will carry every line of code you write for
the rest of the quarter.

{{< vspace 1.5rem >}}

# 1 · Inspecting data (20 min)

We'll interrogate `airquality` — 1973 New York air measurements, built in.

**LM1 — the first look (5 min).** Run each; note what question each answers:

```{r}
# LM1: five inspection functions
head(airquality)
tail(airquality)
dim(airquality)
nrow(airquality)
ncol(airquality)
```

> *Q1. Rows: ____ · Columns: ____ · What is one OBSERVATION here (what does
> one row represent)?*
>
> *Your answer:*

**LM2 — names and structure (4 min).**

```{r}
# LM2
names(airquality)
str(airquality)
glimpse(airquality)
```

> *Q2. `str()` and `glimpse()` answer the same question. Which is easier to
> read, and what do the `<int>` / `int` tags tell you?*
>
> *Your answer:*

**LM3 — the numeric overview (3 min).**

```{r}
# LM3
summary(airquality)
```

> *Q3. Which TWO variables contain missing values, and how many each?
> (summary() tells you directly.)*
>
> *Your answer:*

*(`View(airquality)` opens a spreadsheet view — try it in the Console; it
doesn't belong in a rendered document.)*

> *YT1 (8 min). ➕ Insert a chunk and inspect a dataset you have NOT seen:
> `mtcars`. Use whichever of the functions above you need to answer:*
> *(a) how many rows and columns?*
> *(b) what is one observation (one row)? Hint: run `head()` and look at the
> row labels on the left.*
> *(c) which columns are NOT numeric — or are they all numeric?*
> *(d) any missing values? How do you know?*
> *(e) what is the median `mpg`?*

<!-- ➕ insert a chunk here -->

> *(a) ____ · (b) ____ · (c) ____ · (d) ____ · (e) ____*

{{< vspace 1.5rem >}}

# 2 · Accessing data — just enough (20 min)

Three tools, three jobs. You need to *read* all three; you'll *write* mostly
the first one:

| Tool | Job | Returns |
|---|---|---|
| `$` | one column, **by name** | a vector |
| `[rows, cols]` | a rectangle, by position/name | usually a data frame |
| `[[ ]]` | one element, dug out | the thing itself |

**LM4 — `$`, the workhorse (5 min).**

```{r}
# LM4: a column is a vector
airquality$Temp
mean(airquality$Temp)
```

> *YT2 (5 min). ➕ Insert a chunk. Using `$`, find:*
> *(a) the mean of Wind;*
> *(b) the max Temp;*
> *(c) the mean of Ozone — it comes back `NA`. Fix it with the argument you
> looked up in the help page last week;*
> *(d) how many days had Temp above 90? (Hint: `sum(airquality$Temp > 90)` —
> TRUE counts as 1.)*

<!-- ➕ insert a chunk here -->

> *(a) ____ · (b) ____ · (c) ____ · (d) ____*

**LM5 — `[rows, columns]` and `[[ ]]`, in one pass (5 min).** Predict each
before running. Square brackets take a *position or a name*; double brackets
dig out *one thing*:

```{r}
# LM5: predict, then run
airquality[1, ]                        # first row, all columns
airquality[1:5, c("Ozone", "Temp")]    # a rectangle
airquality[[1, "Temp"]]                # one cell, dug out

# the same three tools on a list (Monday's junk drawer):
my_list <- list(nums = c(1, 2, 3), word = "hello")
my_list$word
my_list[["nums"]]
```

> *Q4. `airquality$Temp` and `airquality[["Temp"]]` give the same vector.
> `airquality["Temp"]` does not — what is it instead? ____*

> *YT3 (5 min). ➕ Insert a chunk. Switch to `iris`. Write ONE line for each:*
> *(a) the Species column as a vector — how long is it?*
> *(b) rows 51–53, all columns — what species are they?*
> *(c) the last row of iris (use `nrow()` inside the brackets — no
> hard-coded 150).*

<!-- ➕ insert a chunk here -->

{{< vspace 1.5rem >}}

# 3 · The pipe `|>` — the symbol you'll type more than any other (15 min)

::: {.callout-important title="Read this one out loud"}
**`|>` means "and then."** Take the thing on the left, hand it to the function
on the right. That's it. Everything from Monday on — every ETV cycle, every
plot, every summary — is a chain of *and then*s.
:::

**LM6 — the same code, two ways (5 min).** Question: *on average, how warm
were the July days?* Without the pipe, you read this **inside-out**, from the
deepest parentheses to the outermost:

```{r}
# LM6a: nested — read from the INSIDE out
round(mean(airquality$Temp[airquality$Month == 7]), 1)
```

With the pipe, you read it **top-to-bottom, like a recipe**:

```{r}
# LM6b: piped — read "and then", top to bottom
airquality$Temp[airquality$Month == 7] |>
  mean() |>
  round(1)
```

Same answer. But the second one you can *say*: "take July's temperatures,
**and then** average them, **and then** round to one decimal."

::: {.callout-note title="How the pipe works — the one rule"}
- `x |> f()` is exactly `f(x)`. The left side becomes the **first argument** of the function on the right
- `x |> f(y)` is `f(x, y)` — extra arguments go where they always did
- So `airquality$Temp |> mean(na.rm = TRUE)` is `mean(airquality$Temp, na.rm = TRUE)`
- Keyboard: **Cmd/Ctrl + Shift + M** types `|>` for you in Positron
- Style: one step per line, indent the continuation. Your eye follows the recipe
:::

**LM7 — a real chain (4 min).** Which days are missing Ozone? Read each
line as "and then":

```{r}
# LM7: WHICH rows are missing Ozone? — read it as a sentence
airquality$Ozone |>
  is.na() |>
  which() |>
  head()
```

> *YT4 (6 min). ➕ Insert a chunk. Rewrite each nested line as a pipe, one
> step per line, and confirm the answer doesn't change:*
> *(a) `sqrt(mean(airquality$Wind))`*
> *(b) `round(mean(airquality$Ozone, na.rm = TRUE), 2)`*
> *(c) `length(unique(iris$Species))`*
> *Then write ONE pipe of your own: three steps, any data, and a comment
> reading it as a sentence.*

<!-- ➕ insert a chunk here -->

{{< vspace 1.5rem >}}

# 4 · How big is data? (13 min)

**LM8 — the ladder (4 min).** Data size is measured in **bytes** — one byte
is roughly one letter. Every step up the ladder multiplies by a thousand:

![Each step is ×1,000. A byte is a letter; a kilobyte is a paragraph; a megabyte is a photo; a gigabyte is an hour of video; a terabyte is a whole laptop.](/assets/byte_ladder.svg){fig-alt="Ladder of storage units from byte to terabyte, each step ×1,000, with an everyday example on each step." width="100%"}

**LM9 — where real things land (4 min).** The bars below are on a *log*
scale — each gridline is another ×1,000. Notice where **your laptop's
memory** sits, because that's the ceiling for what R can hold:

![Real-world sizes on a log scale, from one letter to a national census. The orange "laptop memory" bar is the amount of data R can hold at once.](/assets/how_big_is_it.svg){fig-alt="Horizontal bar chart on a log scale showing sizes of everyday things: a text message (bytes), a survey CSV (KB), a song (MB), an hour of HD video (GB), a laptop's RAM (8–16 GB), a laptop drive (512 GB), and a census extract (1 TB)." width="100%"}

| Unit | ≈ bytes | Something you own that's about that size |
|---|---|---|
| **byte** | 1 | one letter of a text message |
| **kilobyte (KB)** | 1 thousand | a long email; this class's survey as a `.csv`; `airquality` (5.6 KB) |
| **megabyte (MB)** | 1 million | one phone photo (3 MB); one song (4 MB); a year of CA crime data (~100 MB) |
| **gigabyte (GB)** | 1 billion | an hour of HD Netflix (3 GB); your laptop's **memory** (8–16 GB) |
| **terabyte (TB)** | 1 trillion | two of your laptop's drives; a full US census extract |
| **petabyte (PB)** | 1 quadrillion | Netflix's entire catalog; the data behind a weather forecast |

Now measure **our** objects:

```{r}
# LM9: how big are OUR objects?
object.size(airquality)
object.size(iris)
object.size(mtcars)
```

> *Q5. Order the three from smallest to largest, and name the unit they're
> in: ____*

**Why it matters (3 min):** R holds data **in memory** — the orange bar, not
the red one. A 2 GB file on an 8 GB laptop imports (barely); a 200 GB file
does not — you'd need different tools (a database, or a bigger machine). File
size is the first thing to check *before* importing, which is exactly where
next week begins.

> *Q6. Your laptop has 8 GB of memory. Roughly how many copies of a 500 MB
> dataset could fit at once (ignore overhead)? ____ How many hours of HD
> video is that? ____*

> *YT5 (2 min). ➕ Insert a chunk. Run `object.size(iris)` and then
> `object.size(iris$Sepal.Length)`. Roughly what fraction of the whole is
> one column? Does that match "1 of 5 columns"? One sentence on why or why
> not.*

<!-- ➕ insert a chunk here -->

{{< vspace 1.5rem >}}

# 5 · Where this is going (5 min — watch)

Instructor runs one **complete ETV cycle live**, start to finish, no pauses:
import a CSV → transform → plot. Every function you just learned appears in
the first thirty seconds — and every line is joined by `|>`. Monday, you
drive.

# 6 · Exit question (5 min)

> *You're handed `mysterydata`, a data frame you've never seen. Write the
> three lines you'd run FIRST, in order, and say what each tells you. Then
> write ONE pipe that takes the column `score`, removes NAs from its mean,
> and rounds to one decimal.*

```{r}
#| eval: false
# your three inspection lines:


# your one pipe:

```

> *Your explanation:*

Save as `fnd03_yourlastname.qmd` in your `R/` folder and **Render**.
