---
title: "ETV Cycle 02 — Kobe Bryant's First Five Seasons"
subtitle: "STAT/DATA 1810 · In-class activity · Student version"
author: "Your name here"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
  eval: false
---

<!--
CLO ALIGNMENT
  CLO 2 (2.2 read_excel: workbooks, sheets)
  CLO 3 (3.1 dplyr: filter, select, mutate, group_by, summarize, arrange)
  CLO 4 (4.1 ggplot + geom_point; 4.5 defensible claim)
  CLO 12 (12.2 Measurement — totals vs rates; 12.6 Transparency — document the source)
Time budget: 80 minutes. LM = short learning moment.
-->

# 0 · Context & ethics (8 min)

Our data are Kobe Bryant's game-level records compiled from Basketball-Reference.
Before extracting anything:

- **Source & permission:** we did not collect these box scores. Check the site's
  terms and `robots.txt`; today's workbook was prepared for you, and its
  `variable_names` sheet documents the fields.
- **Misuse without context (CLO 12.2):** raw point totals can mislead when
  seasons have different lengths or a player misses games.

> *One sentence: name a way sports stats get misread when context is dropped.*

# 1 · The central question

> **How did Kobe Bryant's scoring performance change across his first five
> NBA seasons?**

We'll measure performance as **points per minute**. A rate lets us compare
seasons with different numbers of games and minutes.

# 2 · E — Extraction (15 min)

**LM1 — the workbook.** Import `Data/nba_goats.xlsx` with `read_excel()`.
First find out what sheets exist.

```{r}
library(tidyverse)
library(readxl)

# LM1a: list the sheets in the workbook
excel_sheets("____")
```

```{r}
# LM1b: read Kobe's game-level sheet
nba_df <- read_excel("____", sheet = "____")
```

**LM2 — inspect.** Confirm what you have: which variables record minutes and
points, and how many rows are in each season?

```{r}
# LM2: inspect columns, then count rows per season
```

# 3 · T — Transformation (27 min)

**LM3 — filter.** Keep Kobe's first five seasons (`season_1` through `season_5`)
and rows with recorded minutes and points.

```{r}
# LM3: filter to the first five seasons and complete game records
kobe_df <- nba_df |>
  filter(____)
```

> *Check yourself: `nrow(kobe_df)` should be 334 complete game rows.*

**LM4 — select + mutate.** Keep the season and performance variables, then
create `season_number` and `ppm = PTS / MP`.

```{r}
# LM4: select what matters and create the rate
```

**LM5 — arrange + summarize.** Group by season and produce the season-level
table the visualization will draw from: season, games, points, minutes, and
points per minute. Arrange chronologically and round `ppm` to 3 digits.

```{r}
# LM5: summarize and arrange by season_number; print the table
kobe_table_df <- ____

kobe_table_df
```

# 4 · V — Visualization (20 min)

**LM6 — geom_point over time.** Plot `ppm` against `season_number` as points.
Time runs left to right; performance runs up.

```{r}
# LM6: time-ordered point plot
ggplot(kobe_table_df, aes(x = ____, y = ____)) +
  geom_point(size = 3) +
  labs(x = ____, y = ____, title = ____)
```

*(If time: add a second plot of raw `points` by season. Which season looks
different across the two plots, and why? Hint: `season_3` was a 50-game
lockout season.)*

# 5 · Interpretation & final answer (10 min)

Answer the central question in 2–3 sentences using the table AND the plot.
Name: the direction of change, the size of the change (first vs fifth season
`ppm`), and the season where a raw total would have misled you.

> *Your answer:*
