---
title: "ETV Cycle 02 — Kobe Bryant's First Five Seasons"
subtitle: "STAT/DATA 1810 · In-class activity · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT (assessed via lab receipt + drills)
  CLO 2.2 read_excel (sheets) · CLO 3.1 filter/select/mutate/group_by/summarize
  CLO 4.1 geom_point · CLO 4.5 claim · CLO 12.2 measurement · 12.6 transparency
Timing: 0 Ethics 8' · E 15' · T 27' · V 20' · Answer 10' = 80'
-->

# 0 · Context & ethics (8 min)

Our data are Kobe Bryant's game-level records compiled from Basketball-Reference.

**Talking points.** Check terms of service and `robots.txt` before scraping;
use the prepared workbook and its `variable_names` sheet for provenance. Raw
point totals also depend on games played and minutes, so a rate makes the
comparison more defensible.

# 1 · The central question

> **How did Kobe Bryant's scoring performance change across his first five
> NBA seasons?** Measure: points per minute.

# 2 · E — Extraction (15 min)

```{r}
library(tidyverse)
library(readxl)

# LM1a: sheets
excel_sheets("Data/nba_goats.xlsx")
```

**Expected:** `KB`, `LJ`, `MJ`, `variable_names`, and `divisions`. The `KB` sheet
contains Kobe's game-level records; the metadata sheet documents the fields.

```{r}
# LM1b: read Kobe's game-level data
nba_df <- read_excel("Data/nba_goats.xlsx", sheet = "KB")
```

```{r}
# LM2: inspect
 glimpse(nba_df)
 nba_df |> count(Season)
```

**Expected:** 1,156 rows and 38 variables. The performance fields are `MP`
(minutes played) and `PTS` (points). The season counts include `season_1` with
82 rows but only 71 complete game records, because some rows have no box-score
values; that is why LM3 filters for non-missing performance values.

# 3 · T — Transformation (27 min)

```{r}
# LM3: first five seasons with complete game records
kobe_df <- nba_df |>
  filter(Season %in% paste0("season_", 1:5),
         !is.na(MP), !is.na(PTS))

nrow(kobe_df)
```

**Expected:** 334 complete game rows.

```{r}
# LM4: select what matters and create the rate
kobe_df <- kobe_df |>
  select(Season, MP, PTS) |>
  mutate(season_number = readr::parse_number(Season),
         ppm = PTS / MP)
```

```{r}
# LM5: summarize and arrange by season_number
kobe_table_df <- kobe_df |>
  group_by(Season, season_number) |>
  summarize(games = n(),
            points = sum(PTS),
            minutes = sum(MP),
            ppm = points / minutes,
            .groups = "drop") |>
  arrange(season_number) |>
  mutate(ppm = round(ppm, 3))

kobe_table_df
```

**Expected table:**

| Season | games | points | minutes | ppm |
|---|---:|---:|---:|---:|
| season_1 | 71 | 539 | 1087.70 | 0.496 |
| season_2 | 79 | 1220 | 2042.93 | 0.597 |
| season_3 | 50 | 996 | 1882.57 | 0.529 |
| season_4 | 66 | 1485 | 2509.82 | 0.592 |
| season_5 | 68 | 1938 | 2778.23 | 0.697 |

# 4 · V — Visualization (20 min)

```{r}
# LM6: time-ordered point plot
ggplot(kobe_table_df, aes(x = season_number, y = ppm)) +
  geom_point(size = 3, color = "#12263A") +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Season number", y = "Points per minute",
       title = "Kobe's scoring rate climbed across his first five seasons")
```

*(Optional companion plot — the measurement beat made visible:)*

```{r}
ggplot(kobe_table_df, aes(x = season_number, y = points)) +
  geom_point(size = 3, color = "#A63D2F") +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Season number", y = "Total points",
       title = "Raw totals dip in season 3, the 50-game lockout season")
```

**Teaching note:** the two plots disagree about season 3. Totals decline
because the season had only 50 games; the rate remains near the surrounding
seasons. That disagreement is the measurement lesson.

# 5 · Interpretation & final answer (10 min)

**Model answer:** *"Kobe's scoring rate rose from about 0.496 points per minute
in season 1 to about 0.697 in season 5, an increase of roughly 41%. The climb
was not perfectly steady: season 3 has fewer total points because the lockout
limited the schedule to 50 games, while its per-minute rate remained near 0.53.
A raw total would have misled us there."*

Required elements: **direction, magnitude, and the lockout caveat** —
direction/size/caveat is the CLO 4.5 trio for this cycle.
