---
title: "ETV Cycle 02 — Kobe Bryant's First Five Seasons"
subtitle: "STAT/DATA 1810 · In-class activity · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right-body
    toc-title: "Contents"
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT (assessed via lab receipt + drills)
  CLO 2.2 read_excel (sheets) · CLO 3.1 filter/select/mutate/group_by/summarize
  CLO 4.1 geom_point · CLO 4.5 claim · CLO 12.2 measurement · 12.6 transparency
Timing: 0 Ethics 8' · E 15' · T 27' · V 20' · Answer 10' = 80'
-->

::: {.callout-warning title="Timing & CLOs"}
- **Budget:** Ethics 8′ · E 15′ · T 27′ · V 20′ · Answer 10′ = **80′**
- **CLOs:** 2.2 `read_excel` (sheets) · 3.1 filter/select/mutate/group_by/summarize · 4.1 `geom_point` · 4.5 claim · 12.2 measurement · 12.6 transparency
:::

# 0 · Context & ethics [8 min]{.timing}

::: {.callout-important title="Talking points"}
- Our data: Kobe Bryant's **game-level records** compiled from Basketball-Reference
- Check terms of service and `robots.txt` before scraping
- Use the prepared workbook and its `variable_names` sheet for provenance
- Raw point totals depend on games played and minutes → a **rate** makes the comparison more defensible
:::

# 1 · The central question

> **How did Kobe Bryant's scoring performance change across his first five
> NBA seasons?** Measure: points per minute.

# 2 · E — Extraction [15 min]{.timing}

```{r}
library(tidyverse)
library(readxl)

# LM1a: sheets
excel_sheets("Data/nba_goats.xlsx")
```

::: {.callout-note title="Expected — LM1a"}
- [LM1a]{.lm} sheets: `KB`, `LJ`, `MJ`, `variable_names`, `divisions`
- `KB` holds Kobe's game-level records; `variable_names` documents the fields
:::

```{r}
# LM1b: read Kobe's game-level data
nba_df <- read_excel("Data/nba_goats.xlsx", sheet = "KB")
```

```{r}
# LM2: inspect
 glimpse(nba_df)
 nba_df |> count(Season)
```

::: {.callout-note title="Expected — LM2"}
- [LM2]{.lm} **1,156 rows, 38 variables**
- Performance fields: `MP` (minutes played) and `PTS` (points)
- `season_1` shows 82 rows but only **71 complete** game records — some rows have no box-score values
- That's why LM3 filters for non-missing performance values
:::

# 3 · T — Transformation [27 min]{.timing}

```{r}
# LM3: first five seasons with complete game records
kobe_df <- nba_df |>
  filter(Season %in% paste0("season_", 1:5),
         !is.na(MP), !is.na(PTS))

nrow(kobe_df)
```

::: {.callout-note title="Expected — LM3"}
- [LM3]{.lm} **334** complete game rows
:::

```{r}
# LM4: select what matters and create the rate
kobe_df <- kobe_df |>
  select(Season, MP, PTS) |>
  mutate(season_number = readr::parse_number(Season),
         ppm = PTS / MP)
```

```{r}
# LM5: summarize and arrange by season_number
kobe_table_df <- kobe_df |>
  group_by(Season, season_number) |>
  summarize(games = n(),
            points = sum(PTS),
            minutes = sum(MP),
            ppm = points / minutes,
            .groups = "drop") |>
  arrange(season_number) |>
  mutate(ppm = round(ppm, 3))

kobe_table_df
```

::: {.callout-note title="Expected table — LM5"}
| Season | games | points | minutes | ppm |
|---|---:|---:|---:|---:|
| season_1 | 71 | 539 | 1087.70 | 0.496 |
| season_2 | 79 | 1220 | 2042.93 | 0.597 |
| season_3 | 50 | 996 | 1882.57 | 0.529 |
| season_4 | 66 | 1485 | 2509.82 | 0.592 |
| season_5 | 68 | 1938 | 2778.23 | 0.697 |
:::

# 4 · V — Visualization [20 min]{.timing}

```{r}
# LM6: time-ordered point plot
ggplot(kobe_table_df, aes(x = season_number, y = ppm)) +
  geom_point(size = 3, color = "#12263A") +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Season number", y = "Points per minute",
       title = "Kobe's scoring rate climbed across his first five seasons")
```

*(Optional companion plot — the measurement beat made visible:)*

```{r}
ggplot(kobe_table_df, aes(x = season_number, y = points)) +
  geom_point(size = 3, color = "#A63D2F") +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Season number", y = "Total points",
       title = "Raw totals dip in season 3, the 50-game lockout season")
```

::: {.callout-important title="Teaching note — the two plots disagree"}
- The plots disagree about **season 3**
- Totals decline because the season had only **50 games** (lockout)
- The rate stays near the surrounding seasons
- That disagreement **is** the measurement lesson (CLO 12.2)
:::

# 5 · Interpretation & final answer [10 min]{.timing}

::: {.callout-tip title="Model answer"}
- *"Kobe's scoring rate rose from about 0.496 points per minute in season 1 to about 0.697 in season 5, an increase of roughly 41%. The climb was not perfectly steady: season 3 has fewer total points because the lockout limited the schedule to 50 games, while its per-minute rate remained near 0.53. A raw total would have misled us there."*
- Required elements (the CLO 4.5 trio for this cycle):
  - **direction** — rate rose
  - **magnitude** — ≈ 0.496 → 0.697, about +41%
  - **caveat** — the lockout season
:::
