---
title: "Lab 02 — Types, Structures, and Looking Inside Data"
subtitle: "STAT/DATA 1810 · Take-home (AI permitted) · Student version"
author: "Your name here"
date: today
format:
  html:
    toc: true
    embed-resources: true
  docx: default
execute:
  warning: false
  message: false
---

```{=html}
<!--
CLO ALIGNMENT (Foundations 02 + 03; see course website → PLOs & CLOs)
  CLO 1  1.4 assignment · 1.5 data types (numeric/character/logical/factor/date)
         · 1.6 containers (vector, data frame/tibble, list) · 1.12 storage
         magnitudes (object.size, KB/MB/GB reasoning)
  CLO 7  7.4 comparison operators · 7.6 is.na()
  CLO 11 11.1 ?function help · 11.x AI-assisted code reading
  CLO 9  practiced, not graded (this IS a rendered Quarto document)
Datasets: airquality, iris (built in) + tibbles you build about Cal Poly.
Target time: 60–90 min. 16 questions.
HOW TO SUCCEED: fill every blank (____), complete every empty chunk, and
answer every "> Your answer:" prompt. The lab must RENDER cleanly to submit.
-->
```

# Welcome to Lab 02

This lab covers **Foundations 02 and 03**: what a value *is* (its type), what
*holds* values (its structure), how to look inside a dataset before you touch
it, how to reach in and take a piece, and how big data actually is. The
examples are Cal Poly examples — lattes near campus, Rec Center headcounts —
plus two built-in datasets, `airquality` and `iris`.

**What this lab checks** (the course learning outcomes it maps to):

| You will show that you can… | CLO |
|---|---|
| tell a data *type* from a data *structure*, and name the type of any value | 1.5, 1.6 |
| build a vector, a tibble, and a list from scratch — and predict what breaks | 1.4, 1.6 |
| explain why a factor is not just text, and put its levels in order | 1.5 |
| inspect an unfamiliar data frame (`glimpse`, `summary`, `dim`) before computing on it | 1.6 |
| pull out a column with `$`, count missing values with `is.na()`, and read a pipe `|>` as "and then" | 1.6, 7.4, 7.6 |
| estimate an object's size in KB/MB/GB and reason about whether it fits in memory | 1.12 |
| read a help page, and read AI-written code you did not write | 11.1 |
| render a reproducible document from a fresh session | 9 |

**The habit from Lab 01 continues.** Every 👀 **Variables check** means: look
at the Variables pane and answer. Before you run anything it should be empty
— if it isn't, restart R.

```{r}
#| label: setup
library(tibble)   # tibble(), tribble(), glimpse() — all this lab needs
```

> 👀 **Variables check.** Lab 01 loaded `tidyverse`; this lab loads only
> `tibble`. Neither added anything to Variables. In one sentence, why do we
> load only what a document needs?
>
> Your answer:

------------------------------------------------------------------------

# Part A · Data types (Q1–Q4)

**Q1.** Predict, then run. Write your four predictions *before* running the
chunk, using the type table from Foundations 02.

> Predictions: `5.45` ____ · `"Scout Coffee"` ____ · `38L` ____ · `5.45 > 4.75` ____

```{r}
# Q1: the four types you'll live with
class(5.45)             # a Starbucks latte in the UU
class("Scout Coffee")   # where the latte came from
class(38L)              # people in the Rec Center at 7 am
class(5.45 > 4.75)      # is the UU latte pricier than Julian's?
```

> Which one, if any, surprised you, and why?
>
> Your answer:

**Q2.** Families. For each variable, write **Q** (quantitative), **C**
(categorical), or **O** (other — a date or time), and name the R type you'd
expect it to be stored as.

> -   the price of your latte: ____ / ____
> -   the shop that made it: ____ / ____
> -   the date you bought it: ____ / ____
> -   whether it was iced: ____ / ____
> -   the number of people in the Rec Center at 5 pm: ____ / ____

**Q3.** Factor vs character. The Rec Center logs a time slot for every
headcount. Run the chunk and study the two `table()` outputs.

```{r}
# Q3: same words, two types
slot_chr <- c("evening", "morning", "afternoon", "morning", "evening", "evening")
slot_fct <- factor(slot_chr, levels = c("morning", "afternoon", "evening"))
class(slot_chr)
class(slot_fct)
table(slot_chr)
table(slot_fct)
```

> 👀 **Variables check.** How does the Variables pane describe `slot_fct`
> differently from `slot_chr`? ____
>
> -   The two tables contain the same counts but in a different order. What
>     decides the order in each? ____
> -   Now add a typo on purpose: in the Console run
>     `factor(c("morning", "Morning"), levels = c("morning", "afternoon", "evening"))`.
>     What did the capital-M value become, and did R give an error? ____
> -   One sentence: when should a column be a factor instead of character?
>
> Your answer:

**Q4.** Help page. Open `?factor` and read the **Arguments** section.

```{r}
#| eval: false
# Q4: open the help page (no output when rendered — that's fine)
?factor
```

> What does the `levels` argument control (one sentence)? What happens, by
> default, if you leave it out — in what order do the levels come?
>
> Your answer:

------------------------------------------------------------------------

# Part B · Data structures from scratch (Q5–Q8)

**Q5.** Vectors and coercion. Build three vectors about lattes at Julian's,
the UU Starbucks, and Scout Coffee, then break one on purpose.

```{r}
# Q5: three vectors, one per column-to-be
shop  <- c("Julian's", "Starbucks", "Scout Coffee")
price <- c(____, ____, ____)          # numbers — what a latte costs at each
iced  <- c(____, ____, ____)          # TRUE / FALSE — would you get it iced?
class(shop); class(price); class(iced)

# now the break: put quotes around ONE price and check the class
price_broken <- c(____, "____", ____)
class(price_broken)
```

> 👀 **Variables check.** `price` and `price_broken` have the same length.
> What does Variables say is different about them? ____
>
> Why did one pair of quotes change the type of *all three* values? Use the
> word from class.
>
> Your answer:

**Q6.** A tibble from your vectors, then a deliberate error.

```{r}
# Q6: build lattes from the vectors in Q5, then inspect it
lattes <- tibble(shop = ____, price = ____, iced = ____)
lattes
nrow(lattes); ncol(lattes); class(lattes)
```

```{r}
#| eval: false
# Q6b: make it FAIL — give price only two values. Copy the error message
# into your answer below, then set this chunk back to eval: false (or delete it)
tibble(shop = shop, price = c(____, ____))
```

> The error message names two things: the sizes involved and the column that
> broke. Copy the message here and underline (or quote) those two parts:
>
> Your answer:
>
> 👀 **Variables check.** Expand `lattes`. Each thing inside has its own type
> and a length of 3. What rule about data frames does that picture show?

**Q7.** A list. Build one that holds text, a number, and your whole `lattes`
tibble.

```{r}
# Q7: a list called me
me <- list(first = "____", n_classes = ____, lattes = ____)
me
length(me)
```

> Could `c()` hold these three things without changing any of them? Try
> `c("Eman", 4)` in the Console — what happened to the 4, and what does that
> tell you a list can do that a vector cannot?
>
> Your answer:

**Q8.** In one sentence each, define **data type** and **data structure**,
using one example of each *from this lab* — at least one must be an object
you built.

> Your answer:

------------------------------------------------------------------------

# Part C · Inspecting & accessing (Q9–Q12)

**Q9.** You've been handed `airquality`. Run the three lines every dataset
gets first.

```{r}
# Q9: inspect before you compute
glimpse(airquality)
summary(airquality)
dim(airquality)
```

> -   Rows: ____ · Columns: ____ · One row represents: ____
> -   Type tag on every column (from `glimpse`): ____ — quantitative or
>     categorical? Is `Month` *really* quantitative? One sentence.
> -   Which two columns have missing values, and how many each (from
>     `summary`)? ____ and ____
>
> 👀 **Variables check.** `airquality` is not listed even though you just
> used it three times. Where does it live? ____

**Q10.** `$` and `is.na()`. Fill the blanks.

```{r}
# Q10: one column at a time
mean(airquality$Temp)
mean(airquality$Ozone)                       # returns NA — why?
mean(airquality$Ozone, ____ = TRUE)          # fix it with the argument from ?mean
____(is.na(airquality$Ozone))                # how many Ozone values are missing?
sum(airquality$Temp > ____)                  # how many days were above 90°F?
```

> -   Mean Temp: ____ · Mean Ozone once fixed: ____ · Missing Ozone: ____ ·
>     Days above 90: ____
> -   `sum(airquality$Temp > 90)` counts days. What is R adding up, and what
>     is each `TRUE` worth?
>
> Your answer:

**Q11.** `[rows, columns]` — just enough. Write ONE line for each.

```{r}
# Q11: rectangles
____                    # (a) rows 1–5, only the Ozone and Temp columns
____                    # (b) the LAST row of airquality — use nrow(), not 153
____                    # (c) rows 51–53 of iris, all columns
```

> (c) shows which species? ____ · Why do we write `nrow(airquality)` inside
> the brackets instead of typing 153?
>
> Your answer:

**Q12.** Access, then assign. Save July's temperatures as their own object.

```{r}
# Q12: pull July out of airquality
july_temp <- airquality$Temp[airquality$Month == ____]
length(july_temp)
mean(july_temp)
```

> 👀 **Variables check.** `july_temp` appears; `airquality` still does not.
> What is the difference between the two, in terms of where each came from?
> ____ · Length of `july_temp`: ____ · Mean July temperature: ____

------------------------------------------------------------------------

# Part D · The pipe and the size of data (Q13–Q15)

**Q13.** Rewrite each nested line as a pipe, **one step per line**, and
confirm the answer is unchanged. Then read one of them aloud as a sentence
with "and then."

```{r}
# Q13: nested -> piped
round(mean(airquality$Ozone, na.rm = TRUE), 1)
# your pipe version:


length(unique(iris$Species))
# your pipe version:


head(which(is.na(airquality$Ozone)))
# your pipe version:

```

> Write the second one as a sentence: "Take ____, and then ____, and then
> ____."
>
> In the first one, `na.rm = TRUE` stays inside `mean()` even in the pipe
> version. Why — what does the pipe fill in, and what does it leave alone?
>
> Your answer:

**Q14.** How big are our objects?

```{r}
# Q14: bytes
object.size(airquality)
object.size(iris)
object.size(lattes)
```

> -   In KB (divide by 1,000): airquality ≈ ____ · iris ≈ ____ · lattes ≈ ____
> -   Which unit from the ladder are all three in — bytes, KB, MB, or GB? ____
> -   One phone photo is about 3 MB. About how many `airquality` datasets fit in
>     one photo? ____

**Q15.** Reason at scale. `airquality` is 153 rows and about 5.6 KB. A weather
station network records the same six columns for **every hour, at 1,000
stations, for a year** (about 8.8 million rows).

> -   Roughly how big would that be? Show the multiplication and give the
>     answer with a unit: ____
> -   Your laptop has 8 GB of memory. Does that dataset fit? Would ten years
>     of it fit? What would you check *before* trying to import it?
>
> Your answer:

------------------------------------------------------------------------

# Part E · Working with AI (Q16)

**Q16.** As in Lab 01, ask an AI assistant to write code you have not fully
learned yet, run it, and explain it.

**Step 1 — copy this exact prompt to your AI:**

> *"In R, using only base R and the tibble package, take the built-in
> airquality data frame, convert its Month column to a factor whose levels are
> the month names May through September, and then use the pipe |> to produce
> a table of how many days are in each month. Explain nothing — just give me
> the code."*

**Step 2 — paste the AI's code below and run it.** If it errors, give the AI
the error message and paste the working version.

```{r}
# Q16: paste the AI-generated code here and run it

```

> 👀 **Variables check.** Did the AI's code *modify* `airquality`, create a
> new object, or print without saving? Use the Variables pane to decide, and
> say how you know.
>
> Your explanations (one plain-English line per piece of the code):
>
> 1.  
>
> 2.  
>
> 3.  
>
> Which line uses something from THIS lab (factor levels, `|>`, `$`)? Which
> line uses something you have not seen yet?

------------------------------------------------------------------------

# Finish: render your lab (required)

1.  Save the file into `R/Labs/` as `lab02_yourlastname.qmd`.
2.  **Restart R** (top-right session menu → Restart). Watch Variables empty
    out. Then click **Render** — once to **HTML** and once to **Word**.
3.  If rendering fails, the error names the chunk. The usual causes: a
    leftover `____`, an empty required chunk, the Q6b chunk not set back to
    `eval: false`, or code that only worked because of something typed in the
    Console.

**Then:** go to Canvas and take the **Lab 02 Quiz**. It asks for values and
answers you produced above.
