---
title: "Lab 02 — Types, Structures, and Looking Inside Data"
subtitle: "STAT/DATA 1810 · Take-home (AI permitted) · SOLUTION"
author: "Instructor key"
date: today
format:
  html:
    toc: true
    embed-resources: true
  docx: default
execute:
  warning: false
  message: false
---

<!--
CLO: 1.4, 1.5, 1.6, 1.12 · 7.4, 7.6 · 11.1 · 9 (practiced)
Grading: this .qmd is completion/receipt (rendered cleanly = credit). The
POINTS come from the Canvas quiz (make_canvas_quiz.py), whose answers can only
be produced by completing this file. Values below feed the quiz key.
The 👀 prompts are habit-building; grade for a genuine attempt.
-->

```{r}
#| label: setup
library(tibble)
```

**Setup 👀.** `library()` loads functions, creates no object. We load only
`tibble` because that is all the document uses — loading nine packages for
one function is slow, hides which package a function came from, and is a
habit that breaks the moment two packages share a name.

# Part A · Data types

**Q1.**
```{r}
class(5.45); class("Scout Coffee"); class(38L); class(5.45 > 4.75)
```
`"numeric"` · `"character"` · `"integer"` · `"logical"`. Usual surprises: `38L`
is integer (the `L`), and a comparison *produces* a logical.

**Q2.** price Q/numeric · shop C/character (accept factor) · date O/Date ·
iced C/logical · headcount Q/numeric (accept integer).

**Q3.**
```{r}
slot_chr <- c("evening", "morning", "afternoon", "morning", "evening", "evening")
slot_fct <- factor(slot_chr, levels = c("morning", "afternoon", "evening"))
class(slot_chr); class(slot_fct)
table(slot_chr); table(slot_fct)
```
👀 Variables: `slot_chr` is *character [6]*; `slot_fct` is *factor with 3
levels*. Order: the character table is **alphabetical**; the factor table
follows the **levels you gave**. Typo: `"Morning"` becomes **`NA`, with no
error** — the value isn't on the menu. Rule: use a factor when a column is a
**category with a known set of values** (and when order matters for
tables/plots); keep free text as character.

**Q4.** `levels` sets the allowed values *and their order*. Left out, the
levels are the unique values **sorted alphabetically**.

# Part B · Data structures from scratch

**Q5.**
```{r}
shop  <- c("Julian's", "Starbucks", "Scout Coffee")
price <- c(4.75, 5.45, 5.95)
iced  <- c(TRUE, FALSE, TRUE)
class(shop); class(price); class(iced)
price_broken <- c(4.75, "5.45", 5.95)
class(price_broken)
```
👀 same length 3; `price` is numeric, `price_broken` is **character**.
**Coercion**: a vector holds one type, so R converts everything to the type
that can hold all values — text.

**Q6.**
```{r}
lattes <- tibble(shop = shop, price = price, iced = iced)
lattes
nrow(lattes); ncol(lattes); class(lattes)
```
3 · 3 · `"tbl_df" "tbl" "data.frame"`. Q6b error: *"Tibble columns must have
compatible sizes. • Size 3: Existing data. • Size 2: Column `price`."* — the
sizes (3 vs 2) and the column (`price`). 👀 Expanding shows three vectors of
length 3: **every column is a vector, and all columns have the same length**.

**Q7.**
```{r}
me <- list(first = "Eman", n_classes = 4, lattes = lattes)
me
length(me)
```
`c("Eman", 4)` → `"Eman" "4"`: the 4 became text. A list keeps each part's own
type (and can hold a whole tibble); a vector coerces to one type.

**Q8.** Model: "A data type is what one value is — `5.45` is numeric. A data
structure is the container — my `lattes` tibble is a data frame whose `iced`
column is logical." Any correct type + structure pair, one self-built.

# Part C · Inspecting & accessing

**Q9.**
```{r}
glimpse(airquality); summary(airquality); dim(airquality)
```
**153 rows × 6 columns**; one row = **one day** (NYC, May–Sep 1973). Every
column is `<int>` — stored quantitative, but `Month` (and `Day`) is really a
**label**, i.e. categorical/other. Missing: **Ozone 37**, **Solar.R 7**. 👀
`airquality` ships in the `datasets` package — Variables shows only what this
session created.

**Q10.**
```{r}
mean(airquality$Temp)
mean(airquality$Ozone)
mean(airquality$Ozone, na.rm = TRUE)
sum(is.na(airquality$Ozone))
sum(airquality$Temp > 90)
```
**77.88** · `NA` because any NA poisons `mean()` · **42.13** · **37** ·
**14 days**. `Temp > 90` is a logical vector; `sum()` counts each `TRUE` as
**1**.

**Q11.**
```{r}
airquality[1:5, c("Ozone", "Temp")]
airquality[nrow(airquality), ]
iris[51:53, ]
```
(c) **versicolor**. `nrow()` inside the bracket keeps working when the data
changes size — a hard-coded 153 silently returns the wrong row tomorrow.

**Q12.**
```{r}
july_temp <- airquality$Temp[airquality$Month == 7]
length(july_temp); mean(july_temp)
```
👀 `july_temp` was created *in this session* with `<-`; `airquality` was not.
Length **31** · mean **83.9°F**.

# Part D · The pipe and the size of data

**Q13.**
```{r}
airquality$Ozone |> mean(na.rm = TRUE) |> round(1)      # 42.1
iris$Species |> unique() |> length()                    # 3
airquality$Ozone |> is.na() |> which() |> head()        # 5 10 25 26 27 32
```
Sentence: "Take the Species column, and then keep the unique values, and then
count them." The pipe fills only the **first argument** of the next function;
every other argument (`na.rm = TRUE`) stays where you wrote it.

**Q14.**
```{r}
object.size(airquality); object.size(iris); object.size(lattes)
```
airquality ≈ **5.6 KB** · iris ≈ **7.3 KB** · lattes ≈ **1–2 KB** (accept
any value in the KB range; depends on strings). All in **KB**. A 3 MB photo
holds about 3,000 / 5.6 ≈ **500** airquality datasets (accept 400–600).

**Q15.** 8.8 million rows ÷ 153 rows ≈ 57,500 × 5.6 KB ≈ **320 MB** (accept
anything from ~300 MB to ~350 MB, or "a few hundred MB"). One year **fits**
in 8 GB; ten years ≈ 3.2 GB **also fits**, though with little room to work
(R copies objects). Check `file.size()` / the file's size on disk *before*
importing, and compare to memory — the E stage's first question.

# Part E · Working with AI

**Q16.** Representative working answer:

```{r}
aq <- airquality
aq$Month <- factor(aq$Month,
                   levels = 5:9,
                   labels = c("May", "June", "July", "August", "September"))
aq$Month |> table()
```

Expected: May 31 · June 30 · July 31 · August 31 · September 30. 👀 Most AI
answers either modify a *copy* (`aq`, a new object appears) or modify
`airquality` directly (a **local copy** of `airquality` now appears in
Variables — the built-in one is untouched); either observation is correct if
it matches the pane. From this lab: `factor(levels = ...)`, `$`, `|>`.
New: the `labels =` argument (or `month.name`, or `dplyr::count()` if the AI
ignored the base-R instruction — fine, they should notice and say so).

**Grading Q16:** credit for running it, getting it to work, and an honest
plain-English read of each line. Naming `labels =` or `table()` as "the thing
I haven't seen" is the intended outcome.

# Rendering & Canvas

Confirm the render to **both** HTML and docx, after **Restart R**. The Canvas
quiz (`make_canvas_quiz.py`) is the graded artifact; this file is the gate.
