---
title: "ETV Cycle 03 — [Crime Data: Title TBD]"
subtitle: "STAT/DATA 1810 · In-class activity · SOLUTION (TEMPLATE)"
author: "Instructor key"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right-body
    toc-title: "Contents"
execute:
  warning: false
  message: false
  eval: false
---

<!--
TEMPLATE STATUS: fill every [bracket] once the dataset is chosen; then flip
execute.eval to true and confirm the expected outputs written below.

Candidate datasets (pick one, then commit):
  [ ] SLO PD open-data incident log (city portal export → saved as .RData)
  [ ] A state open-data portal extract
  [ ] crimedata package sample (pre-cleaned; weakest ethics discussion)
Selection criteria: real, local if possible, no names/addresses at unit level,
a timestamp variable (hour-of-day question works every year), ≥ ~5k rows.

CLO ALIGNMENT
  CLO 2.7 .RData extraction · CLO 3.1 select/filter/mutate/group/summarize/arrange
  CLO 4.1 geom_point · 4.3 counts · 4.5 claim
  CLO 12: recommended set = Privacy (12.4), Representation (12.1),
  Measurement (12.2, reporting/enforcement), Transparency (12.6),
  Consequences (12.7)
Timing: Ethics 10' · E 12' · T 28' · V 20' · Answer 10' = 80'
-->

::: {.callout-caution title="Template status"}
- Fill every `[bracket]` once the dataset is chosen, then flip `execute.eval` to `true` and confirm the expected outputs below
- Candidate datasets (pick one, then commit):
  - SLO PD open-data incident log (city portal export → saved as `.RData`)
  - a state open-data portal extract
  - `crimedata` package sample (pre-cleaned; weakest ethics discussion)
- Selection criteria: real · local if possible · no names/addresses at unit level · a timestamp variable (hour-of-day works every year) · ≥ ~5k rows
:::

::: {.callout-warning title="Timing & CLOs"}
- **Budget:** Ethics 10′ · E 12′ · T 28′ · V 20′ · Answer 10′ = **80′**
- **CLOs:** 2.7 `.RData` extraction · 3.1 select/filter/mutate/group/summarize/arrange · 4.1 `geom_point` · 4.3 counts · 4.5 claim
- **CLO 12 set:** Privacy (12.4) · Representation (12.1) · Measurement (12.2) · Transparency (12.6) · Consequences (12.7)
:::

# 0 · Context & ethics [10 min]{.timing} — instructor script

::: {.callout-important title="Instructor script — five beats"}
- Run all five beats; the two non-negotiables:
  - **Measurement** — these records count *reports and enforcement*, not crime itself
  - **Consequences** — a chart of this data can move policy, prices, and policing; the title must say "reported"
- Model sentence for the student prompt: *"Because unreported offenses never enter the data, every count is a floor, not a level — and the floor is lower in communities that distrust reporting."*
:::

# 1 · The central question

> **[COMMITTED QUESTION — recommended shape: "How do reported [offense
> category] incidents vary by hour of day?" It exercises the timestamp, gives
> geom_point a natural x-axis, and regenerates every semester with new data.]**

# 2 · E — Extraction [12 min]{.timing}

```{r}
library(tidyverse)

# LM1: load — note for students: load() puts objects INTO the environment;
# there is no assignment arrow, which is exactly why we inspect ls() next.
load("Data/[FILENAME].RData")
ls()
```

::: {.callout-note title="Expected — LM1"}
- [LM1]{.lm} `[OBJECT_NAME]` appears in `ls()`
- Contrast with `read_csv()` on the board: *read functions return, `load()` injects*
- That one line is the whole CLO 2.7 lesson
:::

```{r}
# LM2: inspect
glimpse([OBJECT_NAME])
```

::: {.callout-note title="Expected — LM2"}
- [LM2]{.lm} [rows] · [date range]
- One row = one reported incident
:::

# 3 · T — Transformation [28 min]{.timing}

```{r}
# LM3: select
crime_small_df <- [OBJECT_NAME] |>
  select(offense, date_time, [neighborhood_or_beat])
```

```{r}
# LM4: filter — committed rule: [e.g., offense %in% c(...), complete
# date_time, incidents within the chosen year]
crime_clean_df <- crime_small_df |>
  filter([RULE])

nrow(crime_small_df) - nrow(crime_clean_df)
```

::: {.callout-note title="Expected — LM4"}
- [LM4]{.lm} [n removed]
- Defense: [state it]
- Connect to underreporting: we can defend removing *unusable records*; we cannot recover *unreported incidents* — the answer must say so
:::

```{r}
# LM5: recode/create — recommended: hour of day
crime_clean_df <- crime_clean_df |>
  mutate(hour = lubridate::hour(date_time))
```

::: {.callout-important title="LM5 note"}
- [LM5]{.lm} If lubridate hasn't formally arrived yet, this is a one-function toolbox loan — name it, use it, move on
:::

```{r}
# LM6: group + summarize + arrange
crime_summary_df <- crime_clean_df |>
  group_by(hour) |>
  summarize(n = n()) |>
  arrange(desc(n))

crime_summary_df
```

::: {.callout-note title="Expected — LM6"}
- [LM6]{.lm} [peak hours — typically late evening for most offense mixes]
:::

# 4 · V — Visualization & summary [20 min]{.timing}

```{r}
# LM8: the point plot
ggplot(crime_summary_df, aes(x = hour, y = n)) +
  geom_point(size = 3, color = "#12263A") +
  labs(
    x = "Hour of day",
    y = "Number of reported incidents",
    title = "Reported [offense] incidents peak in the [period]",
    subtitle = "Reports, not occurrences — see limitations"
  )
```

::: {.callout-note title="Expected pattern — LM8"}
- [LM8]{.lm} [describe once committed]
- The subtitle is **mandatory** — it is the Transparency beat (12.6) enforced in the artifact itself
:::

# 5 · Interpretation & final answer [10 min]{.timing} — model answer shape

> *"Reported [offense] incidents are concentrated in [period], with roughly
> [X]× as many reports at the peak hour as at the quietest hour. This pattern
> describes when incidents are REPORTED, which reflects both when events occur
> and when people or patrols are positioned to report them — enforcement and
> reporting behavior are baked into every count. [One consequence sentence:
> what this chart should and should not be used to decide.]"*

::: {.callout-tip title="Trio check (CLO 4.5)"}
- **pattern** ✔
- **magnitude** ✔
- **caveat** ✔
- Same rubric as Cycles 01 and 02 — now carrying the heaviest ethical weight of the three
:::
