---
title: "Foundations 03 — Inspecting and Accessing Data"
subtitle: "STAT/DATA 1810 · Week 2 Wednesday · Student version"
author: "Your name here"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT
  CLO 1 (1.6 containers in use · 1.12 storage magnitudes, object.size/file.size)
  CLO 1 accessing: $, [ ], [[ ]]
  CLO 7 (7.6 is.na in practice)
  CLO 9 practiced: you insert your own chunks (➕)
Time: 80 min. Datasets: airquality, iris, mtcars (built in).
Ends with a 5-minute live preview of a full ETV cycle — where all of this is going.
LM = learning moment. YT = your turn. Times in headers are the section budget.
-->

```{r}
#| label: setup
library(tidyverse)
```

# 0 · Today's job (2 min)

You get handed a dataset. Before any analysis, you must answer three
questions: **What is this? How big is it? What's inside?** Today is those
three questions, as functions — and then, once you know what's inside,
how to *reach in and take a piece*.

# 1 · Inspecting data (22 min)

We'll interrogate `airquality` — 1973 New York air measurements, built in.

**LM1 — the first look (5 min).** Run each; note what question each answers:

```{r}
# LM1: five inspection functions
head(airquality)
tail(airquality)
dim(airquality)
nrow(airquality)
ncol(airquality)
```

> *Q1. Rows: ____ · Columns: ____ · What is one OBSERVATION here (what does
> one row represent)?*
>
> *Your answer:*

**LM2 — names and structure (5 min).**

```{r}
# LM2
names(airquality)
str(airquality)
glimpse(airquality)
```

> *Q2. `str()` and `glimpse()` answer the same question. Which is easier to
> read, and what do the `<int>` / `int` tags tell you?*
>
> *Your answer:*

**LM3 — the numeric overview (4 min).**

```{r}
# LM3
summary(airquality)
```

> *Q3. Which TWO variables contain missing values, and how many each?
> (summary() tells you directly.)*
>
> *Your answer:*

*(`View(airquality)` opens a spreadsheet view — try it in the Console; it
doesn't belong in a rendered document.)*

> *YT1 (8 min). ➕ Insert a chunk and inspect a dataset you have NOT seen:
> `mtcars`. Use whichever of the functions above you need to answer:*
> *(a) how many rows and columns?*
> *(b) what is one observation (one row)? Hint: run `head()` and look at the
> row labels on the left.*
> *(c) which columns are NOT numeric — or are they all numeric?*
> *(d) any missing values? How do you know?*
> *(e) what is the median `mpg`?*

<!-- ➕ insert a chunk here -->

> *(a) ____ · (b) ____ · (c) ____ · (d) ____ · (e) ____*

# 2 · Accessing data (33 min)

Three tools, three jobs:

| Tool | Job | Returns |
|---|---|---|
| `$` | one column, **by name** | a vector |
| `[rows, cols]` | a rectangle, by position/name | usually a data frame |
| `[[ ]]` | one element, dug out | the thing itself |

**LM4 — `$`, the workhorse (5 min).**

```{r}
# LM4: a column is a vector
airquality$Temp
mean(airquality$Temp)
```

> *YT2 (4 min). ➕ Insert a chunk. Using `$`, find:*
> *(a) the mean of Wind;*
> *(b) the max Temp;*
> *(c) the mean of Ozone — it comes back `NA`. Fix it with the argument you
> looked up in the help page last week;*
> *(d) how many days had Temp above 90? (Hint: `sum(airquality$Temp > 90)` —
> TRUE counts as 1.)*

<!-- ➕ insert a chunk here -->

> *(a) ____ · (b) ____ · (c) ____ · (d) ____*

**LM5 — `[rows, columns]` (6 min).** Predict each before running:

```{r}
# LM5: predict, then run
airquality[1, ]        # first row, all columns
airquality[, 4]        # all rows, 4th column
airquality[1:5, c("Ozone", "Temp")]
```

> *YT3 (5 min). ➕ Insert a chunk. Write ONE line for each:*
> *(a) rows 10 through 15 of ONLY the Month and Day columns;*
> *(b) the last row of airquality (use `nrow()` inside the brackets — no
> hard-coded 153);*
> *(c) rows 1–3, columns 1–3, by position;*
> *(d) all rows of the Temp column, but as a data frame not a vector (hint:
> leave out the comma).*

<!-- ➕ insert a chunk here -->

**LM6 — `[[ ]]`, digging one thing out (4 min).**

```{r}
# LM6: single element — row 1's Temp
airquality[[1, "Temp"]]

# and on a list:
my_list <- list(nums = c(1, 2, 3), word = "hello")
my_list[["nums"]]
my_list$word
```

**LM7 — the comparison (4 min).** Run all three; look closely at what each
returns:

```{r}
# LM7: three ways at the same column
head(airquality$Temp)          # vector
head(airquality["Temp"])       # still a data frame (note the header!)
head(airquality[["Temp"]])     # vector again
```

> *Q4. Which of the three kept the data frame wrapping? ____*

**LM8 — missing values, located (2 min).**

```{r}
# LM8: WHICH rows are missing Ozone?
which(is.na(airquality$Ozone)) |> head()
```

> *YT4 — access challenge (7 min). ➕ Insert a chunk. Switch to `iris`.*
> *(a) the Species column, as a vector — how long is it?*
> *(b) rows 51–53, all columns — what species are they?*
> *(c) the single value in row 100, column Petal.Width;*
> *(d) the mean Sepal.Length of ONLY the first 50 rows (setosa);*
> *(e) build `my_list2 <- list(flowers = iris, n = nrow(iris))` and pull `n`
> back out with `[[ ]]`, then with `$`. Same answer?*
> *Stretch: which of the three tools did you use most? Why?*

<!-- ➕ insert a chunk here -->

# 3 · How big is data? (10 min)

**LM9 — storage units (4 min).** Data size is measured in bytes:
**byte → KB (thousand) → MB (million) → GB (billion) → TB (trillion)**.
Rough anchors: this class's survey ≈ a few KB · a semester of sensor logs ≈
MBs · a music library ≈ GBs · a national census extract ≈ TBs.

```{r}
# LM9: how big are OUR objects?
object.size(airquality)
object.size(iris)
object.size(mtcars)
```

> *Q5. Order the three from smallest to largest: ____*

**Why it matters:** R holds data **in memory**. A 2 GB file on a 8 GB laptop
imports; a 200 GB file does not — you'd need different tools. File size is
the first thing to check *before* importing, which is exactly where next
week begins.

> *Q6. Your laptop has 8 GB of memory. Roughly how many copies of a 500 MB
> dataset could fit at once (ignore overhead)? ____*

> *YT5 (3 min). ➕ Insert a chunk. Run `object.size(iris)` and then
> `object.size(iris$Sepal.Length)`. Roughly what fraction of the whole is
> one column? Does that match "1 of 5 columns"? One sentence on why or why
> not.*

<!-- ➕ insert a chunk here -->

# 4 · Where this is going (5 min — watch)

Instructor runs one **complete ETV cycle live**, start to finish, no pauses:
import a CSV → transform → plot. Every function you just learned appears in
the first thirty seconds. Monday, you drive.

# 5 · Exit question (8 min)

> *You're handed `mysterydata`, a data frame you've never seen. Write the
> three lines you'd run FIRST, in order, and say what each tells you. Then
> write ONE more line that pulls out a single column named `score` as a
> vector.*

```{r}
#| eval: false
# your three inspection lines:


# your one access line:

```

> *Your explanation:*

Save as `fnd03_yourlastname.qmd` in your `R/` folder and **Render**.
