---
title: "Foundations 02 — Data Types and Data Structures"
subtitle: "STAT/DATA 1810 · Week 2 Monday · Student version"
author: "Your name here"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT
  CLO 1 (1.5 data types: numeric/character/factor/date · 1.6 containers:
         vector, data frame/tibble, list, matrix · 1.4 assignment reinforced)
  CLO 7 preview (7.6 is.na())
  CLO 9 practiced: you insert your own chunks (taught in Foundations 01)
Time: 80 min. Datasets: tiny Cal Poly examples built in class + iris, airquality.
Rule of the day: a TYPE is what one value is; a STRUCTURE is how values are held.
Packages: ONLY tibble today. We load what the lesson needs, not the whole tidyverse.
LM = learning moment. YT = your turn. Times in headers are the section budget.
-->

```{r}
#| label: setup
library(tibble)   # today we need tibble() and tribble() — that's it
```

> 👀 **Variables check.** Loading `tibble` added *nothing* to the Variables
> pane. Why not? (Think: what does `library()` do versus what `<-` does.)

**Name:** \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_ · **Date:** \_\_\_\_\_\_\_\_\_\_

> Before anything else, type your name and today's date on the line above
> (and put your name in the `author:` line at the very top of this file).

**How to read this document.** Two labels show up over and over:

- **LM = Learning Moment.** I demonstrate; you follow along and run the code.
- **YT = Your Turn.** You do it yourself — fill the blank or the empty chunk.
- **👀 Variables check.** Glance at the **Variables** pane (right side, under
  SESSION) and answer the question. It is your session's memory made visible.

# 0 · The one distinction (4 min)

> **Type** = what a single value *is* (number? text? TRUE/FALSE?).
> **Structure** = the container holding many values (vector? data frame? list?).

Everything today hangs off that sentence.

You already know how to insert a chunk (**Cmd/Ctrl + Shift + I**, from last
week). Whenever you see **"➕ insert a chunk"**, make one — every YT today
depends on it.

{{< vspace 1.5rem >}}

# 1 · Data types (28 min)

## 1.1 · The types, defined first (before any code) (5 min)

Every single value in R has a **type**. Here are the ones you'll live with,
each with a Cal Poly example:

| Type | What it is | Example value |
|---|---|---|
| **numeric** (double) | a number that can have decimals | `4.75` — a latte at Julian's |
| **integer** | a whole number (written with an `L`) | `38L` — people in the Rec Center at 7 am |
| **character** | text, in quotes | `"Scout Coffee"` |
| **logical** | `TRUE` or `FALSE` | `TRUE` — the drink is iced |
| **factor** | text from a *fixed menu* of allowed values | `"morning"` from {morning, afternoon, evening} |
| **Date** / **date-time** | a calendar date; a date *and* clock time | `2026-09-22` / `2026-09-22 07:00:00` |
| **NA** | a missing value — a hole, not a type of its own | the headcount nobody recorded |

## 1.2 · Finding the type with `class()` (7 min)

**LM1 — `class()` tells you the type (4 min).** Predict each `class()` from
the table above, *then* run:

```{r}
# LM1: predict from the table, then run
class(4.75)            # Julian's latte
class(38L)             # people in the Rec Center at 7 am
class("Scout Coffee")  # where the latte came from
class(TRUE)            # iced?
```

> *Q1. Predictions: ____ · ____ · ____ · ____*

**LM2 — doubles vs integers (1 min).** Both are numeric; `typeof()` tells them
apart.

```{r}
# LM2
typeof(4.75)
typeof(38L)
```

**LM3 — dates and date-times (2 min).**

```{r}
# LM3
today <- Sys.Date()
class(today)

right_now <- Sys.time()
class(right_now)
```

## 1.3 · Three families: quantitative, categorical, other (6 min)

R's types sort into three families you'll use to pick plots and summaries all
term:

| Family | Question it answers | R types | Cal Poly examples |
|---|---|---|---|
| **Quantitative** | *how much? how many?* — you can average it | numeric, integer | latte price (`4.75`), Rec Center headcount (`38`), miles from campus (`2.3`) |
| **Categorical** | *which kind? which group?* — you count it | character, factor, logical | coffee shop (`"Julian's"`, `"Starbucks"`, `"Scout"`), time slot (`"morning"`), iced (`TRUE`) |
| **Other** (time/date) | *when?* — ordered, but not a plain number | Date, date-time | the day you bought the latte, the clock time you swiped into the Rec |

![The three families and, inside categorical, the character-vs-factor split. The dashed "Morning"? chip is the trap: a value not on a factor's menu becomes NA with no error.](/images/types_families_map.svg){fig-alt="Three columns — quantitative (average it: numeric, integer), categorical (count it: character, factor, logical), other/time-date (Date, date-time) — each with Cal Poly examples and glimpse tags; below, a panel contrasting character (free text, four spellings of morning make four groups, alphabetical order) with factor (a fixed menu of levels in your order, where a value off the menu becomes NA silently)." width="100%"}

> *Q2. Which family: (a) the number of people in the Rec Center at 5 pm ·
> (b) the shop that made your latte · (c) the date of the game vs. UCSB ·
> (d) whether you bought a latte today (yes/no)?*
>
> *Your answer:* (a) ____ · (b) ____ · (c) ____ · (d) ____

**A small distinction — "string" vs "character".** Other languages call text a
*string*. R calls the *type* **character**, and one piece of text (`"Scout
Coffee"`) is a **string** — a character value of length one. So `"latte"` is a
string, its class is character, and `c("latte", "mocha")` is a character
*vector* of two strings. You'll hear both words; they are not in conflict.

## 1.4 · Factors vs characters — why it matters (6 min)

**LM4 — the same text, two types (4 min).** Rec Center headcounts come with a
time slot. As character, the slots are just words; as a **factor**, they come
with a fixed menu of allowed values (**levels**) *in an order you choose*:

```{r}
# LM4: character vs factor
slot_chr <- c("evening", "morning", "afternoon", "morning", "evening", "evening")
slot_fct <- factor(slot_chr, levels = c("morning", "afternoon", "evening"))

class(slot_chr)
class(slot_fct)
levels(slot_fct)

table(slot_chr)   # alphabetical — afternoon first
table(slot_fct)   # the order YOU chose — morning first
```

> 👀 **Variables check.** `slot_chr` and `slot_fct` are both listed — same six
> values. How does Variables describe each one differently?

> *Q3. How many levels does `slot_fct` have, and what order did we give
> them? Why would that order matter on a bar chart?*
>
> *Your answer:*

**LM5 — NA, the missing value (2 min).** `airquality` (built in) has real
missing ozone readings; `is.na()` finds them and `sum()` counts them:

```{r}
# LM5: count the missing values in Ozone
sum(is.na(airquality$Ozone))
```

> *Q4. How many Ozone values are missing? ____*

**LM6 — the is.\* family (2 min).** Test types with functions that return
TRUE/FALSE:

```{r}
#| eval: false
# LM6: fill each blank so all four return TRUE, then delete the eval line
is.numeric(____)
is.character(____)
is.logical(____)
is.factor(____)
```

> *YT1 (4 min). ➕ Insert a chunk below. In it, find the `class()` of each of
> the following, predicting first: `"TRUE"` (with quotes) · `4.75 > 4.25` ·
> `mtcars$cyl` · `Sys.Date() + 7`. Then in the comment, write which one
> surprised you.*

<!-- ➕ insert a chunk here -->

> *Predictions: ____ · ____ · ____ · ____*

{{< vspace 1.5rem >}}

# 2 · Data structures (24 min)

**LM7 — vector: one type, many values (4 min).** Built with `c()`. Three
places to get a latte near campus:

```{r}
# LM7: a vector
shops <- c("Julian's", "Starbucks", "Scout Coffee")
shops
length(shops)
class(shops)
```

> *YT2 (4 min). ➕ Insert a chunk. Build THREE vectors and print each:*
> *(a) `price` — what a latte costs at each shop above (Julian's on campus,
> the Starbucks in the UU, Scout downtown — use real prices if you know
> them, e.g. 4.75, 5.45, 5.95);*
> *(b) `size` — the size you'd order at each, as text ("small", "medium",
> "large");*
> *(c) `iced` — three TRUE/FALSE values, whether you'd get it iced.*
> *Then run `class()` on each. What happens if you run `c(4.75, "five", 5.95)`?
> Try it and look at the class.*

<!-- ➕ insert a chunk here -->

> *Q5. `c(4.75, "five", 5.95)` has class ____. In one phrase, why?*
>
> *Your answer:*

**LM8 — data frame & tibble: vectors standing side by side (6 min).** Each
column is a vector; every column has the same length.

```{r}
# LM8: same data, two containers
latte_df  <- data.frame(shop = shops, price = c(4.75, 5.45, 5.95))
latte_tbl <- tibble(shop = shops, price = c(4.75, 5.45, 5.95))

latte_df
latte_tbl
```

![Same three-latte table printed by `data.frame()` and by `tibble()`, and the class() proof that a tibble is a data frame with extra labels.](/images/tibble_vs_dataframe.svg){fig-alt="Left panel: a data.frame printout with no dimensions and no column types. Right panel: the same table as a tibble announcing '# A tibble: 3 × 2' and showing <chr> and <dbl> type tags. Bottom panel: a nesting diagram of tibbles inside data frames, with class(menu_tbl) returning tbl_df, tbl, data.frame as proof." width="100%"}

> *Q6. Name one visible difference between how the data frame and the tibble
> printed:*
>
> *Your answer:*

> *YT3 (5 min). ➕ Insert a chunk. Build a tibble called `lattes` from YOUR
> vectors from YT2 — columns `shop`, `price`, `size`, `iced`. Print it.
> Then run `nrow(lattes)`, `ncol(lattes)`, and `class(lattes)`. What happens if
> one vector has only two values? Try it (make a broken one on purpose), read
> the error, then fix it.*

<!-- ➕ insert a chunk here -->

> 👀 **Variables check.** `lattes` gets an expand arrow that `price` did not.
> Click it. What is inside, and why does a tibble get one when a vector doesn't?

**LM9 — tribble: a tibble typed row by row, with a factor inside (3 min).**
Rec Center headcounts, three slots. Note the `slot` column is made a factor
*with the order we want*:

```{r}
# LM9: Rec Center headcounts — a tribble, then a factor column
rec <- tribble(
  ~slot,       ~hour, ~people,
  "morning",     7,     38,
  "afternoon",  13,    112,
  "evening",    18,    164
)
rec$slot <- factor(rec$slot, levels = c("morning", "afternoon", "evening"))
rec
```

**LM10 — list: the junk drawer (4 min).** A list holds *anything*, even other
structures:

```{r}
# LM10
my_list <- list(name = "week 2", nums = c(1, 2, 3), data = latte_tbl)
my_list
length(my_list)
```

> *YT4 (3 min). ➕ Insert a chunk. Build a list called `me` with three named
> parts: your first name (text), the number of classes you're taking this
> quarter, and your `lattes` tibble. Print it. Notice how each part keeps its
> own type.*

<!-- ➕ insert a chunk here -->

**LM11 — matrix: all one type, rows × columns (2 min)** (brief):

```{r}
# LM11
matrix(1:6, nrow = 2)
```

> *Q7. Which structure could hold a column of shop names AND a column of
> prices: a matrix or a data frame? Why?*
>
> *Your answer:*

{{< vspace 1.5rem >}}

# 2b · When things break (6 min)

Today's errors are *type* and *structure* errors. Same rule as last week:
read the message slowly, name the family, then fix it.

> *YT5 (6 min). Each chunk below has ONE bug. Read the error (or the silent
> wrong answer!), fix it in place, and write the error family in the comment.*

```{r}
#| eval: false
# YT5a — error family: ____
price <- c(4.75, "5.45", 5.95)
mean(price)
```

```{r}
#| eval: false
# YT5b — error family: ____
tibble(shop = c("Julian's", "Starbucks", "Scout Coffee"), price = c(4.75, 5.45))
```

```{r}
#| eval: false
# YT5c — error family: ____
factor(c("morning", "Morning", "evening"), levels = c("morning", "afternoon", "evening"))
```

> *Which of the three gave NO red error message? ____ What did it give instead?*

{{< vspace 1.5rem >}}

# 3 · In-class activity: type it or contain it (12 min)

For each item, write **T** if the question is about a type and **S** if it's
about a structure, then answer it with code. ➕ Insert **one chunk per item**
(A through F) so each answer is its own runnable piece.

- **A.** What is `rec` (from LM9) — its structure?
- **B.** What is `rec$people` — its type? Which family?
- **C.** Is there any NA anywhere in `airquality$Solar.R`? How many?
- **D.** Build a tibble with 2 columns: three places you eat near campus and
  a rating (1–10) for each.
- **E.** What is `airquality$Month` — its type? Should it be? (Months are
  stored as numbers here — would a factor make more sense? One sentence.)
- **F.** Build a list holding your food tibble from D *and* a character
  vector of what you order at each place.

<!-- ➕ insert chunks A–F here -->

> *A: ____ · B: ____ · C: ____ · D: ____ · E: ____ · F: ____*

{{< vspace 1.5rem >}}

# 4 · Exit question (5 min)

> *In your own words, one sentence each: what is a data type, and what is a
> data structure? Use one example of each from today — and at least one must
> be something YOU built.*
>
> *Your answer:*

# 5 · Save & render (1 min)

Save this file into your `R/Notes/` folder as `fnd02_yourlastname.qmd`. Then
**Restart R** (top-right session menu → Restart) — watch Variables empty out —
and press **Render** (Cmd/Ctrl + Shift + K). If a chunk you inserted breaks
the render, read the error: it names the chunk. Restarting first is how you
prove that everything the document needs is *in* the document.
