---
title: "Foundations 01 — Positron and the R Environment"
subtitle: "STAT/DATA 1810 · Week 1 Wednesday · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right-body
    toc-title: "Contents"
execute:
  warning: false
  message: false
---

```{=html}
<!--
CLO: 1.1, 1.2, 1.3, 1.4, 1.7 · 11.1 · (9 practiced)
Timing: panes+session 12' · running code/chunks 12' · objects 16'
        · functions/pkgs/help 20' · errors 12' · save & render 8' = 80'
Watch the clock at the end of §2 (40' mark) — if behind, cut YT6 and YT7c.
-->
```

**Name:** Instructor key · **Date:** \_\_\_\_\_\_\_\_\_\_

::: {.callout-warning title="Timing & labels"}
- **Budget:** panes + session 12′ · running code 12′ · objects 16′ · functions/packages/help 20′ · errors 12′ · save & render 8′ = **80′**
- **Checkpoint:** end of §2 at the **40′ mark** — if behind, cut YT6 and YT7c
- [LM]{.lm} **Learning Moment** — instructor demonstrates, students follow
- [YT]{.yt} **Your Turn** — students do it themselves
- Say both names out loud the first time each appears — students otherwise guess
:::

# 0 · Getting set up and what you are looking at [12 min]{.timing}

## Setup checklist

::: {.callout-important title="Teaching note — setup"}
- The eight setup steps should be done **before** class: install R → install Positron → download the `.qmd` → make `Desktop/R/{Notes,Labs}` and `Desktop/Data` → move the file into `R/Notes` → open Positron → **File → Open Folder…** on `Desktop/R` → open the file from the sidebar → fill in name and date
- Budget **3 minutes** at the start to rescue whoever isn't there yet
- Most common failures:
  - the file is still in `Downloads`
  - they opened the *file* rather than the *folder*, so the Explorer sidebar is empty
- Fix for both: **File → Open Folder… → `Desktop/R`**
- Make them fill in name and date — first thing they type into a `.qmd`, and how you tell whose render is whose
:::

## The panes [6 min]{.timing}

::: {.callout-important title="Teaching note — panes"}
- Put the annotated screenshot (`images/positron_panes.png`) on the projector
- Have students physically point at each numbered pane — the checklist **is** the activity
- Use Positron's own names consistently: **Editor**, **Console**, **Variables**, **Plots**, **Explorer**
- Two traps:
  - **CONSOLE** and **TERMINAL** tabs sit side by side — the Terminal shows a `%` prompt, not `>`
  - Variables/Plots live inside the **SESSION** panel on the right — if closed, the right side looks blank
- Terminal / Viewer / Connections get one sentence each — resist the tour becoming the lesson
:::

::: {.callout-tip title="Answers"}
- [YT1]{.yt} Variables
- [YT2]{.yt} the Editor (this `.qmd`)
:::

## The Session [6 min]{.timing}

::: {.callout-important title="Teaching note — getting to the Session tab"}
- Point at the top-right interpreter label (`R 4.x.x`, arrow 6 on the screenshot)
- Click it once so they see the menu: running session, Restart, Shutdown, other interpreters — then press Escape
- Point at the **SESSION** panel beneath it: *"that label is the session; this panel is what the session remembers"*
- Say the three facts — **memory / restartable / replaceable** — and move on
- Payoff comes in §4 LM11 when they restart for real
- Show that Variables is empty *now* — the "before" picture makes the restart in §4 land
:::

::: {.callout-tip title="Q1 model answer"}
- [Q1]{.q} The Console runs code once and forgets it
- The Editor holds code you keep, so you can re-run everything tomorrow
:::

# 1 · Running code and code chunks [12 min]{.timing}

::: {.callout-important title="Teaching notes — LM1–LM3"}
- [LM1]{.lm} The up-arrow trick saves students dozens of retypes this term — worth the 20 seconds
- [LM2]{.lm} **Anatomy of a chunk:**
  - put the cursor on the ```` ```{r} ```` line and name its parts out loud: fence, `{r}` = "this is R", fence
  - click the play button so they see output land **both** under the chunk and in the Console
  - the sentence to leave them with: *"a chunk is a saved bundle of Console lines"*
  - students create their own chunks starting Monday (Foundations 02) — today they only run and edit ours
- [LM3]{.lm} One sentence on `#|` options is enough — they'll write `#| label:` and `#| eval: false` themselves in ETV cycles
:::

```{r}
# LM2
2 + 2
3 * 7
10 / 4
```

```{r}
# YT3
5 - 4
100 / 8
2^10
```

::: {.callout-note title="Expected — YT3"}
- [YT3]{.yt} `1` · `12.5` · `1024`
- [Q2]{.q} **both** (output appears under the chunk *and* in the Console)
:::

# 2 · Objects and assignment [16 min]{.timing}

```{r}
# LM4
cups <- 3
cups
```

```{r}
# LM5 — prediction: 13.5
price <- 4.50
total <- cups * price
total
```

```{r}
# LM6
cups <- 10
total
```

::: {.callout-tip title="Q4 model answer — the most important idea of the day"}
- [Q4]{.q} `total` was computed *at the moment* the `<-` line ran, using `cups = 3`
- Changing `cups` later doesn't re-run old lines — **R objects store values, not formulas**
- Spend the full minutes here
- Spreadsheet contrast lands well: Excel cells recompute; R objects don't
:::

```{r}
# YT4 — any sleep value; e.g.
hours   <- 7
minutes <- hours * 60
minutes
hours   <- 8
minutes            # still 420 — same lesson as Q4, now self-discovered
```

::: {.callout-tip title="YT4 (d) answer"}
- [YT4]{.yt} `minutes` does **not** change — Q4 again, but the student built it, so it sticks
- Have two students say their prediction and result aloud
- Then point at Variables: `cups`, `price`, `total`, `hours`, `minutes` — five objects, all session memory
:::

# 3 · Functions, packages, help [20 min]{.timing}

```{r}
# LM7
mean(mtcars$mpg)
nrow(mtcars)
```

::: {.callout-note title="Expected — LM7"}
- [LM7]{.lm} `20.09062` · `32`
:::

```{r}
# YT5
max(mtcars$hp)
min(mtcars$mpg)
ncol(mtcars)
round(mean(mtcars$mpg), 1)
```

::: {.callout-note title="Expected — YT5"}
- [YT5]{.yt} `335` (the Maserati Bora) · `10.4` · `11` · `20.1`
- Part (d) is the first *nested* call — say "inside out" once; don't dwell
:::

::: {.callout-tip title="Q5 · YT6 · Q6 answers"}
- [Q5]{.q} `na.rm = TRUE` tells `mean()` to remove `NA` values before computing; the default `FALSE` returns `NA` if any value is missing
  - plant the flag: this argument returns in Foundations 03 and every ETV cycle
- [YT6]{.yt} `digits`, default `0` — the purpose is reading a Usage line, not the fact itself
- [Q6]{.q} install runs **once per computer**; library runs **once every session**
  - analogy that sticks: install = buying the drill; library = taking it out of the garage
:::

```{r}
library(tidyverse)
```

# 4 · When things break [12 min]{.timing}

```{r}
#| eval: false
meen(mtcars$mpg)
#> Error in meen(mtcars$mpg) : could not find function "meen"
```

```{r}
# LM10 fixed
mean(mtcars$mpg)
```

::: {.callout-tip title="Q7 model answer"}
- [Q7]{.q} "could not find function" means the *name* doesn't exist — a typo, or a package not loaded
- Name the two most common error families now:
  - `could not find function` → typo / `library()` missing
  - `object not found` → name typo / never created
- Students who can *classify* an error stop fearing errors
:::

::: {.callout-caution title="YT7 answers — three error families"}
- [YT7a]{.yt} `mtcars$mgp` → returns `NULL`; `mean` warns "argument is not numeric or logical: returning NA"
  - family: **column-name typo** — `$` on a missing column gives `NULL`, not an error (the sneakiest of the three)
  - fix: `mpg`
- [YT7b]{.yt} `Mtcars` → `Error: object 'Mtcars' not found`
  - family: **object not found** — R is case-sensitive
  - fix: `mtcars`
- [YT7c]{.yt} missing `)` → Console shows a `+` prompt waiting for more; in a chunk, "unexpected end of input"
  - family: **unbalanced parentheses**
  - fix: add `)` — and show them **Escape** in the Console to get out of a `+` prompt
:::

```{r}
mean(mtcars$mpg)
mean(mtcars$mpg)
mean(mtcars$mpg)
```

::: {.callout-important title="LM11 — restart for real"}
- [LM11]{.lm} Everyone restarts → Variables empties
- [Q8]{.q} `cups` gone · code still here · re-running §2 takes ~10 seconds
- That 10-second recovery *is* the argument for putting code in the file
:::

# 5 · Save & render [8 min]{.timing}

::: {.callout-tip title="Exit model answer"}
- `cups` lives in Variables → gone after restart
- this `.qmd` lives in the Editor → survives
- the code that *recreates* `cups` survives — the entire argument for reproducible documents, made on day one without ever saying "CLO 9"
:::
