---
title: "Foundations 01 — Positron and the R Environment"
subtitle: "STAT/DATA 1810 · Week 1 Wednesday · SOLUTION"
author: "Instructor key"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

```{=html}
<!--
CLO: 1.1, 1.2, 1.3, 1.4, 1.7 · 11.1 · (9 practiced)
Timing: panes+session 12' · running code/chunks 12' · objects 16'
        · functions/pkgs/help 20' · errors 12' · save & render 8' = 80'
Watch the clock at the end of §2 (40' mark) — if behind, cut YT6 and YT7c.
-->
```

**Name:** Instructor key · **Date:** \_\_\_\_\_\_\_\_\_\_

**Labels:** **LM = Learning Moment** (instructor demonstrates, students follow). **YT = Your Turn** (students do it themselves). Say both names out loud the first time each appears — students otherwise guess.

# 0 · Getting set up and what you are looking at (12 min)

## Setup checklist

**Teaching note:** the eight setup steps (install R → install Positron → download the `.qmd` → make `Desktop/R/{Notes,Labs}` and `Desktop/Data` → move the file into `R/Notes` → open Positron → **File → Open Folder…** on `Desktop/R` → open the file from the sidebar → fill in name and date) should be done *before* class. Budget 3 minutes at the start to rescue whoever isn't there yet; the most common failures are (a) the file is still in `Downloads`, and (b) they opened the *file* rather than the *folder*, so the Explorer sidebar is empty. The fix for both is the same: File → Open Folder… → `Desktop/R`. Make them fill in name and date — it is the first thing they type into a `.qmd`, and it is how you tell whose render is whose.

## The panes (6 min)

**Teaching note:** put the annotated screenshot (`images/positron_panes.png`) on the projector and have students physically point at each numbered pane; the checklist is the activity. Use Positron's own names consistently — **Editor**, **Console**, **Variables**, **Plots**, **Explorer** — because that is what students see on screen. Two traps: the **CONSOLE** tab and the **TERMINAL** tab sit side by side at the bottom and students click the wrong one (the Terminal shows a `%` prompt, not `>`); and Variables/Plots live inside the **SESSION** panel on the right, so if a student has closed it, the right side looks blank. Terminal/Viewer/Connections get one sentence each — resist the tour becoming the lesson.

**YT1:** Variables. **YT2:** the Editor (this `.qmd`).

## The Session (6 min)

**Teaching note — getting to the Session tab.** Point at the top-right interpreter label (`R 4.x.x`, arrow 6 on the screenshot). Click it once so they see the menu: the running session, Restart, Shutdown, and any other interpreters. Press Escape. Then point at the **SESSION** panel beneath it and say: "that label is the session; this panel is what the session remembers." Say the three facts (memory / restartable / replaceable) and move on. The payoff comes in §4 LM11 when they restart for real. Show that Variables is empty *now* — that "before" picture makes the restart in §4 land.

**Q1 model answer:** the Console runs code once and forgets it; the Editor holds code you keep, so you can re-run everything tomorrow.

# 1 · Running code and code chunks (12 min)

**LM1:** the up-arrow trick saves students dozens of retypes this term — worth the 20 seconds.

**LM2 teaching note — anatomy of a chunk.** Put the cursor on the ```` ```{r} ```` line and name its parts out loud: fence, `{r}` = "this is R", fence. Then click the play button so they see output land **both** under the chunk and in the Console. The sentence to leave them with: *"a chunk is a saved bundle of Console lines."* Students create their own chunks starting Monday (Foundations 02) — today they only run and edit ours.

```{r}
# LM2
2 + 2
3 * 7
10 / 4
```

**LM3:** one sentence on `#|` options is enough. They'll write `#| label:` and `#| eval: false` themselves in ETV cycles.

```{r}
# YT3
5 - 4
100 / 8
2^10
```

**Expected:** 1 · 12.5 · 1024. **Q2:** both.

# 2 · Objects and assignment (16 min)

```{r}
# LM4
cups <- 3
cups
```

```{r}
# LM5 — prediction: 13.5
price <- 4.50
total <- cups * price
total
```

```{r}
# LM6
cups <- 10
total
```

**Q4 model answer:** `total` was computed *at the moment* the `<-` line ran, using `cups = 3`. Changing `cups` later doesn't re-run old lines — R objects store values, not formulas. **This is the single most important idea of the day**; spend the full minutes here. (Spreadsheet contrast lands well: Excel cells recompute; R objects don't.)

```{r}
# YT4 — any sleep value; e.g.
hours   <- 7
minutes <- hours * 60
minutes
hours   <- 8
minutes            # still 420 — same lesson as Q4, now self-discovered
```

**YT4 (d):** `minutes` does not change. This is Q4 again, but the student built it, so it sticks. Have two students say their prediction and result aloud. Then point at Variables: `cups`, `price`, `total`, `hours`, `minutes` — five objects, all session memory.

# 3 · Functions, packages, help (20 min)

```{r}
# LM7
mean(mtcars$mpg)
nrow(mtcars)
```

**Expected:** 20.09062 · 32.

```{r}
# YT5
max(mtcars$hp)
min(mtcars$mpg)
ncol(mtcars)
round(mean(mtcars$mpg), 1)
```

**Expected:** 335 (the Maserati Bora) · 10.4 · 11 · 20.1. Part (d) is the first *nested* call — say "inside out" once; don't dwell.

**Q5 model answer:** `na.rm = TRUE` tells `mean()` to remove `NA` values before computing; the default `FALSE` returns `NA` if any value is missing. (Plant the flag: this argument returns in Foundations 03 and every ETV cycle.)

**YT6:** `digits`, default `0`. The purpose is reading a Usage line, not the fact itself.

```{r}
library(tidyverse)
```

**Q6:** install runs **once** per computer; library runs **once every session**. Analogy that sticks: install = buying the drill; library = taking it out of the garage.

# 4 · When things break (12 min)

```{r}
#| eval: false
meen(mtcars$mpg)
#> Error in meen(mtcars$mpg) : could not find function "meen"
```

```{r}
# LM10 fixed
mean(mtcars$mpg)
```

**Q7 model answer:** "could not find function" means the *name* doesn't exist — a typo, or a package not loaded. **Teaching note:** name the two most common error families now: `could not find function` (typo/library) and `object not found` (name typo / never created). Students who can *classify* an error stop fearing errors.

**YT7 answers:**

-   **7a** `mtcars$mgp` → returns `NULL`, and `mean` warns "argument is not numeric or logical: returning NA". Family: **column-name typo** (`$` on a missing column gives `NULL`, not an error — worth saying, since it's the sneakiest of the three). Fix: `mpg`.
-   **7b** `Mtcars` → `Error: object 'Mtcars' not found`. Family: **object not found** (R is case-sensitive). Fix: `mtcars`.
-   **7c** missing `)` → the Console shows a `+` prompt waiting for more; in a chunk, "unexpected end of input". Family: **unbalanced parentheses**. Fix: add `)`. Show them **Escape** in the Console to get out of a `+` prompt.

```{r}
mean(mtcars$mpg)
mean(mtcars$mpg)
mean(mtcars$mpg)
```

**LM11 — restart for real.** Everyone restarts. Variables empties. Then Q8: `cups` gone · code still here · re-running §2 takes \~10 seconds. That 10-second recovery *is* the argument for putting code in the file.

# 5 · Save & render (8 min)

**Exit model answer:** `cups` lives in Variables (gone after restart); this `.qmd` lives in the Editor (survives). The code that *recreates* `cups` survives — which is the entire argument for reproducible documents, made on day one without ever saying "CLO 9."