---
title: "Foundations 01 — Positron and the R Environment"
subtitle: "STAT/DATA 1810 · Week 1 Wednesday · Student version"
author: "Your name here"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

<!--
CLO ALIGNMENT
  CLO 1 (1.1 what R is · 1.2 Positron · 1.3 projects/directories ·
         1.4 assignment · 1.7 packages vs libraries)
  CLO 11 (11.1 ?function help)
  CLO 9 practiced, not graded (this document IS a Quarto doc)
Time: 80 min. Dataset: mtcars (built in — no import needed yet).
LM = learning moment. YT = your turn (practice). Fill every ____ and every
empty chunk. Times in headers are the budget for that section.
-->

# 0 · What you are looking at (12 min)

**R** is a programming language for working with data. **Positron** is the
program we use to *talk to* R — an integrated development environment (IDE).
R does the work; Positron is the workshop it happens in.

## The panes (6 min)

Positron is divided into **panes**. Find each one on your screen and check it
off — physically point at it:

- [ ] **Editor** (Source) — the big middle area where this `.qmd` file is open.
  Code you *keep* lives here.
- [ ] **Console** — bottom. R answers here, one command at a time. Code you
  type here is *not saved*.
- [ ] **Variables** — right side (RStudio calls it *Environment*). Every object
  you create this session shows up here.
- [ ] **Plots** — right side. Pictures appear here.
- [ ] **Help** — right side. Documentation appears here when you ask for it.
- [ ] **Explorer** (Files) — left sidebar. The files in your project folder.
- [ ] Terminal · Viewer · Connections — know they exist; we'll meet them when
  we need them.

> *YT1 (1 min). A neighbor asks "where do I look to see what objects I've
> made?" Which pane do you point at? ____*
>
> *YT2 (1 min). You type code and want it to be there tomorrow. Which pane
> does it go in? ____*

## The Session (6 min)

R runs inside a **session** — a running copy of R that Positron is talking to.
Look at the **top-right corner** of Positron: it shows which R is running
(something like `R 4.x.x`). That is your session.

Three things to know about the session, today:

1. **It has memory.** Every object you create lives in the session, and shows
   up in Variables.
2. **It can be restarted.** In the Console toolbar (or the top-right session
   menu) there is a **Restart** button. Restarting wipes the session's memory
   clean — Variables empties out. Your *files* are untouched.
3. **It can be replaced.** The same menu lets you pick a different R (or
   Python) if more than one is installed. Don't do that today; just know the
   button exists.

Do this now: create nothing yet, look at Variables (empty), and remember what
empty looks like — we'll come back after we've made some objects.

> *Q1. In one sentence: what is the difference between the Console and the
> Editor?*
>
> *Your answer:*

# 1 · Running code and code chunks (12 min)

**LM1 — the Console (2 min).** Click into the Console, type `2 + 2`, press
Enter. R answers. Now press the up-arrow — the Console remembers your last
line. Edit it to `2 + 3` and press Enter.

**LM2 — a code chunk (5 min).** This document is a **Quarto** file: text and
code mixed together. The gray boxes are **code chunks**. A chunk is:

- opened by a line that reads ```` ```{r} ```` (three backticks, then `{r}`)
- closed by a line of three backticks ```` ``` ````
- everything between the two is R code

Look at the chunk below in the Editor. Notice the small **play button** at
its top-right corner. Click it (or put your cursor on a line and press
**Cmd/Ctrl + Enter** to run just that line):

```{r}
# LM2: this line is a COMMENT — R ignores anything after #
2 + 2
3 * 7
10 / 4
```

Two things happened: the results appeared **under the chunk** *and* the same
code ran in the **Console**. A chunk is just a saved bundle of Console lines.

**LM3 — chunk options (2 min).** Lines that start with `#|` at the top of a
chunk are **options** — instructions to Quarto, not R. You will see
`#| eval: false` in this document; it means "show this code, but don't run
it when rendering." Useful for code that is broken on purpose.

```{r}
#| eval: false
# LM3: this chunk is displayed but never runs
this_would_be_an_error
```

> *YT3 (3 min). Edit the chunk below so it computes THREE things: 5 minus
> something to give 1; 100 divided by 8; and 2 to the 10th power (hint: `^`).*

```{r}
# YT3: edit and run — start with:  5 - ____

```

> *Q2. Where did the output show up — under the chunk, in the Console, or
> both? ____*

# 2 · Objects and assignment (16 min)

**LM4 — the arrow (3 min).** `<-` stores a value under a name. Without it,
R computes and forgets. (Type it fast: `Alt + -` inserts `<-` in Positron.)

```{r}
# LM4: create an object, then call its name
cups <- 3
cups
```

Look at the **Variables pane** — `cups` is there now, with its value.

**LM5 — objects combine (4 min).** Predict before you run: what will `total`
be?

```{r}
# LM5: predict, then run
price <- 4.50
total <- cups * price
total
```

> *Q3. Prediction: ____  · What R said: ____*

**LM6 — overwriting (4 min).** Run this, then check the Variables pane again.

```{r}
# LM6: same name, new value
cups <- 10
total
```

> *Q4. `total` did NOT change to 45. Why not? (Hint: WHEN was `total`
> computed?)*
>
> *Your answer:*

> *YT4 (5 min). In the chunk below:*
> *(a) create `hours` holding how many hours you slept last night;*
> *(b) create `minutes` as `hours` times 60;*
> *(c) print `minutes`;*
> *(d) predict — if you now change `hours` to 8, does `minutes` change? Test it.*

```{r}
# YT4:

```

> *Your prediction for (d): ____ · What happened: ____*

Now look at Variables: how many objects are there? Names are on the left,
values on the right — this pane is your session's memory, made visible.

# 3 · Functions, packages, help (20 min)

**LM7 — a function (4 min).** A function is a named action: `name(arguments)`.
`mtcars` is a dataset built into R — 32 cars from 1974. No import needed.
The `$` pulls one column out of it (much more on `$` next week).

```{r}
# LM7: call functions on built-in data
mean(mtcars$mpg)
nrow(mtcars)
```

> *YT5 (4 min). One line each:*
> *(a) `max()` of horsepower, `mtcars$hp`;*
> *(b) `min()` of miles-per-gallon;*
> *(c) `ncol()` of `mtcars` — how many columns?*
> *(d) `round()` the mean mpg to 1 decimal place. (Hint: `round(x, 1)`.)*

```{r}
# YT5:

```

**LM8 — help (5 min).** Every function has documentation. Run this, and the
Help pane opens:

```{r}
#| eval: false
?mean
```

Read the **Usage** and **Arguments** sections. That's where the answers to
"what does this function want from me?" always live.

> *Q5. From the help page: what does the `na.rm` argument do?*
>
> *Your answer:*

> *YT6 (2 min). Look up `?round`. What is the name of its second argument,
> and what is its default value? ____ · ____*

**LM9 — packages (5 min).** A **package** is a collection of functions
someone wrote. You **install** once (download to your computer), you **load**
every session (open the toolbox):

```{r}
#| eval: false
install.packages("tidyverse")   # ONCE per computer — done in setup
library(tidyverse)              # EVERY session you use it
```

```{r}
# LM9: load it now (already installed)
library(tidyverse)
```

> *Q6. Fill the blanks: install.packages() runs ____ per computer;
> library() runs ____ per session.*

# 4 · When things break (12 min)

**LM10 — read the error (4 min).** Run this broken line. Read the message
*slowly* before fixing it.

```{r}
#| eval: false
# LM10: broken on purpose — fix it in the chunk below
meen(mtcars$mpg)
```

```{r}
# LM10 fixed:

```

> *Q7. What did the error message say, and what did it mean?*
>
> *Your answer:*

> *YT7 (4 min). Each chunk below has ONE bug. Read the error, then fix it
> in place. Write the error family in the comment (typo in a function name?
> object that doesn't exist? something else?).*

```{r}
#| eval: false
# YT7a — error family: ____
mean(mtcars$mgp)
```

```{r}
#| eval: false
# YT7b — error family: ____
mean(Mtcars$mpg)
```

```{r}
#| eval: false
# YT7c — error family: ____
mean(mtcars$mpg
```

**LM11 — the reset button (4 min).** If R gets confused, restart the session:
Console toolbar → **Restart R** (or the top-right session menu). Do it now,
for real. Then look at Variables.

> *Q8. After the restart: is `cups` still in Variables? ____ · Is the code
> that CREATED `cups` still in this file? ____ · Re-run section 2's chunks —
> how long did it take to get `cups` back? ____*

# 5 · Save & render (8 min)

Save this file into your project's `R/` folder as
`fnd01_yourlastname.qmd`, then press **Render** (the button above the
Editor, or Cmd/Ctrl + Shift + K). A webpage appears — your first
reproducible document. Every chunk in it re-ran from a *fresh* session to
build that page.

> *Exit question: name one thing that lives in the Variables pane and one
> thing that lives in the Editor, and say which one survives a restart.*
>
> *Your answer:*
